---
title: <iconify-icon icon="fa6-solid:disease"></iconify-icon> Legionnaire's Disease in the USA
order: 350
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(patchwork)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
    
    theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```


## Introduction

Legionnaires’ disease (LD) is a severe form of pneumonia (∼10–25% fatality rate) caused by inhalation of aerosols containing `Legionella`, a pathogenic gram-negative bacteria. These bacteria can grow, spread, and aerosolize through building water systems. A recent dramatic increase in LD incidence has been observed globally, with a 9-fold increase in the United States from 2000 to 2018, 

Records were also maintained of atmospheric Sulphur Dioxide (SO2) and the acidity i.e. pH of the atmosphere around building water systems such as Cooling Towers (CT) and in Rainwater. 

This data is from this paper: Yu F, Nair AA, Lauper U (2024), <https://doi.org/10.6084/m9.figshare.25157852.v2>

## Read the Modified Data

```{r}
#| echo: false
leg_so2_orig <- read_delim("../../../../../materials/Data/LD/fig3bcopy.csv")
##
ld <- leg_so2_orig %>% slice(1:28) %>% 
  dplyr::select("year" = `@`, "ld" = ...2) %>% 
  dplyr::mutate(year = as.integer(year))
##
so2_Nassau_Erie <- leg_so2_orig %>% slice(30:57) %>% 
  dplyr::select("year" = `@`, "so2_Nassau_Erie" = ...2) %>% 
  dplyr::mutate(year = as.integer(year))
##
so2_US <- leg_so2_orig %>% slice(59:86) %>% 
  dplyr::select("year" = `@`, "so2_US" = ...2) %>% 
  dplyr::mutate(year = as.integer(year))
##
ph_rainwater <- leg_so2_orig %>% slice(88:115) %>% 
  dplyr::select("year" = `@`, "pH_rainwater" = ...2, "upper" = ...3, lower = ...4) %>% 
  dplyr::mutate(year = as.integer(year))
##
ph_ct <- leg_so2_orig %>% slice(117:145) %>% 
  dplyr::select("year" = `@`, "pH_CT" = ...2) %>% 
  dplyr::mutate(year = as.integer(year))

```
<br>
```{r}
#| echo: false
ld %>%
  downloadthis::download_this(
    output_name = "Disease",
    output_extension = ".csv",
    button_label = "Download LD prevalence data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```
<br>
```{r}
#| echo: false
so2_Nassau_Erie %>%
  downloadthis::download_this(
    output_name = "SO2_Nassau",
    output_extension = ".csv",
    button_label = "Download SO2 in Nassau data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```
<br>
```{r}
#| echo: false
so2_US %>%
  downloadthis::download_this(
    output_name = "SO2_US",
    output_extension = ".csv",
    button_label = "Download SO2 in mainland US data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```
<br>
```{r}
#| echo: false
ph_ct %>%
  downloadthis::download_this(
    output_name = "pH_ct",
    output_extension = ".csv",
    button_label = "Download pH Cooling Tower",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```
<br>
```{r}
#| echo: false
ph_rainwater %>%
  downloadthis::download_this(
    output_name = "pH_rainwater",
    output_extension = ".csv",
    button_label = "Download pH Rainwater",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )
```


## Inspect the Data

```{r}
#| label: inspect-skim-glimpse
#| echo: fenced

# Write in

```


## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

Describe how you may plan to transform the data.


## Research Question
::: callout-note
Write in! Look first at the Charts below!
:::

## Join the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

Here is the plot-ready data:

```{r}
#| echo: false
##
ld_mod <- ld %>% 
  dplyr::left_join(so2_Nassau_Erie) %>% 
  left_join(so2_US) %>% 
  left_join(ph_rainwater) %>% 
  left_join(ph_ct)
ld_mod

```





## Plot the Data

Two plots were generated by the researchers with this data. Can you reproduce these? Do these graphs prove/disprove any of your hypotheses? What might have been the Hypotheses that led the creating of these graphs?

```{r}
#| echo: false
#| label: ld-so2
# Set graph theme
theme_set(new = theme_custom())
#
ld_mod %>%
  
  gf_line(ld ~ year, linewidth = 2, 
          xlab = "Year", 
          title = "Legionnaires' Disease (LD) in the US", 
          colour =  ~ "LD",
          subtitle = "Influenced by SO2 Pollution?",
          caption = "Data from: Yu F, Nair AA, Lauper U (2024)\n https://doi.org/10.6084/m9.figshare.25157852.v2") %>%
  gf_point(ylab = "LD Prevalence (relative to 1992)",
           colour = ~ "LD", size = 4) %>%
  
  gf_line(so2_Nassau_Erie ~ year, 
          colour = ~ "SO2 in Nassau") %>% 
  gf_point(so2_Nassau_Erie ~ year, 
           colour = ~ "SO2 in Nassau", 
           shape = 15,size = 3) %>%
  
  gf_line(so2_US ~ year, colour = ~ "SO2 in the US") %>% 
  gf_point(so2_US ~ year, colour = ~ "SO2 in the US",
           shape = 12, size = 3) %>% 

  gf_refine(
    scale_colour_grey("LD"),
    scale_x_continuous(breaks = seq(1992, 2019, 2)),
    scale_y_continuous(
      breaks = seq(1:15),
      labels = scales::label_number(scale = 1, 
                                    suffix = "X"),
      sec.axis = sec_axis(trans = ~./1, 
                          name = "SO2 Concentration")
  )) %>% 
  gf_theme(theme(legend.title = element_blank(),
                 legend.position = "top"))
```

```{r}
#| label: ld-ph
#| echo: false
# Set graph theme
theme_set(new = theme_custom())
#

ld_mod %>%
  gf_line(ld ~ year, linewidth = 2, 
          xlab = "Year", 
          colour =  ~ "LD") %>%
  gf_point(ylab = "LD Prevalence (relative to 1992)",
           colour = ~ "LD", size = 3) %>% 
  gf_line(5*(pH_rainwater-3) ~ year, 
          colour = ~ "pH Rainwater") %>%
  gf_point(5*(pH_rainwater-3) ~ year, 
           colour = ~ "pH Rainwater", size = 3) %>% 
  
  gf_line(5 * (pH_CT-3) ~ year, colour = ~ "pH Cooling Towers", 
          ylab = "pH") %>%
  gf_point(
    5 * (pH_CT - 3) ~ year,
    colour = ~ "pH Cooling Towers",
    shape = 15,
    size = 3
  ) %>%
  
  gf_refine(
    scale_colour_grey("pH"),
    scale_x_continuous(breaks = seq(1992, 2019, 2)),
    scale_y_continuous(
      breaks = seq(1:15),
      labels = scales::label_number(scale = 1, 
                                    suffix = "X"),
      sec.axis = sec_axis(trans = ~(.x/5 +3), 
                          name = "pH"))) %>% 
  
  gf_labs(title = "Legionnaires' Disease (LD) in the US", 
                    subtitle = "Influenced by Atmospheric pH?",
          caption = "Data from: Yu F, Nair AA, Lauper U (2024)\n https://doi.org/10.6084/m9.figshare.25157852.v2") %>% gf_theme(theme_custom()) %>% 
  gf_theme(theme(legend.title = element_blank(),
                 legend.position = "top"))  

```




## Tasks and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown. Combine the multiple datasets into one if needed!
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  - What is a peculiar feature of these graphs?
- What might have been the Hypothesis/Research Question to which the response was Chart?
- What data gathering / research activity might have been carried out to obtain the data graphed here? Provide some details.
- Write a short story based on the chart, describing your inference/surprise. 
- Is there a paradox in this case study? Hint: SO2 is caused by cars/busses running on fossil fuels.
- What Statistical Tests might you run to confirm what the charts are saying?





    

