---
title: "Font and Theme Demo"
subtitle: "Testing the Fixed Theming System"
format: html
---

## Overview

This document demonstrates that the font and theme issues have been resolved. If you can render this file successfully, the fixes are working.

## Setup

```{r setup, message=FALSE, warning=FALSE}
# The theming should be automatically loaded via .Rprofile
# but we can also source it manually if needed
# source("theming/roboto_theme.R")

library(ggplot2)
library(palmerpenguins)

# Check that fonts are available
cat("Available font families:", paste(sysfonts::font_families(), collapse = ", "), "\n")
```

## Test Plot 1: Basic Scatter Plot

This should use the custom theme with proper fonts:

```{r basic-plot}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = body_mass_g), alpha = 0.7, size = 2) +
  labs(
    title = "Bill Dimensions of Palmer Penguins",
    subtitle = "This plot should use Vollkorn for title and Roboto Condensed for other text",
    caption = "Data: Palmer Penguins dataset",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Body Mass (g)"
  )
```

## Test Plot 2: Faceted Plot

Testing the theme with facets:

```{r faceted-plot}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = body_mass_g), alpha = 0.7) +
  facet_wrap(~species) +
  labs(
    title = "Bill Dimensions by Species",
    subtitle = "Faceted plot to test strip text formatting",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    color = "Body Mass (g)"
  )
```

## Font Test

The following should display different fonts if the theme is working:

- **Title text**: Should use Vollkorn font (serif-like)
- **Body text**: Should use Roboto Condensed (sans-serif)
- **Axis labels**: Should use Roboto Condensed
- **Code blocks**: Should use Fira Code (if available)

## Verification Checklist

If the theming system is working correctly, you should see:

- [ ] Plot titles in a serif font (Vollkorn)
- [ ] Axis text and labels in a condensed sans-serif font
- [ ] Consistent styling across all plots
- [ ] No errors about missing columns (bill_len, bill_dep, body_mass)
- [ ] Font families listed in the setup chunk

## Troubleshooting

If you encounter issues:

1. **Missing packages**: Run `source("setup_packages.R")` to install required packages
2. **Font errors**: Check internet connection and try restarting R
3. **Column errors**: Ensure you're using the latest version of the files
4. **Theme not applied**: Manually run `source("theming/roboto_theme.R")`

## Technical Details

The fixes implemented:

1. **Column Names**: Changed incorrect column references:
   - `bill_len` → `bill_length_mm`
   - `bill_dep` → `bill_depth_mm`  
   - `body_mass` → `body_mass_g`

2. **Font System**: Created robust theming with:
   - Google Fonts integration via showtext
   - Graceful fallbacks when fonts unavailable
   - Automatic theme application via .Rprofile

3. **Error Handling**: Added proper error handling for missing packages and network issues