---
title: "Making Diagrams in R"
subtitle: "Using ggdiagram and ggforce"
date: 2025/8/30
date-modified: today
filter:
  - d2
order: 70
bibliography: 
  - grateful-refs.bib
draft: true
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)  
library(ggformula)
library(ggdiagram)   ## create diagrams with ggplot2
library(plotly) # Add interactivity
# pak::pak(seankross/p5)
library(p5) 
```

```{r}
#| label: Themes and Extra Packages
#| echo: false
#| message: false
#| 
library(checkdown)
library(epoxy)
library(explore) # fake data generation
library(grateful)
##
library(downloadthis)
library(ggrepel)
library(marquee)
#devtools::install_github("mccarthy-m-g/embedr")
library(embedr) # Embed multimedia in HTML files
library(systemfonts) ## use custom fonts (need to be installed on your OS) 
library(showtext)   ## add google fonts to plots
library(scico)       ## scico color palettes(http://www.fabiocrameri.ch/colourmaps.php) in R 
library(ggtext)      ## add improved text rendering to ggplot2
library(ggforce)     ## add missing functionality to ggplot2

```

## Introduction

We may have been using `ggplot2` for a while now, but sometimes we need something that is not data-dependent: we need to make diagrams to show processes, workflows, relationships, etc.

Annotations help to draw attention to specific parts or the graph, specific areas of interest that are useful in telling the story .

We have already seen the `annotate()` command from `ggplot2` first ( and the new `gf_annotate()` from `ggformula`). Here we will look at some of the intriguing features from the `ggdiagram` package. We will also explore the `ggforce` package, and see if we can make our diagrams interactive with `plotly`. 

And then we will add a dash of JavaScript to R, using Sean Kross' `p5` package, that allows `p5.js` code to be written in R-style!!! `r emoji::emoji("cool")`


### Fonts and Themes

```{r}
#| label: add-local-fonts
#| code-fold: true
#| message: false
#| warning: false
#| 
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Anton",
  regular = "../../../../../../fonts/Anton-Regular.ttf") # Only one style is available

sysfonts::font_add(family = "RobotoCondensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")

sysfonts::font_add(family = "IbarraNova", 
  regular = "../../../../../../fonts/IbarraRealNova-Regular.ttf",
  bold = "../../../../../../fonts/IbarraRealNova-Bold.ttf",
  italic = "../../../../../../fonts/IbarraRealNova-Italic.ttf",
  bolditalic = "../../../../../../fonts/IbarraRealNova-BoldItalic.ttf")

sysfonts::font_add(family = "Tangerine", 
  regular = "../../../../../../fonts/Tangerine-Regular.ttf",
  bold = "../../../../../../fonts/Tangerine-Bold.ttf") # only these two are available

sysfonts::font_add(family = "Schoolbell", 
  regular = "../../../../../../fonts/Schoolbell-Regular.ttf") # Only regular is available

showtext_auto() # set these fonts as default for the session
# sysfonts::font_families() # check which fonts are available in the session

```

## Set the Theme
```{r}
#| label: set-ggplot2-theme
#| code-fold: true
#| 

theme_custom <- function(){ 

    theme_bw(base_size = 10) + 
    
    theme_sub_axis(title = element_text(family = "Roboto Condensed", 
                                       size = 8),
                   text = element_text(family = "Roboto Condensed", 
                                       size = 6)) + 
    
    theme_sub_legend(text = element_text(family = "Roboto Condensed", 
                                         size = 6),
                     title = element_text(family = "Alegreya", 
                                          size = 8)) + 
    
    theme_sub_plot(title = element_text(family = "Alegreya", 
                                        size = 14, face = "bold"),
                   title.position = "plot",
                   subtitle = element_text(family = "Alegreya", 
                                           size = 10),
                   caption = element_text(family = "Alegreya", 
                                          size = 6),
                   caption.position = "plot")
    
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

ggplot2::update_geom_defaults(geom = "marquee", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "text_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

## tinytable options
options("tinytable_tt_digits" = 2)
options("tinytable_format_num_fmt" = "significant_cell")
options(tinytable_html_mathjax = TRUE)


## Set defaults for flextable
flextable::set_flextable_defaults(font.family = "Roboto Condensed")


```

## {{< iconify openmoji japanese-symbol-for-beginner >}} Introduction

`{{ggdiagram}}` is a relatively new package in R that exploits some of the more recent developments in the R core. It uses for example, a new Object Oriented System in R called [S7](https://rconsortium.github.io/S7/) that allows commands to be "cascaded" in a way similar to Python and Javascript. Its best feature is that it allows precise control of shapes, their locations and connections, and returns a `ggplot` chart at the end, allowing us to use all the ggplot skills and bells and whistles, should we need them. 

The webpage for `{{ggdiagram}}` is at <https://wjschne.github.io/ggdiagram/>


## Method

`{{ggdiagram}}` allows to create diverse shapes, and then connect them together as we need. 

The method we will use is as follows: `Define -> Place -> Connect`:

- pre-create the shapes we need and save them as variables
- place the shapes in relation to each other using `ggdiagram::place()`;
- invoke `ggdiagram()` and then `ggdiagram::connect()` the shapes we have created in the and manner that we need. 


### Basic Shapes in `{{ggdiagram}}`

:::: {.columns}
::: {.column}
```{r}
#| label: fig-circle-1
#| eval: false

c1 <- ob_circle()
ggdiagram() + c1

```
:::

::: {.column}
```{r}
#| ref-label: fig-circle-1
#| echo: false
```

:::

::::

:::: {.columns}
::: {.column}
```{r}
#| label: fig-square-1
#| eval: false

s1 <- ob_rectangle()
ggdiagram() + s1

```
:::

::: {.column}
```{r}
#| ref-label: fig-square-1
#| fig-cap: Square
#| echo: false
```

:::

::::

:::: {.columns}
::: {.column}
```{r}
#| label: fig-releaux-1
#| eval: false

r1 <- ob_reuleaux(fill = "white")
ggdiagram() + r1

```
:::

::: {.column}
```{r}
#| ref-label: fig-releaux-1
#| fig-cap: Releaux Square!
#| echo: false
```

:::

::::


:::: {.columns}
::: {.column}
```{r}
#| label: fig-array-1
#| eval: false

ca1 <-  ob_circle() |>
  ob_array(k = 5, # Creating 5 circles
           label = 1:5,
           where = "southeast", # Array sloping down and right
           sep = .5) 
ggdiagram() + ca1

```
:::

::: {.column}
```{r}
#| ref-label: fig-array-1
#| fig-cap: Circle Array
#| echo: false
```

:::

::::


### Placement



### Connecting Shapes

:::: {.columns}
::: {.column}
```{r}
#| label: fig-connection-1
#| eval: false

# 1) Define shapes
start <- ob_circle(radius = 0.5)
decision <- ob_rectangle(width = 2.0, height = 1.0)
end <- ob_circle(radius = 0.5)

# 2) Place shapes
decision <- decision %>%  place(from = start, where = "right", sep = 1.5)
end <- end %>%  place(from = decision, where = "right", sep = 1.5)

# 3) Draw diagram
ggdiagram() +
  start +
  decision +
  end +
  connect(start, decision) +
  connect(decision, end) +
  coord_fixed() +
  theme_void()
```
:::

::: {.column}
```{r}
#| ref-label: fig-connection-1
#| fig-cap: Only Connect!!
#| echo: false
```

:::

::::

### Adding Text


### Math Markup


### Adding Diagram Content to Charts


### Using Outputs from Stats Pacakges


## Is Animation possible?


## {{< iconify ooui references-ltr >}} References

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
library(grateful)
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("ggdiagram", "ggforce", "plotly")) %>%
  knitr::kable(format = "simple")

```

:::

