---
title: <iconify-icon icon="mingcute:tyre-fill" width="24" height="24"></iconify-icon> ANOVA - Tyre Brands and Mileage
order: 10
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggprism) # Interesting Categorical Axes
library(ggridges)
library(supernova)
#devtools::install_github('cttobin/ggthemr')
library(ggthemr)
library(ggsci)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(ggthemr)
library(ggsci)
library(paletteer)
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 


```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
    
    theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```


```{r}
#| code-fold: true
#| include: false
# Chunk options
knitr::opts_chunk$set(
 fig.width = 7,
 fig.asp = 0.618, # Golden Ratio
 #out.width = "80%",
 fig.align = "center"
)
##
## https://stackoverflow.com/questions/36476751/associate-a-color-palette-with-ggplot2-theme
## 
my_colours <- c("#fd7f6f", "#7eb0d5", "#b2e061", "#bd7ebe", "#ffb55a", "#ffee65", "#beb9db", "#fdcce5", "#8bd3c7")
my_pastels <- c("#66C5CC","#F6CF71","#F89C74","#DCB0F2","#87C55F","#9EB9F3","#FE88B1","#C9DB74","#8BE0A4","#B497E7","#D3B484","#B3B3B3")
my_greys <- c("#000000", "#333333", "#666666","#999999", "#cccccc")
my_vivids <- c("#E58606","#5D69B1","#52BCA3","#99C945","#CC61B0","#24796C","#DAA51B","#2F8AC4","#764E9F","#ED645A","#CC3A8E","#A5AA99")

my_bolds <- c("#7F3C8D","#11A579","#3969AC","#F2B701","#E73F74","#80BA5A","#E68310","#008695","#CF1C90","#f97b72","#4b4b8f","#A5AA99")

font <- "Roboto Condensed"
mytheme <- theme_classic(base_size = 14) + ### %+replace%    #replace elements we want to change
  
  theme(
    text = element_text(family = font),
    panel.grid.minor = element_blank(),
    #text elements
    plot.title = element_text(family = font,
      face = 'bold',
      hjust = 0, #left align
      #vjust = 2 #raise slightly
      margin = margin(0, 0, 10, 0)
    ),
    
    plot.subtitle = element_text(
      family = font,
      hjust = 0,
      margin = margin(2, 0, 5, 0)
    ),
    
    plot.caption = element_text(
      family = font,
      size = 8,
      hjust = 1
    ),
    #right align
    
    axis.title = element_text(#axis titles
      family = font, #font family
      size = 10), #font size
    axis.text = element_text(#axis text
        family = font, #axis family
        size = 8) #font size
    )
theme_av <- list(
  mytheme,
  scale_colour_manual(values = my_bolds, aesthetics = c("colour", "fill"))
)
  
```


## Introduction

This is a dataset pertaining to tyres from different companies and their lifetime mileages.

## Data

```{r}
#| label: read-data
#| echo: false
#| message: false

tyre <- read_csv("../../../../../materials/Data/tyre.csv")
tyre_modified <- tyre %>% 
  mutate(Brands = as_factor(Brands))
tyre_modified
```

## Download the Modified data

```{r}
#| echo: false

library(downloadthis)

tyre_modified %>% 
   downloadthis::download_this(output_name = "tyre", 
      output_extension = ".csv", button_label = "Tyre Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```
## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Plot the Data

```{r}
#| echo: false
#
##
tyre_modified %>% 
  gf_boxplot(Mileage ~ Brands, fill = ~ Brands,
             title = "Tyre Brands and their Mileages",
             caption = "Note the segmented Categorical X-axis made with ggprism") %>% 
  
  gf_theme(theme_av) %>% 

  gf_refine(scale_x_discrete(guide = "prism_bracket"),
            guides(fill = guide_legend(title = "Tyre Brands")))
```

## Task and Discussion: ANOVA
1. Complete the pre-analysis steps for ANOVA

Write in. 

### Model + Table

2. Create the ANOVA model
3. Create the ANOVA table using the `supernova` package

```{r}
#| echo: false
model <- aov(Mileage ~ Brands, data = tyre)
model
supernova::supernova(model)
```


### Post-hoc Analysis and Plots

4. Compute the post-hoc differences in means and plot the pair-wise difference plots

```{r}
#| echo: false
# Set graph theme
theme_set(new = mytheme)
pairwise_model <- model %>% supernova::pairwise(correction = "Tukey", 
                              alpha = 0.05, var_equal = TRUE, 
                              plot = FALSE) 
pairwise_model

plot(pairwise_model)

```


### Conclusion
5. State a conclusion about the effect of `Brands` on `Mileage`.

Write in. 
