---
title: "Changing Scales on Charts"
subtitle: "Percentages, Ratios, and such like"
date: 2025/8/30
date-modified: "`r Sys.Date()`"
order: 50
code-fold: true
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)  
library(ggformula)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(paletteer)  ## access many color palettes
library(showtext)   ## add google fonts to plots
library(kableExtra) ## for nice tables

library(scales)     ## scale functions for visualization. Part of tidyverse.
```

## Introduction

We may have been using `ggplot2` for a while now, learning how to add fonts, colours, and annotations. Here we will look at changing the scales on the axes of our graphs and plots. Scales such as percentages, ratios, and other such like are needed when the quantities depicted demand it, when the data variabls on the axes are not simple counts or measurements. 

Here we will look at some of the neatfeatures from the `scales` package. 

### Fonts and Themes


```{r}
#| label: add-local-fonts
#| code-fold: true
#| 
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Anton",
  regular = "../../../../../../fonts/Anton-Regular.ttf") # Only one style is available

sysfonts::font_add(family = "RobotoCondensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")

sysfonts::font_add(family = "IbarraNova", 
  regular = "../../../../../../fonts/IbarraRealNova-Regular.ttf",
  bold = "../../../../../../fonts/IbarraRealNova-Bold.ttf",
  italic = "../../../../../../fonts/IbarraRealNova-Italic.ttf",
  bolditalic = "../../../../../../fonts/IbarraRealNova-BoldItalic.ttf")

sysfonts::font_add(family = "Tangerine", 
  regular = "../../../../../../fonts/Tangerine-Regular.ttf",
  bold = "../../../../../../fonts/Tangerine-Bold.ttf") # only these two are available

sysfonts::font_add(family = "Schoolbell", 
  regular = "../../../../../../fonts/Schoolbell-Regular.ttf") # Only regular is available

showtext_auto() # set these fonts as default for the session

# sysfonts::font_families() # check which fonts are available in the session

```


```{r}
#| label: set-ggplot2-theme
#| code-fold: true
#| 

theme_custom <- function(){ 
    font <- "RobotoCondensed"   # Assign font family up front
    
    theme_bw(base_size = 14, base_family = font) %+replace%    
      
      ## modify plot elements globally (for all following plots)
      theme(
        axis.ticks = element_line(color = "grey92"),
        axis.ticks.length = unit(.5, "lines"),
        
        panel.grid.minor = element_blank(),
        
        legend.title = element_text(family = "Schoolbell", size = 12),
        legend.text = element_text(family = "IbarraNova", color = "grey30"),
        
        plot.title = element_text(               #title
                   family = "Alegreya",          #set font family
                   size = 18,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                  #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      
        plot.title.position = "plot",
        
        plot.subtitle = element_text(family = "Alegreya", size = 12, 
                                     color = "grey30",
                                     hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
       plot.caption = element_text(             #caption
                   family = font,                #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
       
      plot.caption.position = "plot",            #right align to plot
)
}

## Set the theme
ggplot2::theme_set(new = theme_custom())


## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

```

## Scales: Breaks and Labels

The `scales` package provides functions that are useful for scaling axes and legends in `ggplot2`. It is part of the `tidyverse` and is automatically loaded when you load `ggplot2`.

There are several parts to scaling: the `breaks`, the `labels`, and the `transformations`. 

`Breaks`: These are the points at which ticks appear on an axis. You can specify breaks manually or use functions to generate them automatically.

`Labels`: These are the text labels that appear at the breaks. You can customize the labels to display values in different formats, such as `percentages`, `currency`, `dates`, or scientific notation.

`Transformations`: These are functions that can be applied to the data before plotting, such as `logarithmic` or `square root` transformations.

Scales can be applied all the `aesthetics` in ggplot: `x`, `y`, `color`, `fill`, `size`, `shape`, and even some esoteric ones `area`, `alpha`, `linewidth` etc.

## Case Study: gss-wages

We will work with datasets that we have seen in earlier modules, datasets that deal with variables like currencies, or which vary over several multiples.

Let us take an income dataset from [Vincent Arel-Bundock's website.](https://vincentarelbundock.github.io/Rdatasets/articles/data.html). We will use the dataset `gss_wages` which is available here.<https://vincentarelbundock.github.io/Rdatasets/csv/stevedata/gss_wages.csv>

As always, we will start with a **table** of the data:

```{r}
#| label: tbl-read-gsswages-data
#| tbl-cap: "GSS Wages Dataset"
#| message: false

gss_wages <- read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/stevedata/gss_wages.csv") %>% drop_na()

## Create a nicely formatted table
## uses `kableExtra` package
## 
gss_wages 
# 
# %>%
#   kableExtra::kbl() %>%
#   kableExtra::kable_paper(full_width = FALSE) %>%
#   kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
#   kableExtra::scroll_box(height = "350px")

```

## Basic Plot

Let us create a basic scatter plot of `realrinc` vs `year` and colour by `gender`, and another boxplot of `realrinc` by `gender`:


```{r}
#| label: fig-basic-ggformula-plot
#| fig-cap: "Basic Plots"
#| fig-subcap: 
#| - "Scatter Plot"
#| - "Box Plot"
#| layout-ncol: 2


p1 <- gf_point(data = gss_wages, realrinc ~ year, 
              color = ~ gender, alpha = 0.8, size = 1) %>% 
  gf_labs(y = "Income", x = "Year", title = "GSS-Wages Dataset") %>% 

  gf_refine(scale_color_brewer(palette = "Set1"))
p1

p2 <- gf_boxplot(data = gss_wages, realrinc ~ gender, 
              color = ~gender, fill = ~ gender, alpha = 0.8) %>% 
  gf_labs(y = "Income", x = "Gender", title = "GSS-Wages Dataset") %>% 

  gf_refine(scale_color_brewer(palette = "Set1"))
p2

```

Let us now change the y-axis to show income in thousands of dollars. We will use the `scales` package for this.

```{r}
#| label: fig-dollar-scale
#| fig-cap: "Dollar Scale with Automatic Breaks"
#| fig-subcap: 
#| - "Scatter Plot"
#| - "Box Plot"
#| layout-ncol: 2
#| 
p1 %>% 
  gf_refine(scale_y_continuous(labels = scales::label_currency(scale = 0.001, suffix = "K")))


p2 %>% 
  gf_refine(scale_y_continuous(labels = scales::label_currency(scale = 0.001, suffix = "K")))

```
The `scale` function `label_currency` has several arguments that can be used to customize the display of currency values. Here we have used `scale = 0.001` to convert the values to thousands and `suffix = "K"` to add a "K" suffix to the labels.

We may yet be dissatisfied with this plot and wish to change the y-axis to a logarithmic scale. This is useful when the data spans several orders of magnitude, as is often the case with income data. We will also, for demonstration purposes, setup an excessively detailed y-scale with far too many breaks!!

```{r}
#| label: fig-log-scale
#| fig-cap: "Logarithmic Scale with (Excessively ) Detailed Breaks"
#| fig-subcap: 
#| - "Scatter Plot"
#| - "Box Plot"
#| layout-ncol: 2

p1 %>% 
  gf_refine(scale_y_log10(labels = scales::label_currency(scale = 0.001, 
                                                          suffix = "K"),
            breaks = scales::breaks_log(n = 20, base = 10)))
p2 %>% 
  gf_refine(scale_y_log10(labels = scales::label_currency(scale = 0.001, suffix = "K"), breaks = scales::breaks_log(n = 20, base = 10)))

```


## Case Study: Gapminder

We will work with the `gapminder` dataset from the `gapminder` package. This dataset contains data on life expectancy, GDP per capita, and population for various countries over several years.

```{r}
#| label: tbl-read-gapminder-data
#| tbl-cap: "Gapminder Dataset"

gapminder <- gapminder::gapminder
gapminder

```
There are several variables, most of them "man-made", which tend to vary over serveral orders of magnitude. Let us see how we can depict those. 

```{r}
#| label: fig-gapminder-base-plot
#| fig-cap: "Basic Gapminder Plot"
#| fig-subcap:
#| - "Scatter Plot"
#| - "Box Plot"

p1 <- gf_point(data = gapminder, gdpPercap ~ year, 
              color = ~ continent, alpha = 0.8, size = 1) %>% 
  gf_labs(y = "GDP per Capita", x = "Year", title = "Gapminder Dataset") %>% 

  gf_refine(scale_color_brewer(palette = "Set1"))
p1
p2 <- gf_boxplot(data = gapminder, gdpPercap ~ continent, 
              color = ~continent, fill = ~ continent, alpha = 0.8) %>% 
  gf_labs(y = "GDP per Capita", x = "Continent", title = "Gapminder Dataset") %>% 

  gf_refine(scale_color_brewer(palette = "Set1"))

p2

```


## References

1. R package `scales`: <https://scales.r-lib.org/>
1. 
