---
title: <iconify-icon icon="iconoir:sea-waves"></iconify-icon> <iconify-icon icon="mdi:weed"></iconify-icon> Seaweed Nutrients
order: 250
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggbump)
library(ggprism)
library(paletteer) # fancy colour palettes!
```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
# # 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
#   
```


#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
    
    theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```

## Introduction

Nine types of Seaweed were rated on different parameters and charted as shown below. 


## Read the Data

{{< downloadthis ../../../../../materials/Data/Seaweed_Farming_Marine_Permaculture.xls dname=seaweed label="Download the Seaweed data"
icon=database-fill-down type=info class=data-button >}}

::: callout-note
### Excel Data
The data is an excel sheet. Inspect it first in Excel and  decide which sheet you need, and which part of the data you need. There are multiple sheets!
Then use `readxl::read_xlsx(..)` to read it into R.
NOTE: The sheet that contains our data of interest is titled "seaweed nutrition", range = "A3:R13". 
:::


```{r}
#| label: read-data
#| echo: false
#| message: false

seaweed <- readxl::read_xlsx("../../../../../materials/Data/Seaweed_Farming_Marine_Permaculture.xls", sheet = "seaweed nutrition", range = "A3:R13")

# seaweed %>%
#   downloadthis::download_this(
#     output_name = "seaweed-nutrition",
#     output_extension = ".csv",
#     button_label = "Download Seaweed Nutrition data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )

```

## Inspect the Data

```{r}
#| label: inspect-categorical
#| echo: false
glimpse(seaweed)

```


## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::

## Research Question
::: callout-note
Write in! First look at the chart below!
:::

## Analyse/Transform the Data

```{r}
#| label: data-preprocessing
#| echo: fenced
#
# Write in your code here 
# to prepare this data as shown below  
# to generate the plot that follows

```

```{r}
#| label: seaweed-long
#| echo: false
seaweed %>% 
  drop_na() %>% 
  dplyr::select("common_name" = `common name`, 
         "total_fibre" = `Total fiber`, Carbohydrates, 
         Calcium, Potassium, Zinc) %>% 
  mutate(fibre_rank = dense_rank(desc(total_fibre)),
         carbo_rank = dense_rank(desc(Carbohydrates)),
         calcium_rank = dense_rank(desc(Calcium)),
         potassium_rank = dense_rank(desc(Potassium)),
         zinc_rank = dense_rank(desc(Zinc))) %>% 
  select(common_name, contains("rank")) %>% 
  
  # Convert to Long Form
  pivot_longer(cols = contains("rank"), 
               names_to = "parameter", 
               values_to = "ranks") %>% 
  arrange(parameter) -> seaweed_bump

seaweed_bump

```


## Plot the Data
```{r}
#| echo: false
seaweed_bump %>%
  ggplot(
    aes(
      x = parameter,
      y = ranks,
      group = common_name,
      colour = common_name,
      shape = common_name
    )
  ) +
  ggbump::geom_bump(linewidth = 1) +
  
  scale_x_discrete(labels = c("Calcium", "Carbs", "Fibre", "Potassium", "Zinc"), guide = "prism_bracket") +
  geom_point(size = 3, fill = "black") +
  scale_shape_manual("Seaweed Type", values = c(0, 1, 2, 5, 7, 8, 15, 16, 18)) +
 scale_colour_paletteer_d(name = "Seaweed Type", 
                          palette = "tvthemes::rickAndMorty",
                          dynamic = FALSE) +
  scale_y_reverse(breaks = c(1:9)) +
  scale_linetype_manual("Seaweed Type", values = c(1, 2, 3, 4, 5, 6, 1, 2, 3)) +
  
  labs(
    title = "Nutrients from Seaweed",
    subtitle = "Where can I get Zinc from?",
    caption = "Data from informationisbeautiful.net | Colour Palette = tvthemes::rickAndMorty",
    x = "Nutrient",
    y = "Rank"
  ) +
  theme_custom()

```

## Tasks and Discussion

- Complete the Data Dictionary.
- Select and Transform the variables as shown.
- Create the graphs shown and discuss the following questions:
  -   Identify the type of charts
  -   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
  -   What research activity might have been carried out to obtain the data graphed here? Provide some details.
  -   What might have been the Hypothesis/Research Question to which the response was Chart?
  - Write a 2-line story based on the chart, describing your inference/surprise.
  - Based on the diagram, discuss which one an elderly person might try if they are deficient in calcium. If you were trying to avoid carbs, which seaweed sushi would you try?

    
## References

Over 2500 colour palettes are available in the `paletteer` package. Can you find `tayloRswift`? `wesanderson`? `harrypotter`? `timburton`?

Here are the Qualitative Palettes: 
```{r}
#| echo: false
palettes_d_names %>% reactable::reactable(data = ., filterable = TRUE, minRows = 10)
```

And the Quantitative/Continuous palettes:

```{r}
#| echo: false
palettes_c_names %>% reactable::reactable(data = ., filterable = TRUE, minRows = 10)
```
Use the commands:

```{r}
#| eval: false
#| echo: true

## For Qual variable-> colour/fill: 
scale_colour_paletteer_d(name = "Legend Name", 
                          palette = "package::palette",
                          dynamic = TRUE/FALSE)
                          
## For Quant variable-> colour/fill: 
scale_colour_paletteer_c(name = "Legend Name", 
                          palette = "package::palette",
                          dynamic = TRUE/FALSE)

```

      
