---
title: "Annotations: Drawing Attention to Parts of the Graph"
subtitle: ""
date: 2021/7/30
date-modified: "`r Sys.Date()`"
order: 35
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)  
library(ggformula)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(scico)       ## scico color palettes(http://www.fabiocrameri.ch/colourmaps.php) in R 
library(ggtext)      ## add improved text rendering to ggplot2
library(ggforce)     ## add missing functionality to ggplot2

library(showtext)   ## add google fonts to plots

```

## Introduction

We may have been using `ggplot2` for a while now, but there is always more to learn about the aesthetics of our plots. Here we will look at annotating our graphs and plots. 

Annotations help to draw attention to specific parts or the graph, specific areas of interest that are useful in telling the story .

We have already seen the `annotate()` command from `ggplot2` first ( and the new `gf_annotate()` from `ggformula`). Here we will look at some of the intrigung features from the `ggforce` package. 

### Fonts and Themes

As we learnt in the [Fonts Module](../20-Fonts/index.qmd), let us set the ggplot theme with fonts for both the chart defining elements **and** our anotations.

```{r}
#| label: add-fonts
#| echo: fenced
#| cache: true
#| code-fold: true
font_add_google(name = "Gochi Hand", family = "gochi")
font_add_google("Schoolbell", "bell")
font_add_google("Galada", "galada")
font_add_google("Roboto Condensed", "roboto-condensed")
font_add_google("Noto Sans", "noto")
font_add_google("Uchen", "uchen")
font_add_google("Ibarra Real Nova", "ibarra")
font_add_google("Open Sans", "open")
font_add_google("Anton", "anton")
font_add_google("Tangerine", "tangerine")
showtext_auto() # set the google fonts as default

```

```{r}
#| label: set-ggplot2-theme
#| code-fold: true
theme_set(theme_classic(base_size = 14, 
                        base_family = "roboto-condensed"))

## modify plot elements globally (for all following plots)
theme_update(
  axis.ticks = element_line(color = "grey92"),
  axis.ticks.length = unit(.5, "lines"),
  panel.grid.minor = element_blank(),
  legend.title = element_text(size = 12, family = "ibarra"),
  legend.text = element_text(color = "grey30"),
  plot.title = element_text(size = 18, face = "bold"),
  plot.subtitle = element_text(size = 12, color = "grey30", family = "ibarra"),
  plot.caption = element_text(size = 9, margin = margin(t = 15), family = "anton")
)
# set-geom-default-font

update_geom_defaults(geom = "text", 
                     new = list(family = "roboto-condensed"))

```


## Data

We will again work with a **familiar** dataset, so that we can concentrate on the chart aesthetics, without having to spend time getting used to the data: the `penguins` dataset again, ~~from the `palmerpenguins` package~~ now part of the R installation!. 

As always, we will start with a **table** of the data:

```{r}
#| label: read-penguins-data
#| message: false

penguins <- penguins %>% drop_na() # remove data containing missing data

## Create a nicely formatted table
## uses `kableExtra` package
## 
penguins %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
  kableExtra::scroll_box(width = "700px", height = "500px")

```

## Basic Plot

A basic scatter plot, which we will annotate: 

```{r}
#| label: basic-ggformula-plot
#| 
p1 <- gf_point(data = penguins, bill_depth_mm ~ bill_length_mm, 
              color = ~ body_mass_g, alpha = 0.8, size = 3.5) %>% 

  ## custom colors from the scico package
  gf_refine(scico::scale_color_scico(palette = "tokyo", direction = -1)) %>% 
  
  ## custom labels
  gf_labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (Pygoscelis)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) PLoS ONE',
    x = 'Bill Length (mm)', 
    y = 'Bill Depth (mm)',
    color = 'Body mass (g)'
  )

p1

```



## Using `{ggforce}`

From [Thomas Lin Pedersen's website](www.ggforce.data-imaginist.com) → www.ggforce.data-imaginist.com

> `ggforce` is a package aimed at providing missing functionality to `ggplot2` through the extension system introduced with `ggplot2` v2.0.0. Broadly speaking `ggplot2` has been aimed primarily at explorative data visualization in order to investigate the data at hand, and less at providing utilities for composing custom plots a la `D3.js`. `ggforce` is mainly an attempt to address these “shortcoming” (design choices might be a better description). The goal is to provide a repository of geoms, stats, etc. that are as well documented and implemented as the official ones found in `ggplot2`.



We will start with the basic plot, with the fonts, labels and `ggtext` related work done up to now: First, let us update the theme to use the `ggtext` options, and then we will add the `ggforce` annotations.

```{r}
## Existing theme
theme_set(theme_classic(base_size = 14, 
                        base_family = "roboto-condensed"))

## modify plot elements globally (for all following plots)
theme_update(
  axis.ticks = element_line(color = "grey92"),
  axis.ticks.length = unit(.5, "lines"),
  panel.grid.minor = element_blank(),
  legend.title = element_text(size = 12, family = "ibarra"),
  legend.text = element_text(color = "grey30"),
  
## theme updates here
  plot.title = ggtext::element_markdown(),
  plot.caption = ggtext::element_markdown(),
  axis.title.x = ggtext::element_markdown(),
  axis.title.y = ggtext::element_markdown()
)

```




```{r}
## plot that we will annotate with ggforce afterwards
p3 <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                           colour = body_mass_g)) + 
  geom_point(size = 3.5) + 
  coord_cartesian(xlim = c(25, 65), ylim = c(10, 25)) +
  
  # Add colours
  scico::scale_color_scico(palette = "bamako", direction = -1) +
  rcartocolor::scale_fill_carto_d(palette = "Bold") +
  
  # Add labels
  labs(
    title = "Bill Dimensions of Brush-Tailed Penguins (*Pygoscelis*)",
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = "Data: Gorman, Williams & Fraser (2014) *PLoS ONE*",
    x = "**Bill Length** (mm)", 
    y = "**Bill Depth** (mm)",
    color = "Body mass (g)",
    fill = "Species"
  )
p3

```

### Using ellipses
```{r}
#| label: ellipses
## ellipsoids for all groups
p3 +
  # penguins data is inherited automatically
  # Can use other data also if needed
  ggforce::geom_mark_ellipse( 
    aes(fill = species, 
        label = species), 
    alpha = .15, 
    show.legend = FALSE
  ) +
  
  # Repeat the points!
  # Why? So the ellipses don't hide them!
  geom_point(aes(color = body_mass), 
             alpha = .6, 
             size = 3.5)


```

### Highlighting one group

#### Add an Ellipse
```{r}

## ellipsoids for specific subset
p3 +
  ggforce::geom_mark_ellipse(
    aes(fill = species, 
        label = species, 
        filter = species == 'Gentoo'), # Very Curious syntax here!!
    alpha = 0.25, 
    show.legend = FALSE
  ) +
 geom_point(aes(color = body_mass), 
            alpha = .6, 
            size = 3.5)

```


#### Adding Circles
```{r}

## circles
p3 +
  ggforce::geom_mark_circle(
    aes(fill = species, label = species, 
        filter = species == 'Gentoo'), 
    alpha = 0.25, 
    show.legend = FALSE
  ) +
  geom_point(aes(color = body_mass), 
             alpha = .6, 
             size = 3.5)

```


#### Adding Rectangles
```{r}

## rectangles
p3 +
  ggforce::geom_mark_rect(
    aes(fill = species, 
        label = species, 
        filter = species == 'Gentoo'), 
    alpha = 0.25, 
    show.legend = FALSE
  ) +
  geom_point(aes(color = body_mass), 
             alpha = .6, 
             size = 3.5) 

```


#### Adding a Convex Hull
```{r}
library(concaveman)
## hull
p3 +
  ggforce::geom_mark_hull(
    aes(fill = species, 
        label = species, 
        filter = species == 'Gentoo'), 
    alpha = 0.25, 
    show.legend = FALSE
  ) +
  geom_point(aes(color = body_mass), 
             alpha = .6, 
             size = 3.5)

```

