---
title: <iconify-icon icon="simple-icons:ikea" width="1.2em" height="1.2em"></iconify-icon> Ikea Furniture
order: 10
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(ggridges)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 


```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    
    theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```






## Introduction

This is a dataset pertaining to furniture prices at IKEA, modified for ease of analysis and plotting. 

## Data

```{r}
#| label: read-data
#| echo: false
#| message: false

#library(tidytuesdayR)
ikea <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-11-03/ikea.csv")
#tt <- tt_load("2020-11-03")
#ikea <- tt$ikea 
ikea %>%
  select(-...1) %>%
  mutate(price_usd = 0.27 * price,
         short_description = str_squish(short_description)) %>%
  add_count(category, name = "category_total") %>%
  mutate(
    category = glue("{ category } ({ category_total })"),
    category = fct_reorder(category, price_usd)
  ) %>%
  filter(category_total >= 150) -> ikea_modified
ikea_modified

```

## Download the Modified data

```{r}
#| echo: false

library(downloadthis)

ikea_modified %>% 
   downloadthis::download_this(output_name = "ikea-modified", output_extension = ".csv", button_label = "Ikea furniture Data", button_type = "default", icon = "fa fa-save")

# 
# mtcars %>%
#   downloadthis::download_this(
#     output_name = "mtcars data set",
#     output_extension = ".csv",
#     button_label = "Download data",
#     button_type = "warning",
#     has_icon = TRUE,
#     icon = "fa fa-save"
#   )
```

## Data Dictionary

::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Plot the Data

```{r}
#| echo: false
ggplot2::theme_set(new = theme_custom())

ikea_modified %>%
  ggplot(aes(price_usd, category, 
             fill = other_colors)) +
  geom_density_ridges(alpha = 0.75, aes(linetype = other_colors)) +
  scale_x_log10(labels = scales::label_dollar()) +
  
  scale_fill_brewer("Other Colors?", palette = "Set1") + 
  scale_linetype("Other Colors?") + 
  
  labs(
    x = "Price (USD)",
    y = "",
    title = "How much do items in each IKEA category cost?",
    subtitle = "Are they available in other colors?",
    caption = "TidyTuesday Dataset for 03/11/2020"
  )


```

## Task and Discussion

Complete the Data Dictionary.
Create the graph shown and discuss the following questions:

-   What is the kind of plot used in the chart?
-   What variables have been used in the chart?
-   What can you say about the scale on X-axis?
-   What can you say about prices of items that are available in *single colour* versus those that are available in more than one colour?
-   What is a good hypothesis to interpret the double-humped nature of some of the curves?

