---
title: "Permutation Tests for Linear Regression"
draft: true
date: 03/May/2023
date-modified: "`r Sys.Date()`"
categories: 
  - Linear Regression
  - Quantitative Predictor
  - Quantitative Response
  - Sum of Squares
  - Residuals
  - Permutation
abstract: "Using a Permutation Test to check our Regression Model"
filters:
  - nutshell
---

```{r}
#| label: setup
#| message: true
#| warning: false
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
library(tidyverse)
library(ggformula)
library(mosaic)
library(infer)

```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
library(tidyverse)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
        
    ggplot2::theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```



## Linear Regression using Permutation Tests

We wish to establish the significance of the effect size due to each of
the levels in `TempFac`. From the normality tests conducted earlier we
see that except at one level of `TempFac`, the times are are not
normally distributed. Hence we opt for a Permutation Test to check for
significance of effect.

As remarked in Ernst[^2], the non-parametric permutation test can be
both *exact* and also **intuitively easier** for students to grasp.
Permutations are easily executed in R, using packages such as
`mosaic`[^3].

We proceed with a Permutation Test for `TempFac`. We shuffle the levels
(13, 18, 25) randomly between the Times and repeat the ANOVA test each
time and calculate the F-statistic. The Null distribution is the
distribution of the F-statistic over the many permutations and the
p-value is given by the proportion of times the F-statistic equals or
exceeds that observed.

### Read the Data

```{r}
data("BostonHousing2", package = "mlbench")
housing <- BostonHousing2
inspect(housing)

```


We will use `mosaic` and also try with `infer`.

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}

#### Using mosaic

`mosaic` offers an easy and intuitive way of doing a repeated
permutation test, using the `do()` command. We will `shuffle` the
`TempFac` factor to jumble up the `Time` observations, 10000 times. Each
time we shuffle, we compute the F_statistic and record it. We then plot
the 10000 F-statistics and compare that with the real-world observation
of `F-stat`.

```{r}
#| label: permutation test for ANOVA with mosaic
#| include: false
#| eval: false

obs_F_stat <- frogs_anova %>% 
  broom::tidy() %>% 
  select(statistic)
observed_mosaic <- obs_F_stat$statistic[1]

null_dist_mosaic <- do(10000) * aov(Time ~ shuffle(TempFac), data = frogs_long)
null_dist_mosaic %>% head()

null_dist_mosaic %>% drop_na() %>% 
  select(F) %>% 
  gf_histogram(data = ., ~ F, 
               fill = ~ F >= observed_mosaic,
               title = "Null Distribution of ANOVA F-statistic" )


```

The Null distribution of the F_statistic under permutation shows it
never crosses the real-world observed value, testifying the strength of
the effect of `TempFac` on hatching `Time`. And the p-value is:

```{r}
#| include: false
#| eval: false
p_value <- mean(null_dist_mosaic$F >= observed_mosaic, na.rm = TRUE)
p_value

```

#### Using `infer`

We calculate the observed F-stat with `infer`, which also has a very
direct, if verbose, syntax for doing permutation tests:

```{r}
#| label: ANOVA F-statistic with infer
#| include: false
#| eval: false

observed_infer <- frogs_long %>% 
  specify(Time ~ TempFac) %>% 
  hypothesise(null = "independence") %>% 
  calculate(stat = "F")
observed_infer

```

We see that the observed F-Statistic is of course $385.8966$ as before.
Now we use `infer` to generate a NULL distribution using permutation of
the factor `TempFac`:

```{r}
#| label: Permutation using infer
#| include: false
#| eval: false

null_dist_infer <- frogs_long %>% 
  specify(Time ~ TempFac) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "F")

head(null_dist_infer)

null_dist_infer %>% 
  visualise()

```

As seen, the `infer` based permutation test also shows that the
permutationally generated F-statistics are nowhere near that which was
observed. The effect of `TempFac` is very strong.
:::
