---
title: "Fonts in ggplot"
subtitle: "Using fonts from your machine and from Google"
date: 2025/8/02
date-modified: "`r Sys.Date()`"
order: 20
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)   ## data science package collection (incl. the ggplot2 package)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(showtext)    ## add google fonts to plots
library(scico)       ## Colour palette
library(ggtext)      ## add improved text rendering to ggplot2
library(gfonts)      ## Use Google Fonts offline

```

## {{< iconify noto-v1 package >}} Setting up R Packages

We may have been using `ggplot2` for a while now, but there is always more to learn about the aesthetics of our plots. In this module, we will explore how to use different fonts and themes to make our plots look more polished and professional. 

We may also want to use fonts on our websites and we will look at using packages like `sysfonts` . `gfonts`, and `showtext` to add custom fonts to our plots and websites. For instance you may want your website to look this: [Project Peasants](https://project-peasants.netlify.app) !!

## Using Preferred Fonts

We may prefer specific fonts for **our plots**, and we may also want to use these fonts on our websites. There are (atleast) two ways in which one can use fonts in R: 

- **Live Downloading of Fonts**: Downloading fonts from Google **within the R session**, using the `showtext` package. 
- **Using Locally Installed Fonts**: Using fonts that are already installed/downloaded on your computer, also using the `sysfonts` package. 

::: callout-note
### New Package `systemfonts`
There is also a more recent, and comprehensive package called `systemfonts` We will postpone this for a later presentation (mainly because I am still figuring it out. `r emoji::emoji("sweat_smile")` )
:::


::: callout-important
### GDPR Compliance
Using fonts that are already installed on your computer is [**GDPR compliant**](https://gdpr-info.eu/). Using fonts *directly* from Google may not be, as it may involve sending data to Google servers every time your website is loaded.
:::


### Live-loading Fonts from Google

We will want to add a few new fonts from Google to our graphs. The best way (currently) is to use the `sysfonts` package (which we loaded above) to bring into our work fonts from Google. To view and select the fonts you might want to work with, spend some time looking over [Google Fonts](https://fonts.google.com/). 

Let us add a handful of fonts to our R session. We will use the `font_add_google()` function from the `sysfonts` package to download Google fonts into our session. These fonts will then be included into our session using `showtext_auto()`.

This needs to be included in each Quarto document on your website. 


```{{r}}
#| label: download-google-fonts
#| cache: true

## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
sysfonts::font_add_google(name = "Gochi Hand", family = "GochiHand")
font_add_google("Schoolbell", "Schoolbell")
font_add_google("Roboto Condensed", "RobotoCondensed")
font_add_google("Noto Sans", "Noto")
font_add_google("Jura", "Jura")
font_add_google("Ibarra Real Nova", "Ibarra")
font_add_google("Open Sans", "Open")
font_add_google("Anton", "Anton")
font_add_google("Tangerine", "Tangerine")
font_add_google("Fraunces", "Fraunces")

```

```{{r}}
#| label: enable-showtext

showtext_auto() # set these fonts as default for the session
sysfonts::font_families() # check which fonts are available in the session

```

Ensure that you use the right `name` for the font family, as shown above. It is text-sensitive. The `family` is up to you to choose. Using similar names for `name` and `family` is a good idea, as shown above.


## Adding local fonts from your computer

Using Google fonts as shown above means that every time someone loads your website on their browser, the browser goes off to Google to download the fonts on the fly. As mentioned above, this will typically make your website non-compliant with GDPR. 

One solution is to download fonts from Google *a priori*, or from elsewhere, and then import them into your session. The `sysfonts::font_add()` function allows us to do this. You will need to provide the path to the font files on your computer. That is our next step.

Now you need to import these fonts into ***each Quarto document*** individually ( This website has done that for every document !! ) Think of it as a theme for your document. You will need to provide the path to the font files on your computer. The `../` sequence navigates up one folder level. Adjust the path as needed, to go right from your Quarto document folder up to project `root` and then down to the `fonts` folder.

Let's assume we have gone off to [Google Fonts](https://fonts.google.com/) and downloaded the following fonts to our `fonts` folder: `Alegreya`,`Anton`, `Roboto Condensed`, `Schoolbell`, `Tangerine`, and `Ibarra Real Nova`.

```{r}
#| label: add-local-fonts
#| echo: true
#| 
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Anton",
  regular = "../../../../../../fonts/Anton-Regular.ttf") # Only one style is available

sysfonts::font_add(family = "RobotoCondensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")

sysfonts::font_add(family = "IbarraNova", 
  regular = "../../../../../../fonts/IbarraRealNova-Regular.ttf",
  bold = "../../../../../../fonts/IbarraRealNova-Bold.ttf",
  italic = "../../../../../../fonts/IbarraRealNova-Italic.ttf",
  bolditalic = "../../../../../../fonts/IbarraRealNova-BoldItalic.ttf")

sysfonts::font_add(family = "Tangerine", 
  regular = "../../../../../../fonts/Tangerine-Regular.ttf",
  bold = "../../../../../../fonts/Tangerine-Bold.ttf") # only these two are available

sysfonts::font_add(family = "Schoolbell", 
  regular = "../../../../../../fonts/Schoolbell-Regular.ttf") # Only regular is available

showtext_auto() # set these fonts as default for the session
sysfonts::font_families() # check which fonts are available in the session

```


After running the chunks above in your session, you can check which fonts are available in your session by typing `sysfonts::font_families()` in your console, as shown above. We see that apart from the R-default `sans`, `serif`, and `mono` fonts, we have added the several fonts to our session. 

## Plot Theme using Desired fonts

Now that we have the fonts downloaded, we can use them in our plots. We will use the `theme_set()` function to set a default theme for our plots, and then we will modify the theme elements to use our desired fonts.
Let us set some `ggplot2` theme aspects now!! Here is a handy picture showing ( most of ) the theme-able aspects of a ggplot plot. 

For more info, type `?theme` in your console.

```{r}
#| label: fig-ggplot-theme-elements
#| echo: FALSE
#| fig.alt: "Rosana Ferrero (@RosanaFerrero) on Twitter Sept 11, 2022"
#| fig.cap: "ggplot2 Theme Elements. Source: Rosana Ferrero (@RosanaFerrero) on Twitter Sept 11, 2022"
#| 
knitr::include_graphics("../../../../../materials/images/ggplot_theme_elements.jfif")

```
So now that we understand the elements of the themes, let us set our ggplot theme: 



```{r}
#| label: set-ggplot2-theme
#| 

theme_custom <- function(){ 
    font <- "RobotoCondensed"   # Assign font family up front
    
    theme_bw(base_size = 14, base_family = font) %+replace%    
      
      ## modify plot elements globally (for all following plots)
      theme(
        axis.ticks = element_line(color = "grey92"),
        axis.ticks.length = unit(.5, "lines"),
        
        panel.grid.minor = element_blank(),
        
        legend.title = element_text(family = "Schoolbell", size = 12),
        legend.text = element_text(family = "IbarraNova", color = "grey30"),
        
        plot.title = element_text(               #title
                   family = "Alegreya",          #set font family
                   size = 18,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                  #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      
        plot.title.position = "plot",
        
        plot.subtitle = element_text(family = "Alegreya", size = 12, 
                                     color = "grey30",
                                     hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
       plot.caption = element_text(             #caption
                   family = font,                #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
       
      plot.caption.position = "plot",            #right align to plot
)
}

## Set the theme
ggplot2::theme_set(new = theme_custom())


## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

```

::: callout-note
### Fonts in ggplot

Using `showtext(auto)` as shown above allows us to modify the ***non-graph*** elements of the ggplot chart: the `axes`, the `title`, the `legend` etc. If we wish to add text with fonts to the body of the chart itself, as with geoms such as `text`and `label` annotations , we need to do one more thing: use `update_geom_defaults()` as shown below:

:::

```{r}
#| label: set-geom-default-font
#| 
update_geom_defaults(geom = "text", new = list(family = "Schoolbell"))
update_geom_defaults(geom = "label", new = list(family = "Schoolbell"))

# repeat for other geoms like "text_repel" and any new text-using geom etc.
```


## Data

We will work with a **familiar** dataset, so that we can concentrate on the chart aesthetics, without having to spend time getting used to the data: the `penguins` dataset again, ~~from the `palmerpenguins` package~~ now part of the R installation!. 

As always, we will start with a **table** of the data:

```{r}
#| label: tbl-read-penguins-data
#| tbl-cap: "Penguins Dataset"
#| message: false
#| 
data(penguins, package = "datasets")
penguins <- penguins %>% drop_na() # remove data containing missing data

## Create a nicely formatted table
## uses `kableExtra` package
## 
penguins %>% 
  kableExtra::kbl(caption = "Penguins Dataset") %>%
  kableExtra::kable_paper(full_width = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
  kableExtra::scroll_box(width = "700px", height = "350px")

```


## Basic Plot

A basic scatter plot first, which we will decorate with fonts:

```{r}
#| label: basic-scatter-plot
#| fig.cap: "Scatter plot of bill depth versus bill length, colored by body mass."

## Set the theme
ggplot2::theme_set(new = theme_custom())

p1 <- ggplot(penguins, aes(x = bill_len, y = bill_dep)) +
  geom_point(aes(color = body_mass), alpha = .6, size = 3.5) + 
  
  ## custom colors from the scico package
  scico::scale_colour_scico(palette = "glasgow", direction = -1)+
  
  ## custom labels
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (Pygoscelis)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) PLoS ONE',
    x = 'Bill Length (mm)', 
    y = 'Bill Depth (mm)',
    color = 'Body mass (g)'
  )

p1

```

We see that the ggplot theme uses the fonts that we have assigned to the different elements in the graph. 


## Using `{ggtext}`

From [Claus Wilke's website](https://wilkelab.org/ggtext/) → www.wilkelab.org/ggtext

> The `ggtext` package provides simple Markdown and HTML rendering for ggplot2. Under the hood, the package uses the `gridtext` package for the actual rendering, and consequently it is limited to the feature set provided by `gridtext`.  
Support is provided for Markdown both in theme elements (plot titles, subtitles, captions, axis labels, legends, etc.) and in geoms (similar to `geom_text()`). In both cases, there are two alternatives, one for creating simple text labels and one for creating text boxes with word wrapping.

NOTE: on some machines, the ggtext package may not work as expected. In this case, please do as follows, using your Console:

- remove gridtext: `remove.packages(gridtext)`.   
- Install development version of `gridtext`: `remotes::install_github("wilkelab/gridtext")`

### `element_markdown()`

We can use our familiar markdown syntax right inside the titles and captions of the plot. `element_markdown()` is a theming command made available by the `ggtext` package. 

`element_markdown()` → formatted text elements, e.g. titles, caption, axis text, striptext

```{r}
#| label: ggtext-element-markdown
#| 
## Set the theme
ggplot2::theme_set(new = theme_custom())

p1 + 
  
  ## New code starts here: Two Step Procedure with ggtext
  ## 1. Markdown formatting of labels and title, using asterisks
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (*Pygoscelis*)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) *PLoS ONE*',
    x = '**Bill Length** (mm)', 
    y = '**Bill Depth** (mm)',
    color = 'Body mass (g)'
  ) +
   
  ## 2. Add theme related commands from ggtext
  ## render respective text elements
  theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown()
  )


```

### `element_markdown()` in combination with HTML

This allows us to change fonts and font-effects in titles, labels, and captions:


```{r}
## use HTML syntax to change text color
p1 + 
  labs(title = 'Bill Dimensions of Brush-Tailed Penguins <i style="color:#28A87D;">Pygoscelis</i>') +
  theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)) 
  
## use HTML syntax to change font and text size
## 
p1 + 
 labs(title = 'Bill Dimensions of Brush-Tailed Penguins <b style="font-size:32pt;font-family:Tangerine;">Pygoscelis</b>') +
  theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)) 

```

## Wait, but Why?

- Using appropriate fonts in our plots can make them more readable and visually appealing.
- They may be mandated for publication in some journals, and for publications by organizations.
- Using `sysfonts` , `gfonts`, and `showtext` allows us to use a wide variety of fonts from Google, making it easy to customize our plots.
- Using `ggtext` allows us to add formatted text elements to our plots, making them more informative and visually appealing.
- Note that downloading fonts from Google may not be GDPR compliant, so it is better to use locally installed fonts if possible.
- `gfonts` allows us to programmatically download Google fonts *a priori*, and use them locally, making our websites GDPR compliant.

## Conclusions

In this module, we have learned how to use custom fonts in our ggplot2 plots using the `showtext` package. We have also seen how to use the `ggtext` package to add markdown-formatted text elements to our plots.

## References

1. Thomas Lin Perdersen. <https://www.tidyverse.org/blog/2025/05/fonts-in-r/>
1. Winston Chang. *R Cookbook* (2nd Edition). O'Reilly Media, 2019. <http://www.cookbook-r.com/Graphs/Fonts/>

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
library(grateful)
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("sysfonts", "showtext", "scico")
) %>%
  knitr::kable(format = "simple")

```

:::
