---
title: "Fonts in ggplot"
subtitle: "Using fonts from your machine and from Google"
date: 2025/8/02
date-modified: "`r Sys.Date()`"
order: 20
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)   ## data science package collection (incl. the ggplot2 package)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(showtext)   ## add google fonts to plots
library(scico) ## Colour palette
library(ggtext)      ## add improved text rendering to ggplot2
library(gfonts) ## Use Google Fonts offline
```

## {{< iconify noto-v1 package >}} Setting up R Packages

We may have been using `ggplot2` for a while now, but there is always more to learn about the aesthetics of our plots. In this module, we will explore how to use different fonts and themes to make our plots look more polished and professional. 

We may also want to use fonts on our websites and we will look at using packages like `sysfonts` and `showtext` to add custom fonts to our plots and websites. For instance you may want your website to look this: [Project Peasants](https://project-peasants.netlify.app) !!

## Using Preferred Fonts

We may prefer specific fonts for **our plots**, and we may also want to use these fonts on our websites. There are (atleast) two ways in which one can use fonts in R: 

- **Live Downloading of Fonts**: Downloading fonts from Google within the R session*, using the `showtext` package. 
- **Using Locally Installed Fonts**: Using fonts that are already installed/downloaded on your computer, also using the `sysfonts` package. 

::: callout-note
There is also a more recent, and comprehensive package called `systemfonts` We will postpone this for a later presentation (mainly because I am still figuring it out.)
:::


::: callout-important
### GDPR Compliance
Using fonts that are already installed on your computer is **GDPR compliant**. Using fonts from Google may not be, as it may involve sending data to Google servers everytime your website is loaded.
:::


### Adding Fonts from Google

We will want to add a few new fonts from Google to our graphs. The best way (currently) is to use the `sysfonts` package (which we loaded above) to bring into our work fonts from Google. To view and select the fonts you might want to work with, spend some time looking over [Google Fonts](https://fonts.google.com/). 

Let us add a handful of fonts to our R session. We will use the `font_add_google()` function from the `sysfonts` package to do this. These fonts will then be included into our session using `showtext_auto()`.

```{r}
#| label: add-fonts
#| echo: fenced
sysfonts::font_add_google(name = "Gochi Hand", family = "gochi")
font_add_google("Schoolbell", "bell")
font_add_google("Roboto Condensed", "roboto-condensed")
font_add_google("Noto Sans", "noto")
font_add_google("Jura", "jura")
font_add_google("Ibarra Real Nova", "ibarra")
font_add_google("Open Sans", "open")
font_add_google("Anton", "anton")
font_add_google("Tangerine", "tangerine")
font_add_google("Fraunces", "fraunces")

## Adding local, downloaded fonts
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

showtext_auto() # set these fonts as default for the session
sysfonts::font_families() # check which fonts are available in the session

```


After running the chunk above in your session, you can check which fonts are available in your session by typing `sysfonts::font_families()` in your console. We see that apart from the R-default `sans`, `serif`, and `mono` fonts, we have added the several fonts to our session. 

## Plot Theme using Desired fonts

Now that we have the fonts downloaded, we can use them in our plots. We will use the `theme_set()` function to set a default theme for our plots, and then we will modify the theme elements to use our desired fonts.
Let us set some `ggplot2` theme aspects now!! Here is a handy picture showing ( most of ) the theme-able aspects of a ggplot plot. 

For more info, type `?theme` in your console.

```{r,echo=FALSE, fig.alt = "Rosana Ferrero (@RosanaFerrero) on Twitter Sept 11, 2022", fig.cap= "Rosana Ferrero (@RosanaFerrero) on Twitter Sept 11, 2022"}
knitr::include_graphics("../../../../../materials/images/ggplot_theme_elements.jfif")

```


```{r}
#| label: set-ggplot2-theme
#| 
theme_set(theme_bw(base_size = 12, 
                        base_family = "roboto-condensed"))

## modify plot elements globally (for all following plots)
theme_update(
  axis.ticks = element_line(color = "grey92"),
  axis.ticks.length = unit(.5, "lines"),
  panel.grid.minor = element_blank(),
  legend.title = element_text(family = "bell", size = 12),
  legend.text = element_text(color = "grey30"),
  plot.title = element_text(size = 18, face = "bold"),
  plot.subtitle = element_text(family = "ibarra", size = 12, color = "grey30"),
  plot.caption = element_text(size = 9, margin = margin(t = 15))
)

```

::: callout-note
### Fonts in ggplot

Using `showtext(auto)` as shown above allows us to modify the ***non-graph*** elements of the ggplot chart: the axes, the title, the legend etc. If we wish to add text with fonts to the body of the chart itself, as with geoms and annotations, we need to do one more thing: use `update_geom_defaults()` as shown below:

:::

```{r}
#| label: set-geom-default-font
#| 
update_geom_defaults(geom = "text", new = list(family = "bell"))

```


## Data

We will work with a **familiar** dataset, so that we can concentrate on the chart aesthetics, without having to spend time getting used to the data: the `penguins` dataset again, ~~from the `palmerpenguins` package~~ now part of the R installation!. 

As always, we will start with a **table** of the data:

```{r}
#| label: read-penguins-data
#| message: false
data(penguins, package = "datasets")
penguins <- penguins %>% drop_na() # remove data containing missing data

## Create a nicely formatted table
## uses `kableExtra` package
## 
penguins %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
  kableExtra::scroll_box(width = "700px", height = "500px")

```


## Basic Plot

A basic scatter plot first, which we will decorate with fonts:

```{r}
ggplot(penguins, aes(x = bill_len, y = bill_dep)) +
  geom_point(aes(color = body_mass), alpha = .6, size = 3.5) + 
  
  ## custom colors from the scico package
  scico::scale_colour_scico(palette = "glasgow", direction = -1)+
  
  ## custom labels
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (Pygoscelis)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) PLoS ONE',
    x = 'Bill Length (mm)', 
    y = 'Bill Depth (mm)',
    color = 'Body mass (g)'
  )


```

We see that the ggplot theme uses the fonts that we have assigned to the different elements in the graph. 


## Using `{ggtext}`

From [Claus Wilke's website](https://wilkelab.org/ggtext/) → www.wilkelab.org/ggtext

> The `ggtext` package provides simple Markdown and HTML rendering for ggplot2. Under the hood, the package uses the `gridtext` package for the actual rendering, and consequently it is limited to the feature set provided by `gridtext`.  
Support is provided for Markdown both in theme elements (plot titles, subtitles, captions, axis labels, legends, etc.) and in geoms (similar to `geom_text()`). In both cases, there are two alternatives, one for creating simple text labels and one for creating text boxes with word wrapping.

NOTE: on some machines, the ggtext package may not work as expected. In this case, please do as follows, using your Console:

- remove gridtext: `remove.packages(gridtext)`.   
- Install development version of `gridtext`: `remotes::install_github("wilkelab/gridtext")`

### `element_markdown()`

We can use our familiar markdown syntax right inside the titles and captions of the plot. `element_markdown()` is a theming command made available by the `ggtext` package. 

`element_markdown()` → formatted text elements, e.g. titles, caption, axis text, striptext

```{r}
## assign plot to `gt` - we can add new things to this plot later
## (wrapped in parenthesis so it is assigned and plotted in one step)

(gt <- ggplot(penguins, aes(x = bill_len, y = bill_dep)) +
  geom_point(aes(color = body_mass), alpha = .6, size = 3.5) + 
  scale_x_continuous(breaks = 3:6 * 10, limits = c(30, 60)) +
  scale_y_continuous(breaks = seq(12.5, 22.5, by = 2.5), limits = c(12.5, 22.5)) +
  scico::scale_color_scico(palette = "bamako", direction = -1) +
   
  ## New code starts here: Two Step Procedure with ggtext
  ## 1. Markdown formatting of labels and title, using asterisks
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (*Pygoscelis*)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) *PLoS ONE*',
    x = '**Bill Length** (mm)', 
    y = '**Bill Depth** (mm)',
    color = 'Body mass (g)'
  ) +
   
  ## 2. Add theme related commands from ggtext
  ## render respective text elements
  theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown()
  )
)


```

### `element_markdown()` in combination with HTML

This allows us to change fonts and font-effects in titles, labels, and captions:


```{r}
## use HTML syntax to change text color
gt_mar <- gt +
  labs(title = 'Bill Dimensions of Brush-Tailed Penguins <i style="color:#28A87D;">Pygoscelis</i>') +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10))


## use HTML syntax to change font and text size
gt_mar +
  labs(title = 'Bill Dimensions of Brush-Tailed Penguins <b style="font-size:32pt;font-family:tangerine;">Pygoscelis</b>')


```

## Wait, but Why?

- Using appropriate fonts in our plots can make them more readable and visually appealing.
- They may be mandated for publication in some journals, and for publications by organizations.
- Using `showtext` and `sysfonts` allows us to use a wide variety of fonts from Google, making it easy to customize our plots.
- Using `ggtext` allows us to add formatted text elements to our plots, making them more informative and visually appealing.
- Note that downloading fonts from Google may not be GDPR compliant, so it is better to use locally installed fonts if possible.

## Conclusions

In this module, we have learned how to use custom fonts in our ggplot2 plots using the `showtext` package. We have also seen how to use the `ggtext` package to add markdown-formatted text elements to our plots.

## References

1. Thomas Lin Perdersen. <https://www.tidyverse.org/blog/2025/05/fonts-in-r/>

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
library(grateful)
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("sysfonts", "showtext", "scico")
) %>%
  knitr::kable(format = "simple")

```

:::
