---
title: <iconify-icon icon="openmoji:california-flag"></iconify-icon> <iconify-icon icon="material-symbols:transit-ticket-outline-sharp"></iconify-icon> California Transit Payments 
order: 200
---

## Setting up R Packages

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
library(correlation)
#
library(ggstats)
library(labelled)

```


```{r}
#| label: pedagogical-packages
#| echo: false
library(downloadthis)
library(glue)
library(sjPlot)
library(sjlabelled)
##
###     Adding download buttons
#     data that has been read in
#     {{< downloadthis ../../../../materials/Data/housing_train.csv dname="house_prices" label="Download the House Prices Dataset" icon="database-fill-down" type="info" >}}
#    existing file
#    {{< downloadthis Orange/grouped-summaries.ows dname="grouped_summaries" label="Download the Orange Workflow" icon="database-fill-down" type="info" >}} 
#    
```


#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
    
    theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```




## Introduction

This dataset is the result of a research study on payment options for people using public transit in California.  
The dataset is available on Dataset Dryad:

Pike, Susan (2022). *Transit payment preferences of unbanked passengers.* Dataset Dryad. <https://doi.org/10.25338/B8R04T>

And a brief 2-pager on the research methodology is [here](https://rosap.ntl.bts.gov/view/dot/61204/dot_61204_DS1.pdf).

Yes, peasants, you should read such stuff from other very different domains!

## Read the Data

```{r}
#| label: read-data
#| echo: false
#| message: false

CalPayment <- read_csv("../../../../../materials/Data/doi_10_25338_B8R04T__v20220228/Transit_payment_preferences_of_unbanked_passengers.csv")

CalPayment %>%
  downloadthis::download_this(
    output_name = "Payments-dataset",
    output_extension = ".csv",
    button_label = "Download the Cal Payment data",
    button_type = "warning",
    has_icon = TRUE,
    icon = "fa fa-save"
  )

```

## Data Dictionary
::: callout-note
### Quantitative Variables
Write in.

:::

::: callout-note
### Qualitative Variables
Write in.

:::

::: callout-note
### Observations
Write in.

:::


## Data Munging

```{r}
#| echo: false
#| tbl-cap: Munged Data
CalPayment2 <- CalPayment |> 
  dplyr::select(starts_with("phone")) %>% 
  mutate(across(everything(), ~ str_replace_all(., "1", "2"))) %>% 
  mutate(across(everything(), ~ str_replace_all(., "0", "1"))) %>% 
  mutate(across(everything(), as.numeric))

# ▢ Not having wifi access (1)
# ▢ Ways to add money (2)
# ▢ Identity concerns (3)
# ▢ Monthly fees (4)
# ▢ Knowing the balance (5)
# ▢ Other (please write in) (6)
 
# 
CalPayment2

```


## Summarize and Prepare the Data

```{r}
#| echo: false
CalPayment2 |> count(phone.balance)
CalPayment2 |> count(phone.wifi)
CalPayment2 |> count(phone.money)
CalPayment2 |> count(phone.identity)
CalPayment2 |> count(phone.fees)

variable_labels2 <- c("Wi_Fi access?",
                     "Ways to add money?",
                     "Identity Concerns?",
                     "Monthly Fees?",
                     "Knowing the balance?")
```

Let's label the data variables...

```{r}
#| echo: false
#| tbl-cap: Labelled Data
CalPayment2 <- CalPayment2 %>% 
  sjlabelled::set_label(x = ., label = variable_labels2) %>%
  sjlabelled::set_labels(x = ., labels = c("No", "Yes"))

str(CalPayment2)

```

## Plot the Data


```{r}
#| label: california-payment-likert
#| echo: false

plot_likert(CalPayment2, reverse.colors = F, 
            reverse.scale = F) +
  scale_fill_brewer(palette = "Set2", labels = c("Yes", "No")) +
labs(title = "Challenges in Phone Payment\n in California Transit\n for Un-banked Passengers", caption = "Pike, Susan (2022). Transit payment preferences of unbanked passengers\n [Dataset]Dryad. https://doi.org/10.25338/B8R04T") +
  theme_custom()

```

## Task and Discussion
Complete the Data Dictionary.
Select and Transform the variables as shown.
Create the graph shown below and discuss the following questions:


-   Identify the type of charts
-   Identify the variables used for various geometrical aspects (x, y,
    fill...). Name the variables appropriately.
-   What activity might have been carried out to obtain the data graphed
    here? Provide some details.
-   What would be your recommendation to the Transport Company?
-   To the Phone Companies?
    

