---
title: "Correlation and Regression Explorations"
author: "Arvind V"
format: html
editor: visual
---

## Packages

```{r}
#| label: setup
library(tidyverse)
library(mosaic)
library(broom)
library(ggformula)
ggplot2::theme_set(new = theme_classic())

```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme
#| echo: true
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false

library(systemfonts)
library(showtext)
library(tidyverse)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 
    font <- "Alegreya"   #assign font family up front
    '%+replace%' <- ggplot2::'%+replace%' # nolint
        
    ggplot2::theme_classic(base_size = 14, base_family = font) %+replace%    #replace elements we want to change
    
    theme(
      text = element_text(family = font),  #set base font family
      
      #text elements
      plot.title = element_text(                 #title
                   family = font,          #set font family
                   size = 24,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                    #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      plot.title.position = "plot", 
      
      plot.subtitle = element_text(              #subtitle
                   family = font,          #font family
                   size = 14,                   #font size
                   hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
      plot.caption = element_text(               #caption
                   family = font,          #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
      
      plot.caption.position = "plot",            #right align
      
      axis.title = element_text(                 #axis titles
                   family = "Roboto Condensed",  #font family
                   size = 12),                   #font size
      
      axis.text = element_text(                  #axis text
                   family = "Roboto Condensed",  #font family
                   size = 9),                    #font size
      
      axis.text.x = element_text(                #margin for axis text
                    margin = margin(5, b = 10))
      
      #since the legend often requires manual tweaking 
      #based on plot content, don't define it here
    )
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

```




## Intro

I will work through and "unify" at least two things:

-   Hadley Wickham's chapter on modelling and his analysis of the linear
    model for the diamonds dataset
-   The diagnostic aspects of Linear Regression as detailed in Crawley's
    book

## Explorations into Diagnostic Plots

Let us create dependent y\* variables with different sorts of errors:

```{r}
x <- 0:300
en <- rnorm(301, mean = 0, sd = 5)
eu <- (runif(n = 301) -0.5) * 20
eb <- rnbinom(n = 301,prob = 0.3,size = 2)
eg <- rgamma(n = 301,shape = 1, rate = 1/x)
yn <- x + 10 + en
yu <- x + 10 + eu
yb <- x + 10 + eb
yg <- x + 10 + eg
data <- tibble(x, yn, yu, yb, yg)
data
```

### Normal Errors

```{r}
#| warning: false
lm_norm_aug <- lm(yn ~ x, data = data) %>% 
  augment()
lm_norm_aug %>% gf_point(.resid ~ .fitted)
lm_norm_aug %>% gf_qq(~ .resid) %>% gf_qqline()
```

### Uniform Errors

```{r}
#| warning: false
lm_unif_aug <- lm(yu ~ x, data = data) %>% 
  augment()
lm_unif_aug %>% gf_point(.resid ~ .fitted)
lm_unif_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()
```

### Negative Binom Errors

```{r}
#| warning: false
lm_nbinom_aug <- lm(yb ~ x, data = data) %>% 
  augment()
lm_nbinom_aug %>% gf_point(.resid ~ .fitted)
lm_nbinom_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()

```

### Gamma Errors

```{r}
#| warning: false
lm_gamm_aug <- lm(yg ~ x, data = data) %>% 
  augment()
lm_gamm_aug %>% gf_point(.resid ~ .fitted)
lm_gamm_aug %>% gf_qq(~ .resid, distribution = stats::qnorm) %>% gf_qqline()

```
