---
title: "Inference for Two Independent Means"
subtitle: "How diff-different(sic) are you?"
abstract: "Two independent means from two separate populations."
author: Arvind V. 
date: 22/Nov/2022
date-modified: "`r Sys.Date()`"
order: 110
tags:
- Permutation
- Monte Carlo Simulation
- Random Number Generation
- Distributions
- Generating Parallel Worlds
editor: 
  markdown: 
    wrap: 72
---

:::: pa4
::: {.athelas .ml0 .mt0 .pl4 .black-90 .bl .bw2 .b--blue}
[To be nobody but myself -- in a world which is doing its best, night
and day, to make you everybody else -- means to fight the hardest battle
which any human being can fight, and never stop fighting.]{.f5 .f4-m
.f3-l .lh-copy .measure .mt0}

[--- E.E. Cummings, poet (14 Oct 1894-1962)]{.f6 .ttu .tracked
.fs-normal}
:::
::::

## {{< iconify noto-v1 package >}} Setting up R Packages

```{r}
#| label: setup
#| include: true
#| message: false
#| warning: false

library(tidyverse)
library(mosaic) # Our go-to package
library(ggformula)
library(infer) # An alternative package for inference using tidy data
library(broom) # Clean test results in tibble form
library(skimr) # data inspection
library(tinytable) # Pretty Tables
library(kableExtra) # Pretty Tables
library(ggprism) # Prism-like ggplot themes
##
library(visStatistics) # All-in-one Stats test package
##
library(resampledata3) # Datasets from Chihara and Hesterberg's book
library(openintro) # datasets


```


```{r}
#| label: Extra-Pedagogical-Packages
#| echo: false
#| message: false

library(checkdown)
library(epoxy)
library(TeachHist)
library(TeachingDemos)
library(grateful)
library(downloadthis)
knitr::opts_chunk$set(tidy = 'styler')
library(statlingua) # For Statistics explained in simple English using AI
library(ellmer)
library(latex2exp)

```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme-ggplot4
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false
#| 
library(systemfonts)
library(showtext)
library(ggrepel)
library(marquee)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 

    theme_bw(base_size = 10) + 
    
    theme_sub_axis(title = element_text(family = "Roboto Condensed", 
                                       size = 8),
                   text = element_text(family = "Roboto Condensed", 
                                       size = 6)) + 
    
    theme_sub_legend(text = element_text(family = "Roboto Condensed", 
                                         size = 6),
                     title = element_text(family = "Alegreya", 
                                          size = 8)) + 
    
    theme_sub_plot(title = element_text(family = "Alegreya", 
                                        size = 14, face = "bold"),
                   title.position = "plot",
                   subtitle = element_text(family = "Alegreya", 
                                           size = 10),
                   caption = element_text(family = "Alegreya", 
                                          size = 6),
                   caption.position = "plot")
    
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

ggplot2::update_geom_defaults(geom = "marquee", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "text_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

## tinytable options
options("tinytable_tt_digits" = 2)
options("tinytable_format_num_fmt" = "significant_cell")
options(tinytable_html_mathjax = TRUE)


## Set defaults for flextable
flextable::set_flextable_defaults(font.family = "Roboto Condensed")

```



## {{< iconify openmoji japanese-symbol-for-beginner >}} Introduction

Sometimes, we need to compare if two distinct populations have the same mean. For instance we may wish to compare salaries between two different professions, or test scores between two different teaching methods. In this case, we take a sample ( not necessarily of equal length ) from each population, and compare the means of the two samples.

Here is a diagram showing the workflow of this test based on the kind of data samples we encounter. 


```{mermaid}
%%| echo: false
%%| 
flowchart TD
    A[Inference for Two Independent Means] -->|Check Assumptions| B[Normality: Shapiro-Wilk Test shapiro.test Variances: Fisher F-test var.test]
    B --> C{OK?}
    C -->|Yes, both Parametric| D[t.test]
    C -->|Yes, but not variance Parametric| W[t.test with Welch Correction]
    C -->|No Non-Parametric| E[wilcox.test]
    E <--> G[Linear Model with Signed-Ranks]
    C -->|No Non-Parametric| P[Bootstrap or Permutation]

```

## {{< iconify pajamas issue-type-test-case >}} Case Study #1: Math Anxiety

Let us look at the `MathAnxiety` dataset from the package
`resampledata`. Here we have "anxiety" scores for boys and girls, for
different components of mathematics.

### {{< iconify carbon chart-3d >}} Inspecting and Charting Data

```{r}
#| label: tbl-Math-Anxiety
#| tbl-cap: Math Anxiety Dataset
#| tbl-subcap:
#| - "Dataset (first 10 rows)"
#| - "Qualitative Variables"
#| - "Quantitative Variables"
#| results: hold
data("MathAnxiety",package = "resampledata")
glimpse(MathAnxiety)
skimr::skim(MathAnxiety) %>% tt() %>% 
  theme_striped(font) %>% style_tt(fontsize = 0.75)

```
### {{< iconify streamline dictionary-language-book-solid >}} Data Dictionary

The `MathAnxiety` dataset contains data on students' anxiety levels, with \~600 data observations. 

::: callout-note
### Quantitative Data
- `Age`: Age of the student (in years)
- `AMAS`: Anxiety score from the Abbreviated Math Anxiety Scale (AMAS)
- `RCMAS`: Anxiety score from the Revised Children's Manifest Anxiety Scale (RCMAS)
- `Arith`: Score on Arithmetic Test
:::

::: callout-note
### Qualitative Data

- `Gender`: Boy / Girl
- `Grade`: Grade level of the student (e.g., 5th, 6th, etc.)

:::


### {{< iconify  material-symbols lab-research >}} Hypothesis and Research Questions
- The *target variable-s* for an experiment that resulted in this data might be either of the two Anxiety-related scores. 

::: callout-note
## Research Question

- Is there a difference between `Boy` and `Girl` "anxiety" levels for `AMAS`
(test) in the **population** from which the MathAnxiety dataset is a
sample?
- Ditto for the `RCMAS` scores
- Apropos: We may also be interested in the **correlation** between the *anxiety-related* variables and the `Arith` score obtained, and if it differs between `Boy`s and `Girl`s. 
:::


### {{< iconify mdi chart-areaspline >}} Graphs

First, histograms, densities and counts of the variable we are
interested in, after converting data into long format:

```{r}
#| label: MathAnxiety-summaries
#| code-fold: true
#| layout-ncol: 2
#| message: false

MathAnxiety %>% count(Gender)

MathAnxiety %>% 
  group_by(Gender) %>% 
  summarise(mean = mean(AMAS), n = n())

```


```{r}
#| label: fig-MathAnxiety-graphs
#| fig-cap: AMAS Anxiety Score graphs
#| fig-subcap:
#| - "AMAS Anxiety Score Densities"
#| - "AMAS Anxiety Score Jitter Plot"
#| code-fold: true
#| layout-ncol: 2
#| message: false
#| warning: false

# Set graph theme
theme_set(new = theme_custom())

# https://www.statology.org/r-geom_path-each-group-consists-of-only-one-observation/
MathAnxiety %>%
  gf_density(
    ~ AMAS,
    fill = ~ Gender, colour = ~ Gender,
    alpha = 0.5,
    title = "AMAS Math Anxiety Score Densities",
    subtitle = "Boys vs Girls"
  )
##
MathAnxiety %>% 
  pivot_longer(cols = -c(Gender, Age, Grade),
                             names_to = "type", 
                             values_to = "value") %>% 
  dplyr::filter(type == "AMAS") %>% 
  gf_jitter(
    value ~ Gender,
    group = ~ type, color = ~ Gender,
    width = 0.08, alpha = 0.3,
    ylab = "AMAS Anxiety Scores",
    title = "AMAS Math Anxiety Score Jitter Plots",
    subtitle = "Illustrating Difference in Means"
  ) %>%
  gf_summary(group = ~ 1, # Damn!!! Look at the reference link above
             fun = "mean", geom = "line", 
             colour = ~"MeanDifferenceLine", 
             lty = 1, linewidth = 2) %>% 
  gf_summary(geom = "point", size = 3, colour = "black") %>% 
  gf_refine(scale_x_discrete(
    breaks = c("Boy", "Girl"),
    labels = c("Boys", "Girls"),
    guide = guide_prism_bracket(width = 0.1,outside = FALSE)
  ))  %>% 
  gf_annotate(x = 0.65, y = 25, geom = "text", label = "Mean\n Boys' Anxiety Scores", family = "Roboto Condensed") %>%
    gf_annotate(x = 2.35, y = 25, geom = "text", label = "Mean\n Girls' Anxiety Scores",family = "Roboto Condensed") %>% 
    gf_annotate(x = 1.5, y = 30, geom = "label", label = "Mean Difference Line\nSlope indicates\n differences in mean", fill = "moccasin", family = "Roboto Condensed") %>%
  gf_theme(theme(legend.position = "none", axis.line.x = element_line(linewidth = 1)))

```

The distributions for anxiety scores for boys and girls overlap
considerably and are very similar, though the jitter plot for boys shows
significant outliers. Are they close to being *normal distributions* too?
We should check.

#### A. {{< iconify mdi chart-bell-curve >}} Check for Normality

Statistical tests for means *usually* require a couple of checks[^1]
[^2]:\

-   Are the data normally distributed?\
-   Are the data variances similar?

Let us complete a check for normality: the `shapiro.wilk` test checks
whether a Quant variable is from a normal distribution; the NULL
hypothesis is that the data *are* from a normal distribution. We will
also look at Q-Q plots for both variables:

```{r}
#| label: fig-MathAnxiety-charts
#| fig-cap: AMAS Anxiety Score Normality Check
#| fig-subcap:
#| - "AMAS Anxiety Score Densities"
#| - "AMAS Anxiety Score Q-Q Plots"
#| layout-ncol: 2
#| warning: false
#| code-fold: true

# Set graph theme
theme_set(new = theme_custom())
#
MathAnxiety %>%
  gf_density( ~ AMAS,
              fill = ~ Gender,
              alpha = 0.5,
              title = "Math Anxiety scores for boys and girls") %>%
  gf_facet_grid(~ Gender) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_theme(theme(legend.position = "none")) # No Need!!
##
MathAnxiety %>% 
  gf_qqline(~ AMAS, color = ~ Gender,
            title = "Math Anxiety Scores..",
            subtitle= "Are they Normally Distributed?") %>%
  gf_qq() %>% 
  gf_facet_wrap(~ Gender) %>%  # independent y-axis

  gf_theme(theme(legend.position = "none")) # No Need!!

```

Let us split the dataset into subsets, to execute the normality check
test (Shapiro-Wilk test):

```{r}
boys_AMAS <- MathAnxiety %>% 
  filter(Gender== "Boy") %>% 
  select(AMAS)
##
girls_AMAS <- MathAnxiety %>% 
  filter(Gender== "Girl") %>% 
  select(AMAS)
```

::::: columns
::: {.column width="60%"}
```{r}
#| label: shapiro-test-1
#| eval: false
shapiro.test(boys_AMAS$AMAS)
shapiro.test(girls_AMAS$AMAS)

```
:::

::: {.column width="40%"}
```{r}
#| ref.label: shapiro-test-1
#| echo: false
```
:::
:::::

The distributions for anxiety scores for boys and girls are almost
normal, visually speaking. With the Shapiro-Wilk test we find that the
scores for `girls` **are** normally distributed, but the `boys` scores
are not so. Tch, tch. These boys. 

::: callout-note
The `p.value` obtained in the `shapiro.wilk` test suggests the chances
of the data being so, **given** the Assumption that they ***are normally
distributed***.
:::

We see that `MathAnxiety` contains discrete-level scores for anxiety for
the two variables (for Boys and Girls) anxiety scores. The `boys` score
has a significant outlier, which we saw earlier and perhaps that makes
that variable lose out, perhaps narrowly.

#### B. {{< iconify mdi sigma-lower >}} Check for Variances

Let us check if the two variables have similar variances: the `var.test`
does this for us, with a NULL hypothesis that the variances are not
significantly different:

```{r}
#| message: false
#| results: hold
#| 
MathAnxiety %>% 
  dplyr::select(AMAS, Gender) %>% 
  dplyr::group_by(Gender) %>% 
  dplyr::summarize(variance = var(AMAS), n = n())

var.test(AMAS ~ Gender, data = MathAnxiety, 
         conf.int = TRUE, conf.level = 0.95) %>% 
  broom::tidy()
##
qf(0.975,275,322)

```

The variances are quite similar as seen by the $p.value = 0.82$. We also
saw it visually when we plotted the overlapped distributions earlier.

::: callout-important
### Conditions:

1.  The two variables are not **both** normally distributed.
2.  The two variances are significantly similar.
:::

### {{< iconify academicons hypothesis >}} Hypothesis

Based on the graphs, how would we formulate our Hypothesis? We wish to
infer whether there is any difference in the mean anxiety score between
Girls and Boys, in the **population** from which the dataset
`MathAnxiety` has been drawn. So accordingly:

$$
H_0: \mu_{Boys} - \mu_{Girls} = 0\\
$$

$$
H_a: \mu_{Boys} - \mu_{Girls} \ne 0\\
$$

### {{< iconify academicons hypothesis >}} Observed and Test Statistic

What would be the **test statistic** we would use? The **difference in
means**. Is the observed difference in the means between the two groups
of scores non-zero? We use the `diffmean` function:

```{r}

obs_diff_amas <- diffmean(AMAS ~ Gender, data = MathAnxiety) 
obs_diff_amas

```

Girls' `AMAS` anxiety scores are, on average, $1.76$ points higher than
those for Boys in the dataset/sample.

::: callout-note
### On Observed Difference Estimates

Different tests here will show the difference as positive or negative,
but with the same value! This depends upon the way the factor variable
`Gender` is used, i.e. Boy-Girl or Girl-Boy...
:::

### {{< iconify fluent-mdl2 insights >}} Inference

::::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
#### Parametric t.test

Since the data are not both normally distributed, though the variances
similar, we typically cannot use a parametric `t.test`. However, we can
still examine the results:

```{r}
t1_amas <- mosaic::t_test(AMAS ~ Gender, data = MathAnxiety) 

t1_amas%>% 
  broom::tidy()
```

The `p.value` is $0.001$ ! And the `Confidence Interval` does not
straddle $0$. So the `t.test` gives us good reason to reject the Null
Hypothesis that the means are similar and that there **is** a
significant difference between Boys and Girls when it comes to `AMAS`
anxiety. But can we really believe this, given the non-normality of
data?

#### Mann-Whitney Test

Since the data variables do not satisfy the assumption of being
*normally distributed*, and though the variances are similar, we use the
classical `wilcox.test` (Type `help(wilcox.test)` in your Console.)
which implements what we need here: the *Mann-Whitney U test*:[^3]

> The Mann-Whitney test as a test of **mean ranks**. It first ranks all
> your values from high to low, computes the mean rank *in each group*,
> and then computes the probability that random shuffling of those
> values between two groups would end up with the mean ranks as far
> apart as, or further apart, than you observed. No assumptions about
> distributions are needed so far. (emphasis mine)

$$
mean(rank(AMAS_{Girls})) - mean(rank(AMAS_{Boys})) = diff
$$

$$
H_0: \mu_{Boys} - \mu_{Girls} = 0
$$

$$
H_a: \mu_{Boys} - \mu_{Girls} \ne 0
$$

```{r}
#| label:  fig-Independent-Data-Plots-1
#| fig-cap: AMAS Scores Ranked Data by Gender
#| code-fold: true
#| message: false
#| warning: false
library(ggprism)
library(ggtext)
library(glue)
library(latex2exp)
# https://www.statology.org/r-geom_path-each-group-consists-of-only-one-observation/


# Set graph theme
theme_set(new = theme_custom())
#

MathAnxiety %>% 
  gf_jitter(rank(AMAS) ~ Gender, color = ~ Gender, 
            show.legend = FALSE,
            width = 0.05, alpha = 0.25,
            ylab = "Ranks of AMAS anxiety scores", 
            title = "Ranked Math Anxiety Scores for Boys and Girls") %>% 
  gf_summary(group = ~ 1, # See the reference link above. Damn!!!
             fun = "mean", geom = "line", colour = "lightblue", 
             lty = 1, linewidth = 2) %>% 
  gf_summary(fun = "mean", colour = "firebrick", size = 4, geom ="point") %>% 
  gf_refine(scale_x_discrete(breaks =  c("Boy", "Girl"), 
                             labels = c("Boy", "Girl"), 
                             guide = "prism_bracket")) %>% 
  # gf_annotate("label", label = TeX(r"(\textbf{Slope} $mu_{Rank Boys}-mu_{Rank Girls} \neq 0$ )"), y = 500, x = 1.5, 
  #          color = "black", fill = "moccasin") %>% 
   gf_annotate("label", label = "Mean Rank \nBoys Scores", 
           y = 300, x = 0.75, inherit = FALSE) %>%
   gf_annotate("label", label = "Mean Rank \nGirls Scores", 
           y = 320, x = 2.25, inherit = FALSE)

```

This is pretty similar to @fig-MathAnxiety-graphs-2 with **actual** scores. Here we have plotted the **ranks** of the scores.


```{r}
t2_amas <- wilcox.test(AMAS ~ Gender, data = MathAnxiety, 
            conf.int = TRUE, 
            conf.level = 0.95)
t2_amas %>% broom::tidy()
```

The `p.value` is very similar, $0.00077$, and again the
`Confidence Interval` does not straddle $0$, and we are hence able to
reject the NULL hypothesis that the means are equal and accept the
alternative hypothesis that there is a significant difference in mean
anxiety scores between Boys and Girls.

#### Linear Model Interpretation

We can apply the *linear-model-as-inference* interpretation to the
**ranked data** data to implement the non-parametric test as a Linear
Model:

$$
lm(rank(AMAS) \sim  gender) = \beta_0 + \beta_1 * gender
$$

$$
H_0: \beta_1 = 0\\
$$

$$
H_a: \beta_1 \ne 0\\
$$

```{r}
#| label:  fig-Independent-Data-Plots-2
#| fig-cap: "AMAS Anxiety by Gender (Again)"
#| code-fold: true
#| message: false
#| warning: false
library(ggprism)
library(ggtext)
library(glue)
library(latex2exp)
# https://www.statology.org/r-geom_path-each-group-consists-of-only-one-observation/
ggplot2::theme_set(new = theme_custom())

MathAnxiety %>% 
  gf_jitter(rank(AMAS) ~ Gender, color = ~ Gender, 
            show.legend = FALSE,
            width = 0.05, alpha = 0.25,
            ylab = "Ranks of AMAS anxiety scores", 
            title = "Ranked Math Anxiety Scores for Boys and Girls") %>% 
  gf_summary(group = ~ 1, # See the reference link above. Damn!!!
             fun = "mean", geom = "line", colour = "lightblue", 
             lty = 1, linewidth = 2) %>% 
  gf_summary(fun = "mean", colour = "firebrick", size = 4, geom ="point") %>% 
  gf_refine(scale_x_discrete(breaks =  c("Boy", "Girl"), 
                             labels = c("Boy", "Girl"), 
                             guide = "prism_bracket")) %>% 
  gf_text(label = TeX(r"(\textbf{Slope is} $beta_1$ \textbf{in linear model})", output = "character"), 
          parse = TRUE, 500 ~ 1.5, 
          color = "black", fill = "moccasin") %>% 
   gf_text(label = "Mean Rank \nBoys Scores", 
           300 ~ 0.75, inherit = FALSE) %>%
   gf_text(label = "Mean Rank \nGirls Scores", 
           320 ~ 2.25,inherit = FALSE)

```

```{r}
#| results: hold

lm(rank(AMAS) ~ Gender, 
   data = MathAnxiety) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95)

```

::: callout-tip
#### Dummy Variables in `lm`

Note how the **Qual** variable was used here in Linear Regression! The
`Gender` variable was treated as a binary "dummy" variable[^4].
:::

Here too we see that the `p.value` for the *slope term* ("GenderGirl")
is significant at $7.4*10^{-4}$.

#### Permutation Test

We **pretend** that `Gender` has no *effect* on the `AMAS` anxiety
scores. If this is our position, then the `Gender` labels are
essentially meaningless, and we can pretend that any `AMAS` score
belongs to a Boy or a Girl. This means we can `mosaic::shuffle`
(permute) the `Gender` labels and see how **uncommon** our *real* data
is. And we do not have to really worry about whether the data are
normally distributed, or if their variances are nearly equal.

::: callout-important
The "pretend" position is exactly the NULL Hypothesis!! The "uncommon"
part is the `p.value` under NULL!!
:::

```{r}

null_dist_amas <- 
  do(4999) * diffmean(data = MathAnxiety, AMAS ~ shuffle(Gender))
null_dist_amas
```

```{r}
#| label: fig-NullDist-amas
#| fig-cap: Null Distribution of Difference in AMAS Means
#| fig-subcap:
#| - "Histogram of Null Distribution"
#| - "Cumulative Density of Null Distribution"
#| layout-ncol: 2
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

gf_histogram(data = null_dist_amas, ~ diffmean, bins = 25) %>%
  gf_vline(xintercept = obs_diff_amas, 
           colour = "red", linewidth = 1,
           title = "Null Distribution by Permutation", 
           subtitle = "Histogram") %>% 
  gf_labs(x = "Difference in Means") %>% 
  gf_refine(scale_y_continuous(expand = c(0, 0)))

###
gf_ecdf(data = null_dist_amas, ~ diffmean, 
        linewidth = 1) %>%
  gf_vline(xintercept = obs_diff_amas, 
           colour = "red", linewidth = 1,
           title = "Null Distribution by Permutation", 
           subtitle = "Cumulative Density") %>% 
  gf_labs(x = "Difference in Means") %>% 
  gf_refine(scale_y_continuous(expand = c(0, 0)))

```

```{r}
1-prop1(~ diffmean <= obs_diff_amas, data = null_dist_amas)

```

Clearly the `observed_diff_amas` is much beyond anything we can generate
with permutations with `gender`! And hence there is a significant
difference in weights across `gender`!

#### Visual and by Hand t.test

Here too, as with the [Test for a Single Mean](../100-OneMean/index.qmd), the inner mechanism of the t-test is very similar, but for some important differences: 

1. ***Two Means***: Calculate the `mean` of the two samples $\bar{x}$ and $\bar{y}$.
2. Take the difference between the two sample means $\bar{x} - \bar{y}$:
   $$
   \text{Difference} = \bar{x} - \bar{y}
   $$
2. ***Net Variance***: Here we need to **pool in** the variances of both samples and compute the **net variance**
3. Calculate the `standard error` (using net-variance) as follows:
   $$
   SE = \sqrt{\frac{s_x^2}{n_x} + \frac{s_y^2}{n_y}}
   $$
   where $s_x$ and $s_y$ are the standard deviations of the two samples, and $n_x$ and $n_y$ are their respective sample sizes.
3. Scale the difference by the `standard error` to get a `test statistic` $t$.
   $$
   t = \frac{\text{Difference}}{SE}
   $$
4. If the `test statistic` is more than $\pm 1.96$, we can say that it is beyond the the `confidence interval` of the sample mean difference $\bar{x} - \bar{y}$
   $$
   \text{Confidence Interval} = \bar{x} - \bar{y} \pm 1.96 \times SE
   $$
5. Therefore if the actual difference and bis far beyond the confidence interval, hmm...we cannot think our belief is correct and we change our opinion that the two means $\bar{x}$ and $\bar{y}$ are the same.

Let us translate that mouthful into calculations!

:::: {layout="[15,15,15,40,15]"}

::: {#first-column}
```{r}
(mean_diff_belief <- 0.0) # Assert our belief
```
Null Hypothesis / Belief
:::

::: {#second-column}
```{r}
(obs_diff_amas <- diffmean(AMAS ~ Gender, data = MathAnxiety) %>% as.numeric())
```
Observed Difference in Means
:::

::: {#third-column}
```{r}
#| code-fold: true
#| 
(MathAnxiety %>% 
  dplyr::select(AMAS, Gender) %>% 
  dplyr::group_by(Gender) %>% 
  dplyr::summarize(variance = var(AMAS), n = n()) %>% 
  dplyr::mutate(se_contribution = variance/n) %>% 
  dplyr::summarise(std_error = sqrt(sum(se_contribution))) %>% 
  as.numeric () -> std_error_diff)

```
Standard Error
:::

::: {#fourth-column}
```{r}
(conf_int2 <- tibble(ci_low = obs_diff_amas - 1.96*std_error_diff, ci_high = obs_diff_amas + 1.96*std_error_diff))
```
Confidence Intervals
:::
::: {#fifth-column}
```{r}
(t <- (obs_diff_amas - mean_diff_belief)/std_error_diff)
```
Test Statistic
:::
::::


How can we visualize this? We are going to have to construct a **difference dataset** containing ***all possible differences*** between the two groups, and then plot the distribution of those differences. We can then visualize the observed difference in means, the confidence intervals, and the test statistic. Let'see if that works!

```{r}
#| label: fig-visualizing-t-test-two-means
#| fig-cap: "Visualizing the t-test for two means"
#| message: false
#| warning: false
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

boys_scores <- boys_AMAS %>% rename("boys" = AMAS)
girls_scores <- boys_AMAS %>% rename("girls" = AMAS)
all_diff <- expand_grid(boys_scores, girls_scores) %>% mutate(diff = boys - girls)
##
mosaic::xqnorm(p = c(0.025, 0.975), mean = obs_diff_amas, sd = std_error_diff,
               return = c("value"), verbose = T, plot = F) -> xq1

xqnorm(p = c(0.025, 0.975),  mean = obs_diff_amas, sd = std_error_diff,
       digits = 3, plot = TRUE,
       return = c("plot"), verbose = F, alpha = 0.5, colour = "black",
        system = "gg")  %>% 
  gf_dhistogram(~ diff, data = all_diff,bins = 50, 
                xlab = "Difference in Means",
                ylab = "Density", inherit = F) %>% ## Very important!
  gf_fitdistr(~ diff, data = all_diff, inherit = F) %>% 
  gf_vline(xintercept = mean_diff_belief, colour = "red") %>% 
  gf_vline(xintercept = obs_diff_amas, colour = "purple") %>% 
  
  gf_refine(scale_y_continuous(limits = c(0, 0.9),expand = c(0, 0)), 
            scale_x_continuous(limits = c(-7.5, 7.5),
                breaks = c(-10.0, -5.0, 0, 5, 10, obs_diff_amas),
                labels = scales::number_format(accuracy = 0.001,
                                 decimal.mark = "."))) %>% 
  
    gf_annotate(geom = "label", x = -4.5, y = 0.15, 
            label = "Sample-Difference Histogram and Density", 
            colour = "black") %>% 
  
  ## Observed Difference
  gf_annotate(geom = "label", x = obs_diff_amas, y = 0.8, 
            label = "Observed Mean Difference", colour = "purple") %>% 
## Null Hypothesis

  gf_annotate(geom = "label", x = mean_diff_belief, y = 0.7, 
            label = "Null Hypothesis (Belief)") %>% 

## Confidence Intervals
  gf_annotate("segment", 
            x = conf_int2$ci_low, y = 0.0, 
            xend = conf_int2$ci_low, yend = 0.5,
            color = "purple", linewidth = 0.5) %>% 
  gf_annotate("segment", 
            x = conf_int2$ci_high, y = 0.0, 
            xend = conf_int2$ci_high, yend = 0.5,
            color = "purple", linewidth = 0.5) %>% 
  gf_annotate("segment", 
            x = conf_int2$ci_low, y = 0.35, 
            xend = conf_int2$ci_high, yend = 0.35,
            color = "purple", linewidth = 0.5, 
            arrow = arrow(length = unit(0.25,"cm"), ends = "both")) %>% 
  gf_annotate(geom = "label", x = obs_diff_amas, y = 0.4, 
            label = "95% Confidence Interval", 
            colour = "purple") %>% 
  gf_labs(title = "Visualizing the t-test for Math Anxiety Scores",
       subtitle = "Difference in Means, Confidence Intervals, and Test Statistic") %>% 
  gf_theme(theme_custom())

```
We see that the difference between means is `r t` times the std_error! At a distance of $1.96$ (either way) the probability of this data happening by chance already drops to $2.5 \%$ !! At this distance of `r t`, we would have negligible probability of this data occurring by chance!

:::


### All Tests Together

We can put all the test results together to get a few more insights
about the tests:

```{r}
#| label: tbl-All-Tests-tt-1
#| tbl-cap: All Tests for AMAS Anxiety Scores
#| tbl-subcap:
#| - "t.test"
#| - "wilcox.test"
#| - "Linear Model with Original Data"
#| - "Linear Model with Ranked Data"
#| code-fold: true
#| results: hold

mosaic::t_test(AMAS ~ Gender, 
               data = MathAnxiety) %>% 
  broom::tidy() %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = F, html_font = "Roboto Condensed") %>% 
  kableExtra::column_spec(5, width = "20em", bold = T, background = "gold")


wilcox.test(AMAS ~ Gender, 
               data = MathAnxiety) %>% 
  broom::tidy() %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = F, html_font = "Roboto Condensed") %>% 
  kableExtra::column_spec(2, width = "10em", bold = T, background = "gold")

lm(AMAS ~ Gender, 
   data = MathAnxiety) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95,) %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = F, html_font = "Roboto Condensed") %>% 
  kableExtra::column_spec(1, bold = T, border_right = T) %>%
  kableExtra::column_spec(5, width = "10em", bold = T, background = "gold")

lm(rank(AMAS) ~ Gender, 
   data = MathAnxiety) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95) %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = F, html_font = "Roboto Condensed") %>% 
  kableExtra::column_spec(1, bold = T, border_right = T) %>%
  kableExtra::column_spec(5, width = "10em", bold = T, background = "gold")


```


As we can see, all tests are in agreement that there is a significant
*effect* of `Gender` on the `AMAS` anxiety scores!

### {{< iconify fa-solid ring >}} One Test to Rule Them All: visStatistics

We can use the `visStatistics` package to run all the tests in one go,
using the in-built **decision tree**. This is a very useful package for
teaching statistics, and it can be used to run all the tests we have
seen so far, and more. Here goes: we use the `visstat` function to run
all the tests, and then we can summarize the results. The `visstat`
function takes a dataset, a quantitative variable, a qualitative
variable, and some options for the tests to run.

From the `visStatistics` package documentation:

> visStatistics automatically selects and visualises appropriate
> statistical hypothesis tests between two column vectors of type of
> class "numeric", "integer", or "factor". The choice of test depends on
> the class, distribution, and sample size of the vectors, as well as
> the user-defined ‘conf.level’. The main function `visstat()`
> visualises the selected test with appropriate graphs (box plots, bar
> charts, regression lines with confidence bands, mosaic plots, residual
> plots, Q-Q plots), annotated with the main test results, including any
> assumption checks and post-hoc analyses.

```{r}
#| label: fig-visStatistics-1
#| fig-cap: "visStatistics Output for AMAS Anxiety Scores"


# Generate the annotated plots and statistics
visstat(x = MathAnxiety$Gender, 
        y = MathAnxiety$AMAS,
        conf.level = 0.95, numbers = TRUE) %>% 
  summary()


```

The tool runs the Welch t-test and declares the p-value to be significant. The Shapiro-Wilk test results here also confirm what we had performed earlier. Hence we can say that we may reject the NULL Hypothesis and state that there is a statistically significant difference in `AMAS` anxiety scores between Boys and Girls. 

### An AI Generated Explanation of the Statistical Model

We have used the `statlingua` package to generate an AI explanation of the statistical model `t1_amas` we have used in this module. `statlingua` interfaces to the `ellmer` package, which in turn interfaces to the `ollama` API for AI explanations. The AI model used is `llama3.1`, and the explanation is tailored for a novice audience with moderate verbosity.

::: callout-note
### AI Generated Explanation
The command was run in the Console and the results pasted into this Quarto document. I have not edited this at all. Do look for complex verbiage and outright hallucinatory responses. 

:::

```{r}
#| label: ai-explanations
#| code-fold: true
#| eval: false
library(ellmer)
library(statlingua)
library(chores)
client <- ellmer::chat_ollama(model = "llama3.1", echo = FALSE)
###

exam_context <- "
The model uses a data set on child Math-related Anxiety scores from the resampledata package in R. The dataset has Age, Gender, Grade, and Arithmeitc test score variables, along with AMS and RCMAS math-anxiety scores. 
The hypothesis is that the average Math Anxiety score is no different for Boys and Girls. The model conducted is a two-sample t-test to assess whether the mean of the Math Anxiety scores significantly differ between Boys and Girls."

explanation_amas <- statlingua::explain(t1_amas, client = client, 
                context = exam_context,
                audience = "novice", verbosity = "moderate") # moderate / detailed

```

#### Welch Two Sample t-test Output Explanation

The output given represents the results from conducting a **Welch Two Sample t-test** on the `AMAS` (Math-Related Anxiety Scale) variable in the dataset, specifically examining differences in mean anxiety scores between boys (`Boy`) and girls (`Girl`).

##### Assumptions and Suitability
Given context, the hypothesis to be tested is that average Math Anxiety scores are no different for Boys and Girls. A Welch Two Sample t-test is chosen, suitable given its independence of observations assumption aligned with the stated dataset properties.

It assumes data follows a normal distribution in each group, though the test itself does not check for normality. However, due to non-homogeneous variance typical in many datasets, a t-test variant specifically designed for such cases, like Welch's, is preferable over the (more restrictive) homoscedastic t-test.

##### Output Components

- **data**: ` AMAS by Gender` - This indicates that the analysis comparing means of Math anxiety scores (`AMAS`) between genders (`Boy` and `Girl`) generated this output.
- **Test Statistic**: The calculated test statistic is `t = -3.2918`. A negative t-value suggests an average math-anxiety-score difference in a direction favoring girls (higher average score for them), compared to boys.
- **Degrees of Freedom (`df`)**: The degrees of freedom associated with this test are `df = 580.2`. This value is not an integer, which is typical for Welch’s t-test due to its handling of unequal variances among the groups.
- **p-value**: The p-value from the output, `0.001056`, is considered highly significant at any conventional significance level (e.g., $\alpha = 0.05$). This implies the average Math Anxiety score significantly differs between boys and girls according to the test.
- **Alternative Hypothesis**: This output supports an alternative hypothesis stating that true differences in means (`true difference in means between group Boy and group Girl is not equal to 0`). From the research question's perspective, this confirms that there are significant differing mean Math Anxiety scores for boys and girls.
- **95% Confidence Interval (CI)**: Provided a confidence interval which suggests with 95% confidence, the population average math anxiety score difference can fall between ` -2.8222293` and `-0.7129706`. [Since zero is within this range, one might initially question 
the interpretation of significant differences based on p-values alone]{.black .bg-red}. [***Hallucination!!!***]{.black .bg-gold}
- **Sample Estimates**: Offers summary statistics (means) for individual groups (`mean in group Boy = 21.16718`, `mean in group Girl = 22.93478`), highlighting the observed average Math Anxiety scores per gender.

##### Checking Assumptions and Practical Suggestions

Given this analysis, practical steps to check assumptions include:
- **Normality of Data**: Assess using graphical methods (histograms, Q-Q plots for each group). Check with **Shapiro-Wilk test**.
  
  If data shows clear indication against normality or from formal tests (e.g., Shapiro-Wilk W-statistic indicates distribution significantly deviates from being normally distributed), consider a non-parametric alternative.

- **Equal Variance Assumption**: Since the Welch variant was used here, assessing equal variances isn't directly necessary with its output. But to check formally, one might use **Levene’s Test** or **Barlett's test**, and if significant, reconsider using Welch's for a better estimate of effects in such scenarios.

##### Conclusion
At $\alpha = 0.001056$, the observed differences (higher mean Math Anxiety among girls than boys) are considered statistically significant. Therefore, based on this analysis and context provided, we have sufficient evidence to reject the null hypothesis that there is no difference in average Math-related Anxiety Scores between boys and girls. 



## {{< iconify pajamas issue-type-test-case >}} Case Study #2: Youth Risk Behavior Surveillance System (YRBSS) survey

Every two years, the [Centers for Disease Control and Prevention](https://www.cdc.gov/) in the
USA conduct the **Youth Risk Behavior Surveillance System (YRBSS)**
survey, where it takes data from high-schoolers (9th through 12th grade),
to analyze health patterns. We will work with a selected group of
variables from a random sample of observations during one of the years
the YRBSS was conducted. The `yrbss` dataset is part of the `openintro`
package. Type this in your console: `help(yrbss)`.

### {{< iconify carbon chart-3d >}} Inspecting and Charting Data

```{r}
#| label: Data-Case-Study-2
#| results: hold
data(yrbss, package = "openintro")
yrbss
yrbss_inspect <- inspect(yrbss)
yrbss_inspect$categorical
yrbss_inspect$quantitative

```

We have 13K data entries, and with 13 different variables, some Qual and
some Quant. *Many entries are missing too*, typical of real-world data
and something we will have to account for in our computations. The
meaning of each variable can be found by bringing up the help file. Type
`help(yrbss)` in your Console.

First, histograms, densities and counts of the variable we are
interested in:

```{r}
#| label: EDA-RQ1-1

yrbss_select_gender <- yrbss %>% 
  select(weight, gender) %>% 
  mutate(gender = as_factor(gender)) %>% 
  drop_na(weight) # Sadly dropping off NA data

```

```{r}
#| label: fig-yrbss-weight-gender
#| fig-cap: "Highschoolers' Weights Distributions"
#| fig-subcap:
#| - "Overlapped Density Plot"
#| - "Jitter Plots"
#| layout-ncol: 2
#| warning: false
#| message: false
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

yrbss_select_gender %>%
  gf_density( ~ weight,
              fill = ~ gender, colour = ~ gender,
              alpha = 0.3,
              title = "Highschoolers' Weights by Gender") %>% 
  gf_refine(scale_fill_brewer(name = "Gender", palette = "Set1", 
                              aesthetics = c("colour", "fill")))

###
yrbss_select_gender %>% 
  gf_jitter(weight ~ gender, color = ~ gender, 
            show.legend = FALSE,
            width = 0.05, alpha = 0.25,
            ylab = "Weight", 
            title = "Weights of Boys and Girls") %>% 
  gf_summary(group = ~ 1, # See the reference link above. Damn!!!
             fun = "mean", geom = "line", colour = "lightblue", 
             lty = 1, linewidth = 2) %>% 
  gf_summary(fun = "mean", colour = "black", 
             size = 4, geom ="point") %>% 
  gf_refine(scale_x_discrete(
   breaks =  c("male", "female"), 
            labels = c("male", "female"), 
    guide = guide_prism_bracket(width = 0.1, outside = FALSE)
  ))  %>% 
  gf_annotate(x = 0.75, y = 60, geom = "text", 
              label = "Mean\n Girls' Weights") %>% 
  gf_annotate(x = 2.25, y = 60, geom = "text", 
              label = "Mean\n Boys' Weights") %>% 
  gf_annotate(x = 1.5, y = 100, geom = "label", 
              label = "Slope indicates\n differences in mean", 
              fill = "moccasin") %>% 
  gf_refine(scale_colour_brewer(name = "Gender", palette = "Set1")) %>% 
  gf_theme(theme(legend.position = "none", axis.line.x = element_line(linewidth = 1)))

```

```{r}
yrbss_select_gender %>% count(gender)

```

Overlapped Distribution plot shows some difference in the *means*; and
the Jitter plots show visible difference in the *means*. In this Case
Study, our research question is:

### {{< iconify academicons hypothesis >}} Hypothesis

::: callout-note
## Research Question

Does `weight` of highschoolers in this dataset vary with `gender`?
:::

Based on the graphs, how would we formulate our Hypothesis? We wish to
infer whether there is difference in mean `weight` across `gender`. So
accordingly:

$$
H_0: \mu_{weight-male} = \mu_{weight-female}
$$ 

$$
H_a: \mu_{weight-male} \ne \mu_{weight-female}
$$

#### A. {{< iconify mdi chart-bell-curve >}} Check for Normality

As stated before, statistical tests for means *usually* require a couple
of checks:\

-   Are the data normally distributed?\
-   Are the data variances similar?

We will complete a visual check for normality with plots, and since we
cannot do a `shapiro.test` (length(data) \>= 5000) we can use the
Anderson-Darling test.

Let us plot frequency distribution and Q-Q plots[^5] for both variables.

```{r}
#| label: fig-normality-check-2
#| fig-cap: "Highschoolers' weights Normality Check"
#| fig-subcap:
#| - "Facetted Densities"
#| - "Q-Q Plots"
#| results: hold
#| layout-ncol: 2
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

male_student_weights <- yrbss_select_gender %>% 
  filter(gender == "male") %>% 
  select(weight)
##
female_student_weights <- yrbss_select_gender %>% 
  filter(gender == "female") %>% 
  select(weight)

#shapiro.test(male_student_weights$weight)
#shapiro.test(female_student_weights$weight)

yrbss_select_gender %>%
  gf_density( ~ weight,
              fill = ~ gender, colour = ~ gender,
              alpha = 0.3,
              title = "Highschoolers' Weights by Gender") %>%
  gf_facet_grid(~ gender) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_refine(scale_fill_brewer(name = "Gender", palette = "Set1", aesthetics = c("colour", "fill")))
##
yrbss_select_gender %>% 
  gf_qqline(~ weight | gender, ylab = "weight", colour = ~ gender) %>% 
  gf_qq(title = "Q-Q Plots of Male and Female Weights") %>% 
  gf_refine(scale_fill_brewer(name = "Gender", palette = "Set1", aesthetics = c("colour", "fill")))

```

Distributions are not too close to normal...perhaps a hint of a
rightward skew, suggesting that there are some obese students.

No real evidence (visually) of the variables being normally distributed.

```{r}
library(nortest)
nortest::ad.test(male_student_weights$weight)
nortest::ad.test(female_student_weights$weight)

```

`p-values` are very low and there is no reason to think that the data is
normal.

#### B. {{< iconify mdi sigma-lower >}} Check for Variances

Let us check if the two variables have similar variances: the
`var.test`does this for us, with a NULL hypothesis that the variances
are not significantly different:

```{r}
#| message: false
#| results: hold
var.test( weight ~  gender, data = yrbss_select_gender, 
          conf.int = TRUE,
          conf.level = 0.95) %>% 
  broom::tidy()

#qf(0.975,6164, 6413)

```

The `p.value` being so small, we are able to reject the NULL Hypothesis
that the variances of `weight` are nearly equal across the two exercise
regimes.

::: callout-important
### Conditions

1.  The two variables are not normally distributed.
2.  The two variances are also significantly different.
:::

This means that the parametric `t.test` must be eschewed in favour of
the non-parametric `wilcox.test`. We will use that, and also attempt
`linear models` with rank data, and a final permutation test.

### {{< iconify academicons hypothesis >}} Observed and Test Statistic

What would be the **test statistic** we would use? The **difference in
means**. Is the observed difference in the means between the two groups
of scores non-zero? We use the `diffmean` function, from `mosaic`:

```{r}
#| layout: [[40,10,10]]

obs_diff_gender <- diffmean(weight ~ gender, 
                            data = yrbss_select_gender) 

obs_diff_gender

```

### {{< iconify fluent-mdl2 insights >}} Inference

:::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
#### Mann-Whitney test

Since the data variables do not satisfy the assumption of being
*normally distributed*, and the variances are significantly different,
we use the classical `wilcox.test`, which implements what we need here:
the *Mann-Whitney U test*,

Our model would be:

$$
mean(rank(Weight_{male})) - mean(rank(Weight_{female})) = \beta_1;
$$


$$
H_0: \mu_{weight-male} = \mu_{weight-female}
$$ 


$$
H_a: \mu_{weight-male} \ne \mu_{weight-female}
$$

Recall the earlier graph showing ranks of anxiety-scores against Gender.

```{r}
wilcox.test(weight ~ gender, data = yrbss_select_gender, 
            conf.int = TRUE, 
            conf.level = 0.95) %>% 
  broom::tidy()
```

The `p.value` is negligible and we are able to reject the NULL
hypothesis that the means are equal.

#### Linear Model

We can apply the *linear-model-as-inference* interpretation to the
**ranked data** data to implement the non-parametric test as a Linear
Model:

$$
lm(rank(weight) \sim  gender) = \beta_0 + \beta_1 * gender
$$ 


$$
H_0: \beta_1 = 0
$$ 

$$
H_a: \beta_1 \ne 0\\
$$

```{r}
#| results: hold

# Create a sign-rank function
#signed_rank <- function(x) {sign(x) * rank(abs(x))}

lm(rank(weight) ~ gender, 
   data = yrbss_select_gender) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95)

```

::: callout-tip
### Dummy Variables in `lm`

Note how the **Qual** variable was used here in Linear Regression
`lm()`! The `gender` variable was treated as a binary "dummy"
variable[^6].
:::


#### Permutation Test

For the specific data at hand, we need to shuffle the `gender` and take
the `test statistic` (difference in means) each time.

```{r}
#| label: fig-permutation-test-2
#| fig-cap: Null Distribution of Difference in Weights Means
#| fig-subcap:
#| - "Histogram of Null Distribution"
#| - "Cumulative Density of Null Distribution"
#| code-fold: true
#| results: hold
#| layout-ncol: 2
#| layout-nrow: 2
#| 
ggplot2::theme_set(new = theme_custom())

null_dist_weight <- 
  do(4999) * diffmean(data = yrbss_select_gender, 
                      weight ~ shuffle(gender))
null_dist_weight
###
prop1(~ diffmean <= obs_diff_gender, data = null_dist_weight)
###
gf_histogram(data = null_dist_weight, ~ diffmean, 
             bins = 25) %>%
  gf_vline(xintercept = obs_diff_gender, 
           colour = "red", linewidth = 1,
           title = "Null Distribution by Permutation", 
           subtitle = "Histogram") %>% 
  gf_labs(x = "Difference in Means") %>% 
  gf_refine(scale_y_continuous(expand = c(0, 0)))
###
gf_ecdf(data = null_dist_weight, ~ diffmean, 
        linewidth = 1) %>%
  gf_vline(xintercept = obs_diff_gender, 
           colour = "red", linewidth = 1,
           title = "Null Distribution by Permutation", 
           subtitle = "Cumulative Density") %>% 
  gf_labs(x = "Difference in Means") %>% 
  gf_refine(scale_y_continuous(expand = c(0, 0)))

```


Clearly the `observed_diff_weight` is much beyond anything we can
generate with permutations with `gender`! And hence there is a
significant difference in weights across `gender`!

::::



### All Tests Together

We can put all the test results together to get a few more insights
about the tests:

```{r}
#| label: tbl-All-Tests-tt-2
#| tbl-cap: All Tests for YRBSS Weights
#| tbl-subcap:
#| - "wilcox.test"
#| - "Linear Model with Ranked Data"
#| code-fold: true
#| results: hold

wilcox.test(weight ~ gender, data = yrbss_select_gender, 
            conf.int = TRUE, 
            conf.level = 0.95) %>% 
  broom::tidy() %>% tt() %>% 
  style_tt(i = 1,j = 3,background = "gold",color = "black")

lm(rank(weight) ~ gender, 
   data = yrbss_select_gender) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95) %>% tt() %>% 
  style_tt(i = 1,j = 5,background = "gold",color = "black")


```

The `wilcox.test` and the linear model with rank data offer the same
results. This is of course not surprising!


### {{< iconify fa-solid ring >}} One Test to Rule Them All: visStatistics again

We need to use a smaller sample of the dataset `yrbss_select_gender`, for the (same) reason: `visstat()` defaults to using the `shapiro.wilk` test internally:


```{r}
#| label: fig-visStatistics-2
#| fig-cap: visStatistics Output for YRBSS Weights Model
#| 
yrbss_select_gender_sample <- yrbss_select_gender %>% 
  slice_sample(n = 4999)

visstat(x = yrbss_select_gender_sample$gender, 
        y = yrbss_select_gender_sample$weight,
        conf.level = 0.95, numbers = TRUE) %>% 
  summary()
        
```

Compare these results with those calculated earlier!


## {{< iconify pajamas issue-type-test-case >}} Case Study #3: Weight vs Exercise in the YRBSS Survey

Finally, consider the possible relationship between a highschooler's
weight and their physical activity.

First, let's create a new variable `physical_3plus`, which will be coded
as either "yes" if the student is physically active for *at least* 3
days a week, and "no" if not. Recall that we have several *missing* data
in that column, so we will (sadly) drop these before generating the new
variable:

```{r}
yrbss_select_phy <- yrbss %>% 
  drop_na(physically_active_7d, weight) %>% 
  ## add new variable physical_3plus
  mutate(physical_3plus = if_else(physically_active_7d >= 3, 
                                  "yes", "no"),
         # Convert it to a factor Y/N
         physical_3plus = factor(physical_3plus, 
                                 labels = c("yes", "no"),
                                 levels = c("yes", "no"))) %>% 
  select(weight, physical_3plus)

# Let us check
yrbss_select_phy %>% count(physical_3plus)

```

::: callout-note
## Research Question

Does `weight` vary based on whether students exercise on more or less
than 3 days a week? (`physically_active_7d` \>= 3 days)
:::

### {{< iconify carbon chart-3d >}} Inspecting and Charting Data

We can make distribution plots for `weight` by `physical_3plus`:

```{r}
#| label: fig-yrbss-weight-phy
#| fig-cap: "Highschoolers' Weights by Days of Exercise"
#| fig-subcap:
#| - "Overlapped Density Plot"
#| - "Jitter Plots"
#| layout-ncol: 2
#| message: false
#| warning: false
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

###
gf_density(~ weight,
          fill = ~ physical_3plus, colour = ~ physical_3plus,
          data = yrbss_select_phy, alpha = 0.3, 
          title = "Highschoolers' Weights by Days of Exercise") %>% 
  gf_refine(scale_fill_viridis_d(name = "Days of Exercise >=3",
                                 option = "turbo",
                              aesthetics = c("colour", "fill")))
##
yrbss_select_phy %>% 
  gf_jitter(weight ~ physical_3plus, colour = ~ physical_3plus,
            width = 0.08, alpha = 0.2,
            xlab = "Days of Exercise >=3",
            title = "Student Weights vs Exercise Days") %>%  
   gf_summary(
    group = ~1, # See the reference link above. Damn!!!
    fun = "mean", geom = "line", colour = "lightblue",
    lty = 1, linewidth = 2
  ) %>% 
    gf_summary(fun = "mean", geom = "point", size = 3, colour = "gold") %>%
  gf_refine(scale_x_discrete(
    breaks = c("no", "yes"),
    labels = c("no", "yes"),
    guide = guide_prism_bracket(width = 0.1, outside = FALSE)
  ))  %>%
  gf_refine(
    annotate(x = 0.65, y = 60, geom = "text", label = "Mean Weights\n No Exercise"),
    annotate(x = 2.35, y = 60, geom = "text", label = "Mean Weights\nWith Exercise"),
    annotate(x = 1.5, y = 100, geom = "label", label = "Slope indicates\n differences in mean", fill = "moccasin"),
    scale_colour_viridis_d(name = "Days of Exercise >=3", option = "turbo")) %>%  gf_theme(theme(legend.position = "none", axis.line.x = element_line(linewidth = 1)))

```

The jitter and density plots show the comparison between the two means.
We can also compare the means of the distributions using the following
to first group the data by the `physical_3plus` variable, and then
calculate the mean `weight` in these groups using the `mean` function
while ignoring missing values by setting the `na.rm` argument to `TRUE`.

```{r by-means}
yrbss_select_phy %>%
  group_by(physical_3plus) %>%
  summarise(mean_weight = mean(weight, na.rm = TRUE))

```

There is an *observed difference*, but is this difference large enough
to deem it "statistically significant"? In order to answer this question
we will conduct a hypothesis test. But before that a few more checks on
the data:

#### A. {{< iconify mdi chart-bell-curve >}} Check for Normality

As stated before, statistical tests for means *usually* require a couple
of checks:\

-   Are the data normally distributed?\
-   Are the data variances similar?

Let us also complete a visual check for normality,with plots since we
cannot do a `shapiro.test`:

```{r}
#| label: fig-normality-check-3
#| fig-cap: "Highschoolers' Weights Normality Check"
#| fig-subcap:
#| - "Facetted Densities"
#| - "Q-Q Plots"
#| layout-ncol: 2
#| results: hold
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

yrbss_select_phy %>%
  gf_density( ~ weight,
              fill = ~ physical_3plus, color = ~ physical_3plus,
              alpha = 0.5,
              title = "Highschoolers' Weights by Exercise Frequency") %>%
  gf_facet_grid(~ physical_3plus) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_refine(scale_fill_viridis_d(name = "Days of Exercise >=3",
                                 option = "turbo",
                                 aesthetics = c("colour", "fill")))
##
yrbss_select_phy %>% 
  gf_qq(~ weight | physical_3plus , 
        color = ~ physical_3plus,
        title = "Q-Q Plots of Male and Female Weights") %>% 
  gf_qqline(ylab = "Weight") %>% 
  gf_refine(scale_fill_viridis_d(name = "Days of Exercise >=3",
                                 option = "turbo",
                                 aesthetics = c("colour", "fill")))

```

Again, not normally distributed...

#### B. {{< iconify mdi sigma-lower >}} Check for Variances

Let us check if the two variables have similar variances: the `var.test`
does this for us, with a NULL hypothesis that the variances are not
significantly different:

```{r}
#| message: false
#| results: hold
#| layout: [[85,15]]
var.test( weight ~ physical_3plus, 
          data = yrbss_select_phy, 
          conf.int = TRUE,
          conf.level = 0.95) %>% 
  broom::tidy()

# Critical F value
qf(0.975,4021, 8341)

```

The `p.value` states the probability of the data being what it is,
**assuming** the NULL hypothesis that variances were similar. It being
so small, we are able to reject this NULL Hypothesis that the variances
of `weight` are nearly equal across the two exercise frequencies.
(Compare the `statistic` in the `var.test` with the critical F-value)

::: callout-important
### Conditions

1.  The two variables are not normally distributed.
2.  The two variances are also significantly different.
:::

Hence we will have to use *non-parametric tests* to infer if the means
are similar.

### {{< iconify academicons hypothesis >}} Hypothesis

Based on the graphs, how would we formulate our Hypothesis? We wish to
infer whether there is difference in mean `weight` across
`physical_3plus`. So accordingly:

$$
H_0: \mu_{physical-3plus-Yes} = \mu_{physical-3plus-No}
$$ 


$$
H_a: \mu_{physical-3plus-Yes} \ne \mu_{physical-3plus-No}
$$

### {{< iconify academicons hypothesis >}} Observed and Test Statistic

What would be the **test statistic** we would use? The **difference in
means**. Is the observed difference in the means between the two groups
of scores non-zero? We use the `diffmean` function, from `mosaic`:

```{r}
#| layout: [[40,10,10]]

obs_diff_phy <- diffmean(weight ~ physical_3plus, 
                         data = yrbss_select_phy) 

obs_diff_phy

```

### {{< iconify fluent-mdl2 insights >}} Inference

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
#### *Parametric* t.test

Well, the variables are *not* normally distributed, and the variances
are significantly different so a standard `t.test` is not advised. We
can still try:

```{r}
mosaic::t_test(weight ~ physical_3plus,
               var.equal = FALSE, # Welch Correction
               data = yrbss_select_phy) %>% 
  broom::tidy()
```

The `p.value` is $8.9e-08$ ! And the `Confidence Interval` is clear of
$0$. So the `t.test` gives us good reason to reject the Null Hypothesis
that the means are similar. But can we really believe this, given the
non-normality of data?

#### *Non-Parametric* Wilcoxon test

However, we have seen that the data variables are *not* normally
distributed. So a **non-parametric** Wilcoxon Test, using signed-ranks, is indicated:
(recall the model!)

```{r}
#| warning: false
# For stability reasons, it may be advisable to use rounded data or to set digits.rank = 7, say, 
# such that determination of ties does not depend on very small numeric differences (see the example).

wilcox.test(weight ~ physical_3plus,
            conf.int = TRUE,
            conf.level = 0.95,
            data = yrbss_select_phy) %>% 
  broom::tidy()

```

The nonparametric `wilcox.test` also suggests that the means for
`weight` across `physical_3plus` are significantly different.

#### Linear Model Interpretation

We can apply the *linear-model-as-inference* interpretation to the
**ranked data** data to implement the non-parametric test as a Linear
Model:

$$
lm(rank(weight) \sim  physical.3plus) = \beta_0 + \beta_1 \times physical.3plus
\\
H_0: \beta_1 = 0\\
\\\
H_a: \beta_1 \ne 0\\
$$

```{r}
#| results: hold

lm(rank(weight) ~ physical_3plus, 
   data = yrbss_select_phy) %>% 
  broom::tidy(conf.int = TRUE,
              conf.level = 0.95)

```

Here too, the linear model using `rank` data arrives at a conclusion
similar to that of the Mann-Whitney U test.
:::

### Permutation Tests

For this last Case Study, we will do this in two ways, just for fun: one
using our familiar `mosaic` package, and the other using the package
`infer`.

But first, we need to initialize the test, which we will save as
`obs_diff_**`.

```{r}
#| label:  inf-weight-habit-ht-initial
#| warning: FALSE
#| layout-ncol: 3
obs_diff_infer <- yrbss_select_phy %>%
  infer::specify(weight ~ physical_3plus) %>%
  infer::calculate(stat = "diff in means", 
                   order = c("yes", "no"))
obs_diff_infer
##
obs_diff_mosaic <- 
  mosaic::diffmean(~ weight | physical_3plus, 
                   data = yrbss_select_phy)
obs_diff_mosaic
##
obs_diff_phy

```

::: callout-important
Note that `obs_diff_infer` is a 1 X 1 dataframe; `obs_diff_mosaic` is a
scalar!!
:::

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}
#### Using infer

Next, we will work through creating a permutation distribution using
tools from the **infer** package.

In `infer`, the `specify()` function is used to specify the variables
you are considering (notated `y ~ x`), and you can use the `calculate()`
function to specify the `stat`istic you want to calculate and the
`order` of subtraction you want to use. For this hypothesis, the
statistic you are searching for is the difference in means, with the
order being `yes - no`.

After you have calculated your observed statistic, you need to create a
permutation distribution. This is the distribution that is created by
shuffling the observed weights into new `physical_3plus` groups, labeled
"yes" and "no".

We will save the permutation distribution as `null_dist`.

```{r}
#| label: inf-weight-habit-ht-null
#| warning: FALSE
#| cache: TRUE
#| layout: [[60,40]]

null_dist <- yrbss_select_phy %>%
  specify(weight ~ physical_3plus) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 999, type = "permute") %>%
  calculate(stat = "diff in means", 
            order = c("yes", "no"))
null_dist

```

The `hypothesize()` function is used to declare what the null hypothesis
is. Here, we are assuming that student's weight is independent of
whether they exercise at least 3 days or not.

We should also note that the `type` argument within `generate()` is set
to `"permute"`. This ensures that the statistics calculated by the
`calculate()` function come from a reshuffling of the data (not a
resampling of the data)! Finally, the `specify()` and `calculate()`
steps should look familiar, since they are the same as what we used to
find the observed difference in means!

We can visualize this null distribution with the following code:

```{r}
#| label: fig-inf-weight-habit-ht-visualise
#| fig-cap: Null Distribution of Difference in Weights Means
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

null_dist %>% 
  visualise() + # Note this plus sign! 
  shade_p_value(obs_diff_infer,
                direction = "two-sided") + 
  scale_y_continuous(expand = c(0, 0))

```

Now that you have calculated the observed statistic and generated a
permutation distribution, you can calculate the p-value for your
hypothesis test using the function `get_p_value()` from the infer
package.

```{r}
#| label:  inf-weight-habit-ht-pvalue
null_dist %>%
  get_p_value(obs_stat = obs_diff_infer, 
              direction = "two_sided")

```

What warning message do you get? Why do you think you get this warning
message? Let us construct and record a confidence interval for the
difference between the weights of those who exercise at least three
times a week and those who don't, and interpret this interval in context
of the data.

```{r}
#| label:  inf-weight-habit-ht-confint
null_dist %>%
  infer::get_confidence_interval(
    point_estimate = obs_diff_infer, 
    level = 0.95)

```

It does look like the `observed_diff_infer` is too far away from this
confidence interval. Hence if there was no difference in `weight` caused
by `physical_3plus`, we would never have observed it! Hence the
`physical_3plus` does have an effect on `weight`!

#### Using mosaic

We already have the observed difference, `obs_diff_mosaic`. Now we
generate the null distribution using permutation, with `mosaic`:

```{r}
null_dist_mosaic <- do(999) * 
  diffmean(~ weight | shuffle(physical_3plus), 
           data = yrbss_select_phy)

```

We can also generate the histogram of the null distribution, compare
that with the `obs`erved `diff`rence and compute the `p-value` and
`confidence intervals`:

```{r}
#| label: fig-permutation-test-3
#| fig-cap: Null Distribution of Difference in Weights Means
#| code-fold: true
#| 
ggplot2::theme_set(new = theme_custom())

gf_histogram(~ diffmean, data = null_dist_mosaic) %>% 
  gf_vline(xintercept = obs_diff_mosaic, 
           colour = "red", linewidth = 2) %>% 
  gf_refine(scale_y_continuous(expand = c(0, 0)))
            

```


```{r}
#| label: inf-weight-habit-ht-mosaic-pvalue
# p-value
prop(~ diffmean != obs_diff_mosaic, data = null_dist_mosaic)
# Confidence Intervals for p = 0.95
mosaic::cdata(~ diffmean, p = 0.95, data = null_dist_mosaic)

```

Again, it does look like the `observed_diff_infer` is too far away from
this NULL distribution. Hence if there was no difference in `weight`
caused by `physical_3plus`, we would never have observed it! Hence the
`physical_3plus` does have an effect on `weight`!
:::

Clearly there is a serious effect of Physical Exercise on the body
weights of students in the population from which this dataset is drawn.

## {{< iconify mingcute thought-line >}} Wait, But Why?

-   In our target population for our design intervention, we often perceive groups
-   We need often to infer differences in some Quantitative
    variable between these groups
-   We **treat** our dataset as a **sample** from the population, which
    we cannot access
-   We can apply all the CLT ideas +`t.test` if the two variables in the
    dataset satisfy the conditions of normality, equal variance
-   Else use non-parametric `wilcox.test`
-   And by treating the two variables as *one Quant in two groups*, we
    can simply perform a Permutation test

## {{< iconify fluent-mdl2 decision-solid >}} Conclusion

-   We have learnt how to perform inference for independent means.\
-   We have looked at the conditions that make the regular `t.test`
    possible, and learnt what to do if the conditions of normality and
    equal variance are not met.\
-   We have also looked at how these tests can be understood as
    manifestations of the *linear model*, with data and sign-ranked
    data.\

## {{< iconify openmoji person >}} Your Turn

1.  Try the `SwimRecords` dataset from the `mosaicData` package.

2.  Try some of the datasets in the `moderndive` package. Install it ,
    peasants. And type in your Console `data(package = "moderndive")` to
    see what you have. Teacher `evals` might interest you!

## {{< iconify ooui references-rtl >}} References {#sec-references}

1.  Randall Pruim, Nicholas J. Horton, Daniel T. Kaplan, [*Start
    Teaching with
    R*](https://github.com/ProjectMOSAIC/LittleBooks/raw/master/Starting/MOSAIC-StartTeaching.pdf)
2.  <https://bcs.wiley.com/he-bcs/Books?action=index&itemId=111941654X&bcsId=11307>
3.  <https://statsandr.com/blog/wilcoxon-test-in-r-how-to-compare-2-groups-under-the-non-normality-assumption/>

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c( "resampledata", "openintro", "infer", "TeachHist", "TeachingDemos", "visStatistics", "tinytable")
) %>%
  knitr::kable(format = "simple")

```
:::

[^1]: <https://stats.stackexchange.com/questions/2492/is-normality-testing-essentially-useless>

[^2]: <https://www.allendowney.com/blog/2023/01/28/never-test-for-normality/>

[^3]: <https://stats.stackexchange.com/q/113337>

[^4]: <https://www.middleprofessor.com/files/applied-biostatistics_bookdown/_book/intro-linear-models.html#a-linear-model-can-be-fit-to-data-with-continuous-discrete-or-categorical-x-variables>

[^5]: <https://stats.stackexchange.com/questions/92374/testing-large-dataset-for-normality-how-and-is-it-reliable>

[^6]: <https://en.wikipedia.org/wiki/Dummy_variable_(statistics)>
