---
title: "Highlighting parts of the Chart"
subtitle: ""
author: Arvind V. 
date: 2025/8/30
date-modified: "`r Sys.Date()`"
order: 40
bibliography: 
  - grateful-refs.bib
---

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false
#| 
library(tidyverse)  
library(ggformula)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(paletteer)  ## access to many color palettes
library(showtext)   ## add google fonts to plots
library(ggtext)      ## add improved text rendering to ggplot2
##
library(ggforce)     ## add missing functionality to ggplot2
library(concaveman) # for convex hulls with ggforce
library(ggfocus)  # Scales which help add emphasis to *any* geom!
library(gghighlight) ## highlight parts of ggplot2 plots

```

## Introduction

We may have been using `ggplot2` for a while now, but there is always more to learn about the aesthetics of our plots. Here we will look at ways to highlight specific areas of our plots, to draw attention to them. We will introduce three packages that can help us do this: `{{ggfocus}}` and `{{gghighlight}}`. 

### Fonts and Themes

As we learnt in the [Fonts Module](../20-Fonts/index.qmd), let us set the ggplot theme with fonts for both the chart defining elements **and** our annotations. I am going to use local fonts from my machine, that I have downloaded from Google Fonts earlier and saved in a `fonts` folder at the root of my project.

```{r}
#| label: add-local-fonts
#| code-fold: true
#| 
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Anton",
  regular = "../../../../../../fonts/Anton-Regular.ttf") # Only one style is available

sysfonts::font_add(family = "RobotoCondensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")

sysfonts::font_add(family = "IbarraNova", 
  regular = "../../../../../../fonts/IbarraRealNova-Regular.ttf",
  bold = "../../../../../../fonts/IbarraRealNova-Bold.ttf",
  italic = "../../../../../../fonts/IbarraRealNova-Italic.ttf",
  bolditalic = "../../../../../../fonts/IbarraRealNova-BoldItalic.ttf")

sysfonts::font_add(family = "Tangerine", 
  regular = "../../../../../../fonts/Tangerine-Regular.ttf",
  bold = "../../../../../../fonts/Tangerine-Bold.ttf") # only these two are available

sysfonts::font_add(family = "Schoolbell", 
  regular = "../../../../../../fonts/Schoolbell-Regular.ttf") # Only regular is available

showtext_auto() # set these fonts as default for the session

```


```{r}
#| label: set-ggplot2-theme
#| code-fold: true
#| 

theme_custom <- function(){ 
    font <- "RobotoCondensed"   # Assign font family up front
    
    theme_bw(base_size = 14, base_family = font) %+replace%    
      
      ## modify plot elements globally (for all following plots)
      theme(
        axis.ticks = element_line(color = "grey92"),
        axis.ticks.length = unit(.5, "lines"),
        
        panel.grid.minor = element_blank(),
        
        legend.title = element_text(family = "Schoolbell", size = 12),
        legend.text = element_text(family = "IbarraNova", color = "grey30"),
        
        plot.title = element_text(               #title
                   family = "Alegreya",          #set font family
                   size = 18,                    #set font size
                   face = 'bold',                #bold typeface
                   hjust = 0,                  #left align
                   margin = margin(t = 5, r = 0, b = 5, l = 0)), #margin
      
        plot.title.position = "plot",
        
        plot.subtitle = element_text(family = "Alegreya", size = 12, 
                                     color = "grey30",
                                     hjust = 0,                   #left align
                   margin = margin(t = 5, r = 0, b = 10, l = 0)), #margin
      
       plot.caption = element_text(             #caption
                   family = font,                #font family
                   size = 9,                     #font size
                   hjust = 1),                   #right align
       
      plot.caption.position = "plot",            #right align to plot
)
}

## Set the theme
ggplot2::theme_set(new = theme_custom())


## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text",new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "RobotoCondensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

```



## Data

We will again work with a **familiar** dataset, so that we can concentrate on the chart aesthetics, without having to spend time getting used to the data: the `penguins` dataset again, ~~from the `palmerpenguins` package~~ now part of the R installation! 

As always, we will start with a **table** of the data:

```{r}
#| label: tbl-read-penguins-data
#| tbl-cap: "Penguins Dataset"
#| message: false
#| code-fold: true

penguins <- penguins %>% drop_na() # remove data containing missing data

## Create a nicely formatted table
## uses `kableExtra` package
## 
penguins %>% 
  kableExtra::kbl(caption = "Penguins Dataset") %>%
  kableExtra::kable_paper(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    position = "center"
  ) %>%
  kableExtra::scroll_box(height = "350px")

```

## Basic Plots

A basic scatter plot, which we will use to highlight and draw attention to specific areas of the plot. We will include, for the sake of retaining context, the fonts and labels that we have used in previous modules. And we will also use `ggtext` to format the axis titles and the plot title.

We will colour one plot using a **Quant** variable, and the other using a **Qual** variable, to show that the highlighting methods work with both types of variables.


```{r}
#| label: fig-basic-ggformula-scatter
#| fig-cap: "Basic Scatter Plot coloured by Body Mass"
#| code-fold: true


p1 <- gf_point(data = penguins, bill_dep ~ bill_len, 
              color = ~ body_mass, alpha = 0.8, size = 3.5) %>% 

  ## custom colors from the scico package
  gf_refine(scico::scale_color_scico(palette = "tokyo", direction = -1)) %>% 
  
  ## custom labels
  gf_labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (**Pygoscelis**)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) PLoS ONE',
    x = '**Bill Length** (mm)', 
    y = '**Bill Depth** (mm)',
    color = 'Body mass (g)') %>% 
   
  ## 2. Add theme related commands from ggtext
  ## render respective text elements
  gf_theme(theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown()
  ))


p1

```
And another scatter plot:

```{r}
#| label: fig-basic-ggplot2-scatter
#| fig-cap: "Basic Scatter Plot coloured by Species"
#| code-fold: true
#| 
p2 <- ggplot(penguins, aes(x = bill_len, y = bill_dep, 
                           colour = species)) + 
  geom_point(size = 3.5) + 
  
  # Add colours
  scale_color_paletteer_d("ggsci::category10_d3") +
  
  # Add labels
  labs(
    title = "Bill Dimensions of Brush-Tailed Penguins (*Pygoscelis*)",
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = "Data: Gorman, Williams & Fraser (2014) *PLoS ONE*",
    x = "**Bill Length** (mm)", 
    y = "**Bill Depth** (mm)",
    color = "Body mass (g)",
    fill = "Species"
  )  + 
  ## 2. Add theme related commands from ggtext
  ## render respective text elements
  gf_theme(theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown()
  ))

p2

```

## Using `{ggfocus}`

`{{ggfocus}}` provides ***scales*** which help add emphasis to *any* geom! It works by *fading out the non-focused data*, and can be used with any geom, including those from other packages. It is particularly useful for highlighting points in scatter plots, lines in line plots, or bars in bar plots. Some examples of these scales are:

- `scale_focus_color()`: Highlights points based on a condition, changing their color and adjusting the transparency of non-highlighted points.
- `scale_focus_fill()`: for filled geoms like bars or areas.
- `scale_focus_size()`: adjusts the size of the highlighted points instead of color.
- `scale_focus_alpha()`: Adjusts the transparency of points based on a condition, making non-highlighted points more transparent.
- `scale_focus_linetype()`: Highlights lines in line plots based on a condition by changing their linetype and adjusting the transparency of non-highlighted lines.

```{r}
#| label: fig-ggfocus-example
#| fig-cap: "Using `ggfocus` to highlight points in a scatter plot"

p2 +
  ggfocus::scale_color_focus(
    focus_levels  = c("Gentoo"),
    color_focus = "red",          # color for highlighted points
    color_other = "grey80",       # color for non-highlighted points
  ) + 
  labs(subtitle = "Highlighting Gentoo Penguins using ggfocus")

```

We can also highlight for `fill`, using the `ggformula` idiom this time:

```{r}
#| label: fig-ggfocus-example-2
#| fig-cap: "Using `ggfocus` to highlight points in a histogram based on a categorical condition"
#| message: false
#| warning: false

gf_histogram(~ body_mass, data = penguins, fill = ~ species) %>%
  
  gf_refine(ggfocus::scale_fill_focus(
    focus_levels = c("Gentoo"), # condition to highlight points
    color_focus = "dodgerblue",          # color for highlighted points
    color_other = "grey80",       # color for non-highlighted points
  )) %>% 
  
  gf_labs(title = "Penguins Body Mass Histograms",
          subtitle = "Highlighting Gentoo Penguins using ggfocus")

```

We see that the scales from `ggfocus` directly use the ***existing*** aesthetics of the plot, and modify them to highlight specific data points. This is different from the other two packages we will see, which add new geoms to the plot to create the highlighting effect.

## Using `{gghighlight}`

To be Written Up. Yeah. 


## {{< iconify ooui references-ltr >}} References

1. Yutani, H. (2023). *ggfocus: Focus Scales for ggplot2*. R package version 0.1.0. <https://CRAN.R-project.org/package=ggfocus>

::: {#refs style="font-size: 60%;"}
###### {{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
library(grateful)
#scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("gghighlight", "ggforce", "ggtext", "ggfocus")
) %>%
  knitr::kable(format = "simple")

```
:::
