---
title: "Inference for Correlation"
date: 25/Nov/2022
date-modified: today
order: 150
keywords: Statistics ; Tests; p-value
abstract: "Statistical Significance Tests for Correlations between two Variables"
bibliography: 
  - grateful-refs.bib
citation: true
#suppress-bibliography: true
editor: 
  markdown: 
    wrap: 72
fig-align: "center"
format: 
  html: default
  revealjs: 
    theme: default
    output-file: counts-reveal.html
    smaller: true
    center: false
    scroll-view: true
    scrollable: true
    slide-level: 3
    slide-number: true
    chalkboard: true
    df-print: paged
    highlight-style: tango
    menu: true
    overview: true
    preload-iframes: true
    preview-links: true
    touch: true

---

## {{< iconify noto-v1 package >}} Setting up R packages

```{r}
#| label: setup
#| warning: false
#| message: false
#| include: true

library(tidyverse) # Sine Qua Non
library(mosaic) # Our bag of tricks
library(broom) # Tidying model outputs
library(crosstable) # tabulated summary stats
library(openintro) # datasets and methods
library(resampledata3) # datasets
library(mosaicData) # datasets
library(statsExpressions) # datasets and methods
library(ggstatsplot) # special stats plots
library(ggExtra)

# Non-CRAN Packages
# remotes::install_github("easystats/easystats")
library(easystats)

```

```{r}
#| label: Extra-Pedagogical-Packages
#| echo: false
#| message: false

library(checkdown)
library(epoxy)
library(TeachHist)
library(TeachingDemos)
library(grateful)
library(ggrepel)
library(marquee)

```


#### Plot Fonts and Theme

```{r}
#| label: plot-theme-ggplot4
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false
#| 
library(systemfonts)
library(showtext)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 

    theme_bw(base_size = 10) + 
    
    # theme(panel.widths = unit(11, "cm"), 
    #       panel.heights = unit(6.79, "cm")) + # Golden Ratio
    
    theme(plot.margin = margin_auto(t = 1, r = 2, b = 1, l = 1, unit = "cm"),
          plot.background = element_rect(fill = "bisque",
                                         colour = "black",
                                         linewidth = 1)) + 
            
    theme_sub_axis(title = element_text(family = "Roboto Condensed", 
                                       size = 10),
                   text = element_text(family = "Roboto Condensed", 
                                       size = 8)) + 
    
    theme_sub_legend(text = element_text(family = "Roboto Condensed", 
                                         size = 6),
                     title = element_text(family = "Alegreya", 
                                          size = 8)) + 
    
    theme_sub_plot(title = element_text(family = "Alegreya", 
                                        size = 14, face = "bold"),
                   title.position = "plot",
                   subtitle = element_text(family = "Alegreya", 
                                           size = 10),
                   caption = element_text(family = "Alegreya", 
                                          size = 6),
                   caption.position = "plot") 
    
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

ggplot2::update_geom_defaults(geom = "marquee", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "text_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

## tinytable options
options("tinytable_tt_digits" = 2)
options("tinytable_format_num_fmt" = "significant_cell")
options(tinytable_html_mathjax = TRUE)


## Set defaults for flextable
flextable::set_flextable_defaults(font.family = "Roboto Condensed")

```


## {{< iconify openmoji japanese-symbol-for-beginner >}} Introduction

Correlations define how one variables varies with another. One of the basic Questions we would have of our data is: Does some variable have a significant correlation score with another in some way? Does $y$ vary with $x$? A **Correlation Test** is designed to answer exactly this question.  The block diagram below depicts the statistical procedures available to test for the significance of correlation scores between two variables. 

In this module we will explore the **correlation coefficient** and how to test for its significance. We will also see how to use the **linear model** method to perform correlation tests, and how to use the **permutation test** to do so without any assumptions.


### {{< iconify flat-color-icons workflow >}} Basic Definitions

Before we begin, let us recap a few basic definitions:

We have already encountered the `variance` of a variable:

$$
\begin{align*}
var_x &= \frac{\sum_{i=1}^{n}(x_i - \mu_x)^2}{(n-1)}\\
where ~ \mu_x &= mean(x)\\
n &= sample\ size
\end{align*}
$$
The *standard deviation* is:

$$
\sigma_x = \sqrt{var_x}\\
$$
The **covariance** of two variables is defined as:

$$
\begin{align}
cov(x,y) &= \frac{\sum_{i = 1}^{n}(x_i - \mu_x)*(y_i - \mu_y)}{n-1}\\
&= \frac{\sum{x_i *y_i}}{n-1} - \frac{\sum{x_i *\mu_y}}{n-1} - \frac{\sum{y_i *\mu_x}}{n-1} + \frac{\sum{\mu_x *\mu_y}}{n-1}\\
&= \frac{\sum{x_i *y_i}}{n-1} - \frac{\sum{\mu_x *\mu_y}}{n-1}\\
\end{align}
$$

Hence covariance is *the expectation of the product minus the product of the expectations* of the two variables. 

So, finally, the coefficient of **correlation** between two variables is defined as:

$$
\begin{align}
correlation ~ r &= \frac{cov(x,y)}{\sigma_x * \sigma_y}\\
&= \frac{\sum_{i = 1}^{n}(x_i - \mu_x)*(y_i - \mu_y)}{(\sigma_x * \sigma_y)(n-1)}\\
&= \frac{\sum_{i = 1}^{n}\left(\frac{x_i - \mu_x}{\sigma_x}\right)*\left(\frac{y_i - \mu_y}{\sigma_y}\right)}{(n-1)}\\
\end{align}
$${#eq-corr}

which is the average of the product of the **z-scores** of the two variables.

::: callout-tip
### Correlation uses [z-scores](../../../10-Descriptive/Modules/22-Quantities/index.qmd)!
Note that in both cases we are dealing with **z-scores**: variable minus its mean, $\frac{x_i - \mu_x}{\sigma_x}$, which we have seen when dealing with the CLT and the Gaussian Distribution. 
:::

## {{< iconify pajamas issue-type-test-case >}} Case Study #1: Galton's famous dataset

How can we start, except by using the famous `Galton` dataset, now part of the `mosaicData` package?

### {{< iconify flat-color-icons workflow >}} Workflow: Read and Inspect the Data

```{r}
#| label: read-galton
data("Galton", package = "mosaicData")
Galton
```

```{r}
skimr::skim(Galton)

```

So there are several correlations we can explore here: Children's `height` vs that of `father` or `mother`, based on `sex`. In essence we are replicating Francis Galton's famous study.

### {{< iconify carbon clean >}} Data Munging

Note that `nkids`, while coded as `int`, is actually a `factor` variable. Let us convert it to a factor:

```{r}
Galton_modified <- Galton %>% 
  mutate(nkids = as_factor(nkids))
```

### {{< iconify streamline dictionary-language-book-solid >}} Data Dictionary

::: callout-note
### Qualitative Variables

- `sex`(fct): sex of the child
- `family`(int): an ID for each family
- `nkids`(fct): Number of children in each family
:::

::: callout-note
### Quantitative Variables

- `father`(dbl): father's height in inches
- `mother`(dbl): mother's height in inches
- `height`(dbl): Child's height in inches

:::

### {{< iconify flat-color-icons workflow >}} Workflow: Experiment, and Research Questions

We can say that Galton may have measured the heights of fathers/mothers and their children, and recorded the `sex` of the child. He may have been interested in knowing if there is a correlation between the heights of the parents and their children, and if this correlation is different for sons and daughters. Hence the children's `height` is the *response* variable, and the `father`'s/`mother`'s heights are *explanatory* variables, as is `sex` of the child.

::: callout-note
## Question 1
1.  Based on this sample, what can we say about the correlation between a son's height and a father's height in the population?
:::

::: callout-note
## Question 2
1.  Based on this sample, what can we say about the correlation between a daughter's height and a father's height in the population?
:::
Of course we can formulate more questions, but these are good for now! And since we are going to infer correlations by `sex`, let us split the dataset into two parts, one for the sons and one for the daughters, and quickly summarise them too:

```{r}
Galton_sons <- Galton_modified %>% 
  dplyr::filter(sex == "M") %>% 
  rename("son" = height)
Galton_daughters <- Galton_modified %>% 
  dplyr::filter(sex == "F") %>% 
  rename("daughter" = height)

```

:::: {layout="[50,50]"}

::: {#first-column}
```{r}
dim(Galton_sons)

```

Sons Data

:::

::: {#second-column}
```{r}
dim(Galton_daughters)

```
Daughters data

:::
::::





:::: {layout="[50,50]"}

::: {#first-column}

```{r}
Galton_sons %>% 
  summarize(across(.cols = c(son, father), 
                   .fns = list(mean = mean, sd = sd)))

```

Sons Summary Stats

:::

::: {#second-column}
```{r}
Galton_daughters %>% 
  summarize(across(.cols = c(daughter, father), 
                   .fns = list(mean = mean, sd = sd)))
```

Daughters Summary Stats
:::


::::


::: callout-note
### What did filtering do?
Why are `father` means different for sons and daughters?? 
When we filtered the dataset into two, the filtering by `sex` of the child also effectively filtered the heights of the `father` (and `mother`). This is proper and desired; but think!

:::

### {{< iconify flat-color-icons workflow >}} Workflow: Visualization

Let us first quickly plot a graph that is relevant to each of the two research questions.

```{r}
#| label: fig-plot-galton
#| fig-cap: "Heights of Sons and Daughters vs Fathers"
#| fig-subcap:
#|   - Sons
#|   - Daughters
#| layout-ncol: 2
#| code-fold: true
#| out-width: "75%"
#| out-height: "75%"
#| 
ggplot2::theme_set(new = theme_custom())

Galton_sons %>% 
  gf_point(son ~ father) %>% 
  gf_lm() %>% 
  gf_labs(x = "Father's Height", y = "Son's Height",
          title = "Heights: Sons vs Fathers",
          subtitle = "Galton dataset")
##
Galton_daughters %>% 
  gf_point(daughter ~ father) %>% 
  gf_lm() %>% 
  gf_labs(x = "Father's Height", y = "Daughter's Height",
          title = "Heights: Daughters vs Fathers",
          subtitle = "Galton dataset")
```

We might even plot the **overall** `height`s together and colour by `sex` of the child:

```{r}
#| out-width: "50%"
#| out-height: "50%"
#| fig-align: center
#| 
ggplot2::theme_set(new = theme_custom())

Galton %>% 
  gf_point(height ~ father, 
           group = ~ sex, colour = ~ sex) %>% 
  gf_lm() %>% 
  gf_refine(scale_color_brewer(palette = "Set1")) %>%
  gf_labs(x = "Father's Height", y = "Children's Height",
          title = "Heights: Children vs Fathers",
          subtitle = "Galton dataset")

```

So daughters are shorter than sons, generally speaking, and both sets of heights seem related to that of the father. 

### {{< iconify flat-color-icons workflow >}} Workflow: Assumptions

For the *classical* correlation tests, we need that the variables are normally distributed. As before we check this with the `shapiro.test`:

```{r}
#| layout-ncol: 2
shapiro.test(Galton_sons$father)
shapiro.test(Galton_sons$son)
##
shapiro.test(Galton_daughters$father)
shapiro.test(Galton_daughters$daughter)

```

Let us also check the densities and quartile plots of the `height`s in the dataset: 

```{r}
#| label: fig-plot-galton-density-quartile
#| fig-cap: "Checking Children's Heights for Normality"
#| fig-subcap:
#|   - Densities
#|   - Q-Q plots
#| layout-ncol: 2
#| code-fold: true
#| out-width: "75%"
#| out-height: "75%"
#| 
ggplot2::theme_set(new = theme_custom())

Galton %>% group_by(sex) %>% 
  gf_density(~ height, group = ~ sex, 
             fill = ~ sex) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_refine(scale_fill_brewer(palette = "Set1")) %>%
  gf_facet_grid(vars(sex)) %>% 
  gf_labs(title = "Facetted Density Plots") %>% 
  gf_theme(legend.position = "none") # Think!
##
Galton %>% group_by(sex) %>% 
  gf_qq(~ height, group = ~ sex, 
             colour = ~ sex, size = 0.5) %>% 
  gf_qqline(colour = "black") %>% 
  gf_refine(scale_color_brewer(palette = "Set1")) %>%
  gf_facet_grid(vars(sex)) %>% 
  gf_labs(title = "Facetted QQ Plots",
          x = "Theoretical quartiles",
          y = "Actual Data") %>% 
  gf_theme(legend.position = "none") # Think!

```

and the `father`'s heights:

```{r}
#| label: fig-plot-galton-fathers-distributions
#| fig-cap: "Are Dads Normally Distributed?"
#| fig-subcap:
#|   - Fathers Distributions facetted by Child Sex
#|   - Fathers Q-Q Plots facetted by Child Sex
#| layout-ncol: 2
#| code-fold: true
#| out-width: "75%"
#| out-height: "75%"
#| 
ggplot2::theme_set(new = theme_custom())

Galton %>% group_by(sex) %>% 
  gf_density(~ father, group = ~ sex, # no this is not weird
             fill = ~ sex) %>% 
  gf_fitdistr(dist = "dnorm") %>% 
  gf_refine(scale_fill_brewer(name = "Sex of Child", palette = "Set1")) %>%
  gf_facet_grid(vars(sex)) %>% 
  gf_labs(title = "Fathers: Facetted Density Plots",
                    subtitle = "By Sex of Child") %>% 
  gf_theme(legend.position = "none") # Think!

Galton %>% group_by(sex) %>% 
  gf_qq(~ father, group = ~ sex, # no this is not weird
             colour = ~ sex, size = 0.5) %>% 
  gf_qqline(colour = "black") %>% 
  gf_facet_grid(vars(sex)) %>% 
  gf_refine(scale_colour_brewer(name = "Sex of Child", palette = "Set1")) %>% 
  gf_labs(title = "Fathers Heights: Facetted QQ Plots",
          subtitle = "By Sex of Child",
          x = "Theoretical quartiles",
          y = "Actual Data") %>% 
  gf_theme(legend.position = "none") # Think!

```

The `shapiro.test` informs us that the child-related `height` variables are *not normally distributed*; though visually there seems nothing much to complain about. Hmmm...

[Dads are weird anyway](https://xpressenglish.com/our-stories/conversation-with-father/), so we must not expect `father` heights to be normally distributed. 


### {{< iconify flat-color-icons workflow >}} Workflow: Inference

Let us now see how Correlation Tests can be performed  based on this dataset, to infer patterns in the population from which this dataset/sample was drawn.

We will go with classical tests first, and then set up a permutation test that does not need any assumptions. 

::: {.panel-tabset .nav-pills style="background: whitesmoke;"}

#### Classical Tests
We perform the Pearson correlation test first: the data is **not normal** so we cannot really use this. We should use a *non-parametric* correlation test as well, using a `Spearman` correlation.

```{r}
#| label: corr-tests-sons
# Pearson (built-in test)
cor_son_pearson <- mosaic::cor.test(son ~ father, 
                        method = "pearson", 
                        data = Galton_sons) %>% 
  broom::tidy() %>% 
  mutate(term = "Pearson Correlation r")
cor_son_pearson

cor_son_spearman <- mosaic::cor.test(son ~ father, 
                                     method = "spearman", 
                                     data = Galton_sons) %>% 
  broom::tidy() %>% 
  mutate(term = "Spearman Correlation r")
cor_son_spearman

```

Both tests state that the correlation between `son` and `father` is significant. 

```{r}
#| label: corr-tests-daughters
# Pearson (built-in test)
cor_daughter_pearson <- mosaic::cor.test(daughter ~ father, 
                        method = "pearson", 
                        data = Galton_daughters) %>% 
  broom::tidy() %>% 
  mutate(term = "Pearson Correlation r")
cor_daughter_pearson

cor_daughter_spearman <- mosaic::cor.test(daughter ~ father, 
                                          method = "spearman", 
                                          data = Galton_daughters) %>% 
  broom::tidy() %>% 
  mutate(term = "Spearman Correlation r")
cor_daughter_spearman

```

Again both tests state that the correlation between `daughter` and `father` is significant. 

#### {{< iconify game-icons dragon-head size=2x >}} Intuitive

What is happening under the hood in `cor.test`?

Given that the correlation coefficient $r$ is a measure of the linear relationship between two variables, we can test its significance using a t-test. The formula for the t-value in a correlation test is derived from the relationship between the correlation coefficient and the t-distribution.

The formula for the t-value is given by:
$$t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}$${#eq-tstat}
where:

- $t$ is the t-statistic,
- $r$ is the Pearson correlation coefficient,
- $n$ is the number of paired observations (sample size).
- The degrees of freedom for this t-test is $df = n - 2$.

The derivation of this t-statistic stems from the fact that the correlation coefficient can be expressed in terms of the slope of the regression line when one variable is regressed on the other. The t-test essentially tests whether the slope of this regression line is significantly different from zero, which would indicate a significant linear relationship between the two variables.

::: callout-warning
#### {{< iconify game-icons dragon-head size=2x >}} Dragons Ahead!!
OK, if you like, you can stop here! But for you intrepid, [beamish people who possess vorpal swords](https://www.poetryfoundation.org/poems/42916/jabberwocky), here come the Dragons!!
:::

In [Linear Regression](../../../30-Modelling/Modules/20-LinReg/index.qmd), we saw that the `F-statistic` is a ratio of variances, and that it follows an `F-distribution`. How does this relate to correlation?

- In regression we have a *target* variable (sons' `height`s) and a *predictor* variable (`father`s` heights), same as with our present study of correlation.
- We look at how much one variable *explains*, or reduces the variance the other
- i.e. How much the **variance** of the target variable (sons' `height`s) is reduced by the fact that we know the value(s) of a predictor variable (`father`s` heights).
- Our measure of how well this reduction is happening is a ratio: a ratio of variances, also denoted as $r^2$, i.e. the square of the correlation coefficient.
- We take the variance of the target variable, and the variance of the target variable *after* we have accounted for the predictor variable. 
- The ratio of variances follows an `F-distribution` with appropriate degrees of freedom
- This gives us our `F-value`, computed from our data. 
    - We compare our computed `F-value` with the **critical F-value** `F-crit` for a probability of error of $5\%$, given by the `F-distribution` with appropriate degrees of freedom.
    - If our `F-value` is well above `F_crit`, we state there is there very low probability (i.e.`p-value`) that this reduction happened simply by change and accept the hypothesis that there is significant correlation between the two variables. 
- We relate the idea of *Regression* to *Correlation* by noting that the `t-value` for correlation must be simply the square root of the `F-value` for regression, since the `F-distribution` is for $r^2$ and the `t-distribution` is for $r$.

::: callout-important
### A Distribution for a Variance-Ratio??
Why does the variance-ratio have a distribution??? The two variances appear to be single numbers !!! Is there anything that is ***not random*** in statistics?

- Remember, we treat our data as a ***sample*** from a ***population***, in order to estimate the `regression slope` for the **population**.
- The sample is random, and hence the variance of the sample is also random. 
- If we took another sample, we would get a different variance. 
- So the variance of a sample is a random variable, and hence the *ratio-of-variances* also has a distribution.
Phew!
:::

::: callout-important
### The F-statistic and its Distribution
Why does the *ratio-of-variances* have an F-distribution? 

- In our case, our **residuals** (deviations from the means) are assumed to be **normal**. 
- The variance calculation ***squares*** these normally-distributed residuals, leading to a *chi-square* distribution for the individual variances. 
- *The ratio of these two independent chi-squared variables, each divided by their respective degrees of freedom, follows an F-distribution*. 
- This is a fundamental result in statistics that underpins the use of the F-test in statistical analysis.
:::

#### {{< iconify game-icons dragon-head size=2x >}} Derivation

To derive the formula for the `t-value` in a correlation test, starting with a ratio of variances, we need to focus on the t-test for the significance of the Pearson correlation coefficient. The t-value assesses whether the observed correlation coefficient $r$ is significantly different from zero. Let’s proceed step-by-step, connecting the t-test to variances and ensuring a clear derivation.

###### Step 1: Understanding the Pearson Correlation Coefficient

The Pearson correlation coefficient $r$ measures the linear relationship between two variables $X$ and $Y$. It is defined as:

$$
\begin{equation}
\begin{split}
r &= \frac{\text{Cov}(X, Y)}{\sqrt{\text{Var}(X) \text{Var}(Y)}}\\
&= \frac{\sum (x_i - \bar{x})(y_i - \bar{y}) / (n-1)}{\sqrt{\left( \sum (x_i - \bar{x})^2 / (n-1) \right) \left( \sum (y_i - \bar{y})^2 / (n-1) \right)}}
\end{split}
\end{equation}
$$
where:

$\text{Cov}(X, Y)$ is the covariance of $X$ and $Y$,
$\text{Var}(X)$ and $\text{Var}(Y)$ are the variances of $X$ and $Y$,
$x_i$ and $y_i$ are the data points,
$\bar{x}$ and $\bar{y}$ are the means,
$n$ is the sample size.

Simplifying, we get:
$$r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}$$

This formula shows $r$ as a standardized measure of covariance relative to the product of standard deviations (square roots of variances).

###### Step 2: Hypothesis Testing for Correlation

To test whether the correlation is significantly different from zero, we use a t-test. The null hypothesis ($H_0$) is that the *population correlation coefficient* $\rho = 0$, and the alternative hypothesis ($H_a$) is $\rho \neq 0$ (for a two-tailed test). The t-statistic for testing the significance of $r$ is commonly given as:

$$t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}$${#eq-tstat}

This t-statistic follows a t-distribution with $n - 2$ degrees of freedom under the null hypothesis. Our goal is to derive this formula, starting from a perspective involving variances.

###### Step 3: Connecting to Variances via Linear Regression

The t-test for the correlation coefficient is closely related to the t-test for the slope of a linear regression model. Suppose we regress $Y$ on $X$:
$$Y = \beta_0 + \beta_1 X + \epsilon$$
The slope $\beta_1$ estimates the change in $Y$ per unit change in $X$. The sample slope $\beta_1$ is:
$$\beta_1 = \frac{\text{Cov}(X, Y)}{\text{Var}(X)} = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2}$$
Notice the relationship between $b_1$ and $r$. Since:
$$r = \frac{\text{Cov}(X, Y)}{\sqrt{\text{Var}(X) \text{Var}(Y)}}$$

We can express the covariance as:
$$\text{Cov}(X, Y) = r \sqrt{\text{Var}(X) \text{Var}(Y)}$$
So:
$$
\beta_1 = \frac{r \sqrt{\text{Var}(X) \text{Var}(Y)}}{\text{Var}(X)} = r \sqrt{\frac{\text{Var}(Y)}{\text{Var}(X)}} = r \frac{s_y}{s_x}
$${#eq-slope-corr}

where $s_x = \sqrt{\text{Var}(X)} = \sqrt{\sum (x_i - \bar{x})^2 / (n-1)}$ and $s_y = \sqrt{\text{Var}(Y)}$ are the sample standard deviations.

###### Step 4: t-Test for the Regression Slope

To test $H_0: \beta_1 = 0$, we use the `t-statistic` for the slope:

$$t = \frac{\beta_1}{\text{SE}(\beta_1)}$$
where $\text{SE}(\beta_1)$ is the standard error of the `slope`. The variance of the slope estimate is derived from the regression model. Assuming the errors $\epsilon$ are normally distributed with variance $\sigma^2$, the variance of $\beta_1$ is:

$$\text{Var}(\beta_1) = \frac{\sigma^2}{\sum (x_i - \bar{x})^2}$$
The standard error is:
$$\text{SE}(b_1) = \sqrt{\frac{\sigma^2}{\sum (x_i - \bar{x})^2}}$$
Since $\sigma^2$ is unknown, we estimate it with the residual variance $s^2$:
$$s^2 = \frac{\sum (y_i - \hat{y}_i)^2}{n - 2}$$
where $\hat{y}_i = \bar{y} + b_1 (x_i - \bar{x})$ are the fitted values, and $n - 2$ accounts for the degrees of freedom (two parameters estimated: $\beta_0$ and $\beta_1$).
Thus:
$$\text{SE}(b_1) = \frac{s}{\sqrt{\sum (x_i - \bar{x})^2}}$$
So the t-statistic is:

$$t = \frac{\beta_1 \sqrt{\sum (x_i - \bar{x})^2}}{s}$$

###### Step 5: Relating the Residual Variance to $r$

The residual sum of squares is:
$$\sum (y_i - \hat{y}_i)^2 = \sum (y_i - \bar{y} - b_1 (x_i - \bar{x}))^2$$

To connect this to $r$, consider the proportion of variance explained by the regression. The coefficient of determination $r^2$ (for simple linear regression, this is the square of the correlation coefficient) is:
$$r^2 = \frac{\text{SSR}}{\text{SST}} = 1 - \frac{\text{SSE}}{\text{SST}}$$
where:

- $\text{SSR} = \sum (\hat{y}_i - \bar{y})^2$ (sum of squares due to regression),
- $\text{SSE} = \sum (y_i - \hat{y}_i)^2$ (sum of squares of errors),
- $\text{SST} = \sum (y_i - \bar{y})^2$ (total sum of squares).

Since $\hat{y}_i - \bar{y} = \beta_1 (x_i - \bar{x})$, we have:

$$\text{SSR} = \sum (\beta_1 (x_i - \bar{x}))^2 = \beta_1^2 \sum (x_i - \bar{x})^2$$

Substitute $\beta_1 = r \frac{s_y}{s_x}$:

$$\beta_1^2 = r^2 \frac{s_y^2}{s_x^2} = r^2 \frac{\sum (y_i - \bar{y})^2 / (n-1)}{\sum (x_i - \bar{x})^2 / (n-1)} = r^2 \frac{\sum (y_i - \bar{y})^2}{\sum (x_i - \bar{x})^2}$$

So:
$$\text{SSR} = r^2 \frac{\sum (y_i - \bar{y})^2}{\sum (x_i - \bar{x})^2} \cdot \sum (x_i - \bar{x})^2 = r^2 \sum (y_i - \bar{y})^2$$
Thus:
$$r^2 = \frac{\text{SSR}}{\text{SST}} = \frac{r^2 \sum (y_i - \bar{y})^2}{\sum (y_i - \bar{y})^2} = r^2$$

This confirms consistency. Now, the residual sum of squares is:
$$\text{SSE} = \text{SST} (1 - r^2) = (1 - r^2) \sum (y_i - \bar{y})^2$$

The residual variance is:
$$s^2 = \frac{\text{SSE}}{n - 2} = \frac{(1 - r^2) \sum (y_i - \bar{y})^2}{n - 2}$$
So:
$$s = \sqrt{\frac{(1 - r^2) \sum (y_i - \bar{y})^2}{n - 2}}$$

###### Step 6: Substitute into the t-Statistic

Recall:
$$t = \frac{\beta_1 \sqrt{\sum (x_i - \bar{x})^2}}{s}$$

Substitute $\beta_1 = r \frac{s_y}{s_x}$, where $s_y = \sqrt{\sum (y_i - \bar{y})^2 / (n-1)}$, $s_x = \sqrt{\sum (x_i - \bar{x})^2 / (n-1)}$:

$$\beta_1 = r \frac{\sqrt{\sum (y_i - \bar{y})^2 / (n-1)}}{\sqrt{\sum (x_i - \bar{x})^2 / (n-1)}} = r \sqrt{\frac{\sum (y_i - \bar{y})^2}{\sum (x_i - \bar{x})^2}}$$

Now compute:
$$\beta_1 \sqrt{\sum (x_i - \bar{x})^2} = r \sqrt{\frac{\sum (y_i - \bar{y})^2}{\sum (x_i - \bar{x})^2}} \cdot \sqrt{\sum (x_i - \bar{x})^2} = r \sqrt{\sum (y_i - \bar{y})^2}$$

The standard error term involves:
$$s = \sqrt{\frac{(1 - r^2) \sum (y_i - \bar{y})^2}{n - 2}}$$

So:
$$t = \frac{r \sqrt{\sum (y_i - \bar{y})^2}}{\sqrt{\frac{(1 - r^2) \sum (y_i - \bar{y})^2}{n - 2}}} = \frac{r \sqrt{\sum (y_i - \bar{y})^2} \cdot \sqrt{n - 2}}{\sqrt{(1 - r^2) \sum (y_i - \bar{y})^2}}$$

The $\sqrt{\sum (y_i - \bar{y})^2}$ terms cancel out:
$$t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}$$

###### Step 7: Linking to Ratio of Variances

Our inquiry started with a “ratio of variances.” In the context of the t-test, the t-statistic can be interpreted through the lens of explained versus unexplained variance. From the regression perspective:
$$r^2 = \frac{\text{SSR}}{\text{SST}} = \frac{\text{Explained Variance}}{\text{Total Variance}}$$
The unexplained variance is:
$$1 - r^2 = \frac{\text{SSE}}{\text{SST}}$$
The t-statistic can be related to the F-statistic for regression, where:
$$F = \frac{\text{SSR}/1}{\text{SSE}/(n-2)} = \frac{r^2 / 1}{(1 - r^2)/(n-2)}$$
For simple linear regression, the t-statistic for the slope is the square root of the F-statistic:
$$t^2 = F$$
Let’s compute:
$$F = \frac{r^2 (n - 2)}{1 - r^2}$$
$$t = \sqrt{F} = \sqrt{\frac{r^2 (n - 2)}{1 - r^2}} = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}$$
This matches our derived t-statistic, confirming that the ratio of explained to unexplained variance underpins the test.

###### Summary

The t-value for testing the significance of the Pearson correlation coefficient $r$, derived from the perspective of variances in a regression framework, is:

$$t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}$$
This formula arises from the ratio of explained to unexplained variance in the regression model, where $r^2$ represents the proportion of variance in $Y$ explained by $X$, and $1 - r^2$ represents the unexplained variance, adjusted by the degrees of freedom $n - 2$.

$$\Large{\boxed{t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}}}$$

On to the computations!

#### By Hand

[< WORK IN PROGRESS >]{style="background-color: yellow;"}

:::: {layout="[15,15,15,15,15]"}

::: {#first-column}
A. Data Dimensions
```{r}
(dim(Galton_sons))
```
Sons Data: n = 465
:::
::: {#second-column}
B. Variances
```{r}
(var_son <- var(Galton_sons$son))
```
Variance of Sons' Heights
:::

::: {#third-column}
B. Variances
```{r}
(var_father <- var(Galton_sons$father))
```
Variance of Fathers' Heights
:::
::: {#fourth-column}
C. Covariance
```{r}
(cov_son_father <- cov(Galton_sons$son, Galton_sons$father))
```
Covariance of Sons' and Fathers' Heights
:::

::: {#fifth-column}
E. Estimate
```{r}
(corr_estimate <- cov_son_father / (sqrt(var_son) * sqrt(var_father)))

```
Correlation Estimate for Sons
:::
::::


F. t-statistics

We can now compute the `t-statistic` using the formula:
$$
t_{statistic} = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}
$$
We can look up the critical value of t from the `t-distribution` with $df = 463$ at a probability of error of $5\%$:

:::: {layout="[50,50]"}
::: {#first-column}

```{r}
(t_statistic<- corr_estimate/sqrt((1 - corr_estimate^2)/(465 - 2)))
```
F statistic for Sons
:::
::: {#second-column}

```{r}
(t_critical <- qt(p = 0.975, df = 463, log.p = FALSE))

```
Critical t value for Sons
:::
::::

G. p-value

Finally we can compute the `p-value` for this `t-statistic`:

```{r}
(p_value_son <- 2.0 * (1 - pt(q = t_statistic, df = 465 -2 )))

```
We see that the `p-value` is very small, and we can reject the null hypothesis of "no correlation" between `son` and `father` heights.

H. Plotting the t-distribution

```{r}
#| label: fig-t-distribution
#| layout-ncol: 2
#| fig-cap: "t-distribution with df = 463"
#| fig-subcap: 
#|   - "t-statistic and t-critical marked"
#|   - "Estimate and Confidence Intervals"
#| code-fold: true
#| out-width: "75%"
#| out-height: "75%"
#| fig-align: center

mosaic::xqt(p = c(0.025, 0.975), df = 463, 
            return = c("plot"), alpha = 0.5, 
            pattern = "rings",
            colour = "black", system = "gg") %>% 
  gf_vline(xintercept = t_statistic, color = "red") %>%
  gf_vline(xintercept = t_critical, color = "blue") %>%
  gf_annotate(geom = "label", x = t_statistic, y = 0.3, 
            label = "t-statistic", colour = "purple", size = 4) %>% 
  gf_annotate(geom = "label", x = t_critical, y = 0.3,
            label = "t-critical", colour = "purple", size = 4) %>% 
  gf_labs(title = "t-distribution with df = 463",
          subtitle = "Sons' Heights vs Fathers' Heights",
          x = "t-value", y = "Density") %>% 
  gf_refine(
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))),
    scale_x_continuous(
      breaks = c(-3, -2, -1, 0, 1, 2, 3, t_statistic, t_critical, 10),
      limits = c(-4, 10),
      labels = scales::number_format(accuracy = 0.01)))
###
df_corr <- cor_son_pearson$parameter
mean_value <- cor_son_pearson$estimate

gf_fun(dt(x = (x - mean_value) / sqrt(df_corr / (df_corr - 2)), 
          df = df_corr) * (1 / sqrt(df_corr / (df_corr - 2))) ~ x, xlim = c(-5, 5)) %>% 
  gf_labs(title = "t-Distribution with Non-Zero Mean",
          subtitle = "Sons' Heights vs Fathers' Heights",
          x = "Correlation Estimate", y = "Density") %>% 
  gf_vline(xintercept = mean_value, color = "red") %>%
  gf_annotate(geom = "label", x = mean_value, y = 0.3,
            label = "Corr Estimate", colour = "purple", size = 4) %>% 
  gf_vline(xintercept = cor_son_pearson$conf.low, color = "blue", linetype = "dashed") %>%
  gf_vline(xintercept = cor_son_pearson$conf.high, color = "blue", linetype = "dashed") %>%
  gf_annotate("segment", x = cor_son_pearson$conf.low, 
              xend = cor_son_pearson$conf.high, y = 0.25, yend = 0.25,
            arrow = arrow(ends = "both", length = unit(0.2, "inches")), color = "blue", size = 1)


```


#### Permutation Test

We can of course use a randomization based test for correlation. How would we mechanize this, what aspect would be randomize?

Correlation is calculated on a **vector-basis**: each individual observation of `variable#1` is multiplied by the **corresponding** observation of `variable#2`. Look at @eq-corr! So we might be able to randomize the *order* of this multiplication to see *how uncommon* this particular set of multiplications are. That would give us a `p-value` to decide if the **observed correlation** is close to the truth. So, onwards with our friend `mosaic`:

```{r}
obs_daughter_corr <- cor(Galton_daughters$father, Galton_daughters$daughter)
obs_daughter_corr

```

```{r}
#| out-width: "75%"
#| out-height: "75%"
#| fig-align: center
#| 
ggplot2::theme_set(new = theme_custom())

corr_daughter_null <- 
  do(4999) * mosaic::cor.test(daughter ~ shuffle(father), 
                                          data = Galton_daughters) %>% 
  broom::tidy()
corr_daughter_null

corr_daughter_null %>% 
  gf_histogram(~ estimate, bins = 50) %>% 
  gf_vline(xintercept = obs_daughter_corr, 
           color = "red",linewidth = 1) %>% 
  gf_labs(title = "Permutation Null Distribution",
          subtitle = "Daughter Heights vs Father Heights")
##
p_value_null <- 2.0*mean(corr_daughter_null$estimate >= obs_daughter_corr)
p_value_null

```

We see that will all permutations of `father`, we are never able to hit the actual `obs_daughter_corr`! Hence there is a definite correlation between `father` height and `daughter` height.

#### The Linear Model

The premise here is that **many common statistical tests are special cases of the linear model**. A **linear model** estimates the relationship between dependent variable or  
"response" variable `height` and an explanatory variable or "predictor", `father`. It is assumed that the relationship is **linear**. $\beta_0$ is the *intercept* and $\beta_1$
is the slope of the linear fit, that **predicts** the value of `height` based the value of `father`.

$$
height = \beta_0 + \beta_1 \times father
$$
The model for Pearson Correlation tests is exactly the Linear Model:

$$
\begin{aligned}
height = \beta_0 + \beta_1 \times father\\
\\
H_0: Null\ Hypothesis\ => \beta_1 = 0\\\
H_a: Alternate\ Hypothesis\ => \beta_1 \ne 0\\
\end{aligned}
$$

Using the *linear model* method we get:

```{r}
# Linear Model
lin_son <- lm(son ~ father, data = Galton_sons) %>% 
  broom::tidy() %>% 
  mutate(term = c("beta_0", "beta_1")) %>% 
  select(term, estimate, p.value)
lin_son
##
lin_daughter <- lm(daughter ~ father, data = Galton_daughters) %>% 
  broom::tidy() %>% 
  mutate(term = c("beta_0", "beta_1")) %>% 
  select(term, estimate, p.value)
lin_daughter

```

Why are the respective $r$-s and $\beta_1$-s different, though the `p-value`-s is **suspiciously the same**!? Did we miss a factor of $\frac{sd(son/daughter)}{sd(father)} = ??$ somewhere...??

Let us **scale** the variables to within `{-1, +1}` : (subtract the mean and divide by sd) and re-do the Linear Model with **scaled** versions of `height` and `father`:

```{r}
# Scaled linear model
lin_scaled_galton_daughters <- lm(scale(daughter) ~ 1 + scale(father), data = Galton_daughters) %>% 
  broom::tidy() %>% 
  mutate(term = c("beta_0", "beta_1")) 
lin_scaled_galton_daughters 

```

Now you're talking!! The `estimate` is the same in both the classical test and the linear model! So we conclude:

1.  **When both target and predictor have the same standard deviation, the slope from the linear model and the Pearson correlation are the same**.

2.  There is this relationship between the **slope in the linear model** and **Pearson correlation**:

$$
Slope\ \beta_1 = \frac{sd_y}{sd_x} * r
$$

The slope is usually much more interpretable and informative than the correlation coefficient.

2.  Hence a linear model using `scale()` for both variables will show slope = **r**.

Slope_Scaled: `r lin_scaled_galton_daughters$estimate[2]` = Correlation: `r cor_daughter_pearson$estimate`

3.  Finally, the *p-value* for Pearson Correlation and that for the *slope* in the linear model is the same ($0.04280043$). Which means we cannot reject the NULL hypothesis of "no relationship" between `daughter`-s and `father`-s heights.

Can you complete this for the sons?

:::

:::{.hidden}
## {{< iconify pajamas issue-type-test-case >}} Case Study #2: Study and Grades


In some cases the **LINE** assumptions may not hold.

Nonlinear relationships, non-normally distributed data ( with large
**outliers** ) and working with *ordinal* rather than continuous data:
these situations necessitate the use of Spearman's *ranked* correlation
scores. (**Ranked**, not **sign-ranked**.).

See the example below: We choose to look at the `gpa_study_hours`
dataset. It has two numeric columns `gpa` and `study_hours`:

```{r gpa_study_hours}
glimpse(gpa_study_hours)

```

We can plot this:

```{r Pearson_example_3}
ggplot2::theme_set(new = theme_custom())

ggplot(gpa_study_hours, aes(x = study_hours, y = gpa)) + 
  geom_point() + 
  geom_smooth() + 
  labs(title = "GPA vs Study Hours",
       subtitle = "Pearson Correlation Test")

```

Hmm...not normally distributed, and there is a sort of increasing
relationship, however is it linear? And there is some evidence of
heteroscedasticity, so the LINE assumptions are clearly in violation.
Pearson correlation would not be the best idea here.

Let us quickly try it anyway, using a Linear Model for the **scaled**
`gpa` and `study_hours` variables, from where we get:

```{r Pearson_example_1}
# Pearson Correlation as Linear Model
model_gpa <-
  lm(scale(gpa) ~ 1 + scale(study_hours), data = gpa_study_hours)
##
model_gpa %>%
  broom::tidy() %>% 
  mutate(term = c("beta_0", "beta_1")) %>% 
  cbind(confint(model_gpa) %>%                                              as_tibble()) %>%  
  select(term, estimate, p.value, `2.5 %`, `97.5 %`)

```

The correlation estimate is $0.133$; the `p-value` is $0.065$ (and the
`confidence interval` includes $0$).

Hence we fail to reject the NULL hypothesis that `study_hours` and `gpa`
have no relationship. But can this be right?

Should we use another test, that does not **need** the LINE assumptions?

#### "Signed Rank" Values

Most statistical tests use the **actual values** of the data variables.
However, in some *non-parametric* statistical tests, the data are used
in **rank-transformed** sense/order. (In some cases the **signed-rank**
of the data values is used instead of the data itself.)

**Signed Rank** is calculated as follows:

1.  Take the absolute value of each observation in a sample

2.  Place the <u>*ranks*</u> in order of (absolute magnitude). The
    smallest number has *rank = 1* and so on. This gives is **ranked
    data**.

3.  Give each of the ranks the sign of the original observation ( + or
    -). This gives us **signed** ranked data.

```{r signed_rank_function}
signed_rank <- function(x) {sign(x) * rank(abs(x))}

```

#### Plotting Original and Signed Rank Data

Let us see how this might work by comparing data and its signed-rank
version...A quick set of plots:

```{r data_plots}
#| include: false
#| eval: false
ggplot2::theme_set(new = theme_custom())

p1 <- ggplot(mydata_long, aes(x = group, y = value)) +
  geom_jitter(width = 0.02,
              height = 0,
              aes(colour = group),
              size = 4) +
  geom_segment(data = mydata_wide,
               aes(
                 y = 0,
                 yend = 0,
                 x = .75,
                 xend = 1.25
               )) +
  geom_text(aes(x = 1, y = 0.5, label = "0")) +
  geom_segment(data = mydata_wide,
               aes(
                 y = 0.3,
                 yend = 0.3,
                 x = 1.75 ,
                 xend = 2.25
               )) +
  geom_text(aes(x = 2, y = 0.6, label = "0.3")) +
  geom_segment(data = mydata_wide,
               aes(
                 y = 0.5,
                 yend = 0.5,
                 x = 2.75,
                 xend = 3.25
               )) +
  geom_text(aes(x = 3, y = 0.8, label = "0.5")) +
  labs(title = "Original Data",
       subtitle = "Black Lines show Means") +
  ylab("Response Variable")

p2 <- mydata_long %>%
  group_by(group) %>%
  mutate(s_value = signed_rank(value)) %>%
  ggplot(., aes(x = group, y = s_value)) +
  geom_jitter(width = 0.02,
              height = 0,
              aes(colour = group),
              size = 4) +
  stat_summary(
    fun = "mean",
    geom = "point",
    colour = "red",
    size = 8
  ) +
  labs(title = "Signed Rank of Data",
       subtitle = "Red Points are means of Ranks!") +
  ylab("Signed Rank of Response Variable")

patchwork::wrap_plots(p1, p2, nrow = 1, guides = "collect") 

```

So the means of the **ranks** three separate variables seem to be in the
same order as the means of the data variables themselves.

How about associations between data? Do ranks reflect well what the data
might?

```{r Spearman_Plot}
#| eval: false
#| include: false
# Plot the data
ggplot2::theme_set(new = theme_custom())

p1 <- ggplot(mydata_wide, aes(x, y1)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle(" Pearson Correlation\n and Linear Models")

# Plot ranked data
p2 <- mydata_wide %>% 
  mutate(x_rank = rank(x),
         y_rank = rank(y1)) %>%
  ggplot(.,aes(x_rank, y_rank)) + 
  geom_point(shape = 15, size = 2) +
  geom_smooth(method = "lm") + 
  ggtitle(" Spearman Ranked Correlation\n and Linear Models")

patchwork::wrap_plots(p1,p2, nrow = 1, guides = "collect")

```

The slopes are almost identical, $0.25$ for both original data and
ranked data for $y1\sim x$. So maybe *ranked* and even *sign_ranked*
data could work, and if it can work despite LINE assumptions not being
satisfied, that would be nice!

#### How does Sign-Rank data work?

TBD: need to add some explanation here.

Spearman correlation = Pearson correlation using the rank of the data
observations. Let's check how this holds for a our x and y1 data:

So the Linear Model for the Ranked Data would be:

$$
\begin{aligned}
y = \beta_0 + \beta_1 \times rank(x)\\
\\
H_0: Null\ Hypothesis\ => \beta_1 = 0\\\
H_a: Alternate\ Hypothesis\ => \beta_1 \ne 0\\
\end{aligned}
$$

#### Code

```{r}
#| include: false
#| eval: false
# Spearman
cor1 <- cor.test(x, y1, method = "spearman") %>%
  broom::tidy() %>% mutate(term = "Spearman Correlation ") %>% select(term, estimate, p.value)
cor1

```

```{r}
#| include: false
#| eval: false
# Pearson using ranks
cor2 <- cor.test(rank(y1), rank(x), method = "pearson") %>%
  broom::tidy() %>% select(estimate, p.value)
cor2
```

```{r}
#| include: false
#| eval: false
# Linear Models using rank
cor3 <- lm(rank(y1) ~ 1 + rank(x),data = mydata_wide) %>% 
  broom::tidy() %>% select(estimate, p.value)
cor3
```

Notes:

1.  When ranks are used, the slope of the linear model ($\beta_1$) has
    the same value as the Spearman correlation coefficient ( $\rho$ ).

2.  Note that the slope from the linear model now has an intuitive
    interpretation: **the number of ranks y changes for each change in
    rank of x**. ( Ranks are "independent" of `sd` )

#### Example

We examine the `cars93` data, where the numeric variables of interest
are `weight` and `price`.

```{r Spearman_example_1}
ggplot2::theme_set(new = theme_custom())

cars93 %>% 
  ggplot(aes(weight, price)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, lty = 2) + 
  labs(title = "Car Weight and Car Price have a nonlinear relationship") + theme_classic()
```

Let us try a Spearman Correlation score for these variables, since the
data are not linearly related and the variance of `price` also is not
constant over `weight`

```{r Spearman_example_2}
#| warning: false
#| message: false
ggplot2::theme_set(new = theme_custom())

cor.test(cars93$price, cars93$weight, method = "spearman") %>% broom::tidy()

# Using linear Model
lm(rank(price) ~ rank(weight), data = cars93) %>% summary()

# Stats Plot
ggstatsplot::ggscatterstats(data = cars93, x = weight, 
                            y = price,
                            type = "nonparametric",
                            title = "Cars93: Weight vs Price",
                            subtitle = "Spearman Correlation")
```

We see that using ranks of the `price` variable, we obtain a Spearman's
$\rho = 0.882$ with a `p-value` that is very small. Hence we are able to
reject the NULL hypothesis and state that there is a relationship
between these two variables. The **linear** relationship is evaluated as
a correlation of `0.882`.

```{r}
# Other ways using other packages
mosaic::cor_test(gpa ~ study_hours, data = gpa_study_hours) %>%
  broom:: tidy() %>% 
  select(estimate, p.value, conf.low, conf.high)

```

```{r}

statsExpressions::corr_test(data = gpa_study_hours, 
                            x = study_hours, 
                            y = gpa)
```
:::

## {{< iconify mingcute thought-line >}} Wait, But Why?

Correlation tests are useful to understand the relationship between two variables, but they do not imply causation. A high correlation does not mean that one variable causes the other to change. It is essential to consider the context and other factors that may influence the relationship.

Correlations also become an important thing to evaluate in Linear Regression. 

## {{< iconify fluent-mdl2 decision-solid >}} Conclusion

Correlation tests are a powerful way to understand the relationship between two variables. They can be performed using classical methods like Pearson and Spearman correlation, or using more robust methods like permutation tests. The linear model approach provides a deeper understanding of the relationship, especially when the assumptions of normality and homoscedasticity are met.

## {{< iconify openmoji person >}} Your Turn

1. Try the datasets in the `infer` package. Use `data(package = "infer")` in your Console to list out the data packages. Then simply type the name of the dataset in a Quarto chunk ( e.g. `babynames`) to read it. 

2. Same with the `resampledata` and `resampledata3` packages. 


## {{< iconify ooui references-rtl >}} References

1. *Common statistical tests are linear models (or: how to teach
    stats)* by [Jonas Kristoffer Lindeløv](https://lindeloev.github.io/tests-as-linear/)\
    [CheatSheet](https://lindeloev.github.io/tests-as-linear/linear_tests_cheat_sheet.pdf)\
1.    *Common statistical tests are linear models: a work through* by [Steve Doogue](https://steverxd.github.io/Stat_tests/)\
1.    [Jeffrey Walker "Elements of Statistical Modeling for Experimental Biology"](https://www.middleprofessor.com/files/applied-biostatistics_bookdown/_book/)\
1.   Diez, David M & Barr, Christopher D & Çetinkaya-Rundel, Mine: [OpenIntro Statistics](https://www.openintro.org/book/os/)\
1.   Modern Statistics with R: From wrangling and exploring data to inference and predictive modelling by [Måns
    Thulin](http://www.modernstatisticswithr.com/)\
1.    [Jeffrey Walker "A
    linear-model-can-be-fit-to-data-with-continuous-discrete-or-categorical-x-variables"](https://www.middleprofessor.com/files/applied-biostatistics_bookdown/_book/intro-linear-models.html#a-linear-model-can-be-fit-to-data-with-continuous-discrete-or-categorical-x-variables)
1. <https://grok.com/share/c2hhcmQtNA%3D%3D_5a4873eb-9c38-4ee2-a6dc-de1505c415d5>
    
## Examples

1. <https://stats.libretexts.org/Courses/Citrus_College/Statistics_C1000%3A_Introduction_to_Statistics/10%3A_Correlation_and_Linear_Regression/10.04%3A_Hypothesis_Test_for_a_Correlation_Using_t-Test>

::: {#refs style="font-size: 60%;"}

```{r}
#| echo: false
# scan_packages()
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("easystats", "ggExtra", "ggstatsplot",
           "openintro", "resampledata3", "statsExpressions")
) %>%
  knitr::kable(format = "simple")

```

:::


