---
title: "Permutation Tests"
author: "Arvind V."
date: 21/Nov/2022
date-modified: "`r Sys.Date()`"
abstract: A bunch of case studies with Permutation Tests
code-summary: "Show the Code"
df-print: paged
execute: 
  freeze: auto
---

```{r setup, include=FALSE}

#knitr::opts_chunk$set(echo = TRUE, out.width="70%", out.height = "70%", fig.align = "center")
library(ggformula)
library(tidyverse)
library(skimr)
library(mosaic)
library(resampledata3)
library(openintro)

```

#### Plot Fonts and Theme

```{r}
#| label: plot-theme-ggplot4
#| cache: false
#| code-fold: true
#| messages: false
#| warning: false
#| 
library(systemfonts)
library(showtext)
library(ggrepel)
library(marquee)
## Clean the slate
systemfonts::clear_local_fonts()
systemfonts::clear_registry()
##
showtext_opts(dpi = 96) #set DPI for showtext
sysfonts::font_add(family = "Alegreya",
  regular = "../../../../../../fonts/Alegreya-Regular.ttf",
  bold = "../../../../../../fonts/Alegreya-Bold.ttf",
  italic = "../../../../../../fonts/Alegreya-Italic.ttf",
  bolditalic = "../../../../../../fonts/Alegreya-BoldItalic.ttf")

sysfonts::font_add(family = "Roboto Condensed", 
  regular = "../../../../../../fonts/RobotoCondensed-Regular.ttf",
  bold = "../../../../../../fonts/RobotoCondensed-Bold.ttf",
  italic = "../../../../../../fonts/RobotoCondensed-Italic.ttf",
  bolditalic = "../../../../../../fonts/RobotoCondensed-BoldItalic.ttf")
showtext_auto(enable = TRUE) #enable showtext
##
theme_custom <- function(){ 

    theme_bw(base_size = 10) + 
    
    theme_sub_axis(title = element_text(family = "Roboto Condensed", 
                                       size = 8),
                   text = element_text(family = "Roboto Condensed", 
                                       size = 6)) + 
    
    theme_sub_legend(text = element_text(family = "Roboto Condensed", 
                                         size = 6),
                     title = element_text(family = "Alegreya", 
                                          size = 8)) + 
    
    theme_sub_plot(title = element_text(family = "Alegreya", 
                                        size = 14, face = "bold"),
                   title.position = "plot",
                   subtitle = element_text(family = "Alegreya", 
                                           size = 10),
                   caption = element_text(family = "Alegreya", 
                                          size = 6),
                   caption.position = "plot")
    
}

## Use available fonts in ggplot text geoms too!
ggplot2::update_geom_defaults(geom = "text", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

ggplot2::update_geom_defaults(geom = "marquee", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "text_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)
ggplot2::update_geom_defaults(geom = "label_repel", new = list(
  family = "Roboto Condensed",
  face = "plain",
  size = 3.5,
  color = "#2b2b2b"
)
)

## Set the theme
ggplot2::theme_set(new = theme_custom())

## tinytable options
options("tinytable_tt_digits" = 2)
options("tinytable_format_num_fmt" = "significant_cell")
options(tinytable_html_mathjax = TRUE)


## Set defaults for flextable
flextable::set_flextable_defaults(font.family = "Roboto Condensed")

```



## Introduction

We saw from the diagram created by Allen Downey that *there is only one
test*! We will now use this philosophy to develop a technique that
allows us to mechanize several *Statistical Models* in that way, with
nearly identical code.

We will use two packages in R, `mosaic` and the relatively new `infer`
package, to develop our intuition for what are called **permutation**
based statistical tests.

## Hypothesis Testing using Permutation

From Reference #1:

> Hypothesis testing can be thought of as a 4-step process:
>
> 1.  State the null and alternative hypotheses.
>
> 2.  Compute a test statistic.
>
> 3.  Determine the p-value.
>
> 4.  Draw a conclusion.
>
>     In a traditional introductory statistics course, once this general
>     framework has been mastered, the main work is in **applying the
>     correct formula** to compute the standard test statistics in step
>     2 and using a table or computer to **determine the p-value** based
>     on the known (usually approximate) **theoretical distribution of
>     the test statistic** under the null hypothesis.
>
>     In a **simulation-based approach**, steps 2 and 3 change. In Step
>     2, it is no longer required that the test statistic be normalized
>     to conform with a known, named distribution. Instead, natural test
>     statistics, like the difference between two sample means $y1 − y2$
>     can be used.
>
>     In Step 3, we use **randomization to approximate the sampling
>     distribution of the test statistic**. Our lady tasting tea example
>     demonstrates how this can be done from first principles. More
>     typically, we will use randomization **to create new simulated
>     data sets** ( "*Parallel Worlds*") that are like our original data
>     in some ways, but make the null hypothesis true. For each
>     simulated data set, we calculate our test statistic, just as we
>     did for the original sample. Together, this collection of test
>     statistics computed from the simulated samples constitute our
>     randomization distribution.
>
>     When creating a randomization distribution, we will attempt to
>     satisfy 3 guiding principles.
>
> 5.  Be consistent with the null hypothesis. We need to **simulate a
>     world** in which the null hypothesis is true. If we don't do this,
>     we won't be testing our null hypothesis.
>
> 6.  Use the data in the **original sample**. The original data should
>     shed light on some aspects of the distribution that are not
>     determined by null hypothesis. For example, a null hypothesis
>     about a mean doesn't tell us about the shape of the population
>     distribution, but the data give us some indication.
>
> 7.  Reflect the way the original data were collected.

From Chihara and Hesterberg:

> This is the core idea of statistical significance or classical
> hypothesis testing -- to calculate how often pure random chance would
> give an effect as large as that observed in the data, in the absence
> of any real effect. If that probability is small enough, we conclude
> that the data provide convincing evidence of a real effect.


## Permutations tests using mosaic::`shuffle()`

The `mosaic` package provides the `shuffle()` function as a synonym for
`sample()`. When used without additional arguments, this will permute
its first argument.

```{r, echo=TRUE}
# library(mosaic)
shuffle(1:10)

```

Applying `shuffle()` to an *explanatory variable* in a model allows us to
test the null hypothesis that the explanatory variable has, in fact, no
explanatory power. This idea can be used to test

-   the equivalence of two or more means,
-   the equivalence of two or more proportions,
-   whether a regression parameter is 0. (Correlations between two
    variables)

Coupled with `mosaic::do()` we can repeat a `shuffle` many times,
computing a desired statistic each time we shuffle. The distribution of
this computed statistic is a NULL distribution, which can then be
compared with the observed statistic to decide upon the Hypothesis Test
and p-value.


### Case Study-1: Hot Wings Orders vs Gender

A student conducted a study of hot wings and beer consumption at a Bar.
She asked patrons at the bar to record their consumption of hot wings
and beer over the course of several hours. She wanted to know if people
who ate more hot wings would then drink more beer. In addition, she
investigated whether or not gender had an impact on hot wings or beer
consumption.

```{r}

Beerwings <- resampledata3::Beerwings # From resampledata3 package
skimr::skim(Beerwings)

```

Let us calculate the observed difference in `Hotwings` consumption
between Males and Females ( `Gender`)

```{r}

mean(Hotwings ~ Gender, data = Beerwings)
obs_diff_wings <- mosaic::diffmean(data = Beerwings, Hotwings ~ Gender)
obs_diff_wings 

```

```{r}

gf_boxplot(data = Beerwings, Hotwings ~ Gender, 
           orientation = "x", fill = ~ Gender,
           title = "Hotwings Consumption by Gender")

```

The observed difference in mean consumption of Hotwings between Males
and Females is $5.2$. Could this have occurred by chance? Here is our
formulation of the Hypotheses:

$$
NULL\ Hypothesis\ H_0:  \mu_F =  \mu_M\\
\\
\\
Alternative\ Hypothesis\
H_a: \mu_F \ne \mu_M\\
$$

So we perform a Permutation Test to check:

```{r}

null_dist_wings <- do(1000) * diffmean(Hotwings ~ shuffle(Gender), 
                                       data = Beerwings)
null_dist_wings %>% head()

gf_histogram(data = null_dist_wings, ~ diffmean,
             fill = ~ (diffmean >= obs_diff_wings)) %>% 
  gf_vline(xintercept = obs_diff_wings, colour = "red") %>% 
  gf_refine(scale_fill_manual(values = c("lightseagreen", "tomato")))

prop1(~ diffmean >= obs_diff_wings, data = null_dist_wings)

```

The $\color{red}{red\ line}$ shows the actual measured mean difference
in Hot Wings consumption. The probability that our Permutation
distribution is able to equal or exceed that number is $0.001998002$ and
we have to reject the Null Hypothesis that the means are identical.

::: {.content-hidden}
To test whether eating more hotwings would lead to increased beer
consumption, we need a regression model, which we can again test with a
permutation test.

```{r}
lm(Beer ~ Hotwings, data = Beerwings)
```
:::


### Case Study-2: Verizon

Verizon was an Incumbent Local Exchange Carrier (ILEC), responsible for maintaining land-line phone service in certain areas. Verizon also sold
long-distance service, as did a number of competitors, termed
Competitive Local Exchange Carriers (CLEC). When something went wrong,
Verizon was responsible for repairs, and was supposed to make repairs as
quickly for CLEC long-distance customers as for their own. 

The New York Public Utilities Commission (PUC) monitored fairness by comparing repair times for Verizon and different CLECs, for different classes of repairs
and time periods. In each case a hypothesis test was performed at the 1%
significance level, to determine whether repairs for CLEC's customers
were significantly slower than for Verizon's customers. There were
hundreds of such tests. If substantially more than 1% of the tests were
significant, then Verizon would pay large penalties. 

These tests were performed using t tests; Verizon proposed using permutation tests
instead.

```{r}

verizon <- resampledata3::Verizon # From resampledata3 package
verizon
skim(verizon)

```

Let us make a boxplot of the repair times for Verizon and CLEC:

```{r}

gf_boxplot(data = verizon, Group ~ Time, 
           orientation = "y", fill = ~ Group,
           title = "Repair Times by Operator")

```

```{r}

mean(Time ~ Group, data = verizon)
obs_diff_verizon <- diffmean(Time ~ Group, data = verizon)
obs_diff_verizon

```

```{r}

null_dist_verizon <- do(1000) * diffmean(Time ~ shuffle(Group), 
                                         data = verizon)
gf_histogram(data = null_dist_verizon, ~ diffmean,
             fill = ~ (diffmean >= obs_diff_verizon)) %>% 
  gf_vline(xintercept = obs_diff_verizon, colour = "red") %>% 
  gf_refine(scale_fill_manual(values = c("lightseagreen", "tomato")))


prop1(~ diffmean >= obs_diff_wings, data = null_dist_verizon)

```

It seems that the repair `Time` for CLEC is significantly higher than for the ILEC and there only about a $1.2\%$ chance that this could have occurred by chance. So we reject the Null Hypothesis that the mean repair `Time`-s are identical.


### Case Story-3: Recidivism

Do criminals released after a jail term commit crimes again?

```{r}

recidivism <- resampledata3::Recidivism # From resampledata3 package
recidivism
skim(recidivism)

```

There are some missing values in the variable <tt> `Age25`</tt>. The
<tt> `complete.cases`</tt> command gives the row numbers where values
are not missing. We create a new data frame omitting the rows where
there is a missing value in the <tt> 'Age25' </tt> variable.

```{r}

recidivism_na <- recidivism %>% tidyr::drop_na(Age25)

```

Also, the variable `Recid`  is a factor variable coded "Yes" or
"No". We convert it to a numeric variable of 1's and 0's.

```{r}

recidivism_na <- recidivism_na %>% 
  mutate(Recid2 = ifelse(Recid=="Yes", 1, 0))

obs_diff_recid <- diffmean( Recid2 ~ Age25, data = recidivism_na)
obs_diff_recid

null_dist_recid <- do(1000) * diffmean( Recid2 ~ shuffle(Age25), data = recidivism_na)

gf_histogram( ~ diffmean, data = null_dist_recid) %>% 
  gf_vline(xintercept = obs_diff_recid, colour = "red")

```

The probability that the difference in means of `Recid2` between the two age groups is as large as the observed value is about $0.15$, and hence we cannot reject the Null Hypothesis that the means are identical.


### Case Study-4: Matched Pairs: Results from a diving championship.

```{r}

Diving2017 <- resampledata3::Diving2017 # From resampledata3 package
head(Diving2017)
skim(Diving2017)

```

The data is made up of **paired** observations per swimmer. So we need
to take the difference between the two swim records for *each* swimmer
and then shuffle the differences to either polarity. Another way to look
at this is to shuffle the records between `Semifinal` and `Final` on a
per Swimmer basis.

```{r}

obs_diff_swim <- mean(~ Final - Semifinal, data = Diving2017)
obs_diff_swim

```

```{r}

polarity <- c(rep(1, 6), rep(-1,6))
polarity
null_dist_swim <- do(100000) * mean(data = Diving2017, 
                                    ~(Final - Semifinal) * resample(polarity,
                                                    replace = TRUE))
null_dist_swim %>% head()
gf_histogram(data = null_dist_swim, ~mean) %>% 
  gf_vline(xintercept = obs_diff_swim, colour = "red")

```

It appears that there no reason to accept the Alternate Hypothesis that the Diving scores are different across semi-final and final events. 




### Case Study #5: Flight Delays

LaGuardia Airport (LGA) is one of three major airports that serves the
New York City metropolitan area. In 2008, over 23 million passengers and
over 375 000 planes flew in or out of LGA. United Airlines and America
Airlines are two major airlines that schedule services at LGA. The data
set `FlightDelays` contains information on all 4029 departures of these
two airlines from LGA during May and June 2009.

```{r}

flightDelays <- resampledata3::FlightDelays # From resampledata3 package
skim(flightDelays)

```

The variables in the `flightDelays` dataset are:

| Variable     | Description                                                 |
|--------------|-------------------------------------------------------------|
| Carrier      | UA=United Airlines, AA=American Airlines                    |
| FlightNo     | Flight number                                               |
| Destination  | Airport code                                                |
| DepartTime   | Scheduled departure time in 4 h intervals                   |
| Day          | Day of the Week                                             |
| Month        | May or June                                                 |
| Delay        | Minutes flight delayed (negative indicates early departure) |
| Delayed30    | Departure delayed more than 30 min? Yes or No               |
| FlightLength | Length of time of flight (minutes)                          |

: flightDelay dataset variables

a)  Let us compute the proportion of times that each carrier's flights
    was delayed more than 20 min. We will conduct a two-sided test to
    see if the difference in these proportions is statistically
    significant.

```{r flightDelays-a}

prop(data = flightDelays, (Delay >= 20) ~ Carrier)
obs_diff_delay <- diffprop(data = flightDelays, Delay >= 20 ~ Carrier)
obs_diff_delay

```

We see carrier AA has a 17.13% chance of delays\>= 20, while UA has
22.26% chance. The difference is 5.12%. Is this statistically
significant? We take the Delays for both Carriers and perform a
permutation test by `shuffle` on the `carrier` variable:

```{r}

null_dist_delay <- do(4999) * diffprop(data = flightDelays, Delay >= 20 ~ shuffle(Carrier))
null_dist_delay %>% head()
gf_histogram(data = null_dist_delay, ~ diffprop) %>% gf_vline(xintercept = obs_diff_delay, color = "red")

```

It appears that the difference in delay times is significant. We can
compute the `p-value` based on this test:

```{r}

2 * mean(null_dist_delay >= obs_diff_delay)

```

which is very small. Hence we reject the null Hypothesis that there is
no difference between `carrier`s on `delay times`.

b)  Compute the variance in the flight delay lengths for each carrier.
    Conduct a test to see if the variance for United Airlines differs
    from that of American Airlines.

```{r}

var(data = flightDelays, Delay ~ Carrier)

# There is no readymade function in mosaic called `diffvar`...so...we construct one
obs_diff_var <- diff(var(data = flightDelays, Delay ~ Carrier))
obs_diff_var

```

The difference in variances in `Delay` between the two `carrier`s is
$-431.0677$. In our Permutation Test, we `shuffle` the `Carrier`
variable:

```{r}

obs_diff_var <- diff(var(data = flightDelays, Delay ~ Carrier))
null_dist_var <-
  do(4999) * diff(var(data = flightDelays, Delay ~ shuffle(Carrier)))
null_dist_var %>% head()

# The null distribution variable is called `UA`
gf_histogram(data = null_dist_var, ~ UA) %>% 
  gf_vline(xintercept = obs_diff_delay, color = "red")
2 * mean(null_dist_var >= obs_diff_var)

```

Clearly there is no case for a significant difference in variances!


### Case Study #6: Walmart vs Target

Is there a difference in the price of groceries sold by the two
retailers Target and Walmart? The data set `Groceries` contains a sample
of grocery items and their prices advertised on their respective web
sites on one specific day.

a)  Inspect the data set, then explain why this is an example of matched
    pairs data.
b)  Compute summary statistics of the prices for each store.
c)  Conduct a permutation test to determine whether or not there is a
    difference in the mean prices.
d)  Create a ~~histogram~~ bar-chart of the difference in prices. What
    is unusual about Quaker Oats Life cereal?
e)  Redo the hypothesis test without this observation. Do you reach the
    same conclusion?

```{r}
groceries <- resampledata3::Groceries %>% # From resampledata3 package
  mutate(Product = stringr::str_squish(Product))
groceries
skim(groceries)

```

We see that the comparison is to be made between two prices for the
*same* product, and hence this is one more example of `paired data`, as
in Case Study #4. Let us plot the prices for the products:

```{r}
#| code-fold: true
#| label: fig-walmart-vs-target
#| fig-cap: "Figure: Walmart vs Target Prices"
#| 
gf_col(data = groceries,
       Target ~ Product,
       fill = "#0073C299",
       width = 0.5 ) %>% 
  gf_col(data = groceries,
         -Walmart ~ Product,
         fill = "#EFC00099",
         ylab = "Prices",
         width = 0.5
       ) %>% 
  gf_col(data = groceries %>% filter(Product == "Quaker Oats Life Cereal Original"), 
         -Walmart ~ Product,
         fill = "red", 
         width = 0.5) %>% 
  gf_theme(theme_classic()) %>%
  gf_theme(ggplot2::theme(axis.text.x = element_text(
    size = 8,
    face = "bold",
    vjust = 0,
    hjust = 1
  ))) %>% gf_theme(ggplot2::coord_flip())

```

We see that the price difference between Walmart and Target prices is
highest for the `Product` named `Quaker Oats Life Cereal Original`. Let
us check the mean difference in prices:

```{r}

diffmean(data = groceries, Walmart ~ Target, only.2 = FALSE)
obs_diff_price = mean( ~ Walmart - Target, data = groceries)
obs_diff_price

```

Let us perform the pair-wise permutation test on prices, by shuffling
the two store names:

```{r}

polarity <- c(rep(1, 15), rep(-1,15))
polarity
null_dist_price <- do(100000) * mean(data = groceries, 
                                    ~(Walmart-Target) * resample(polarity,
                                                    replace = TRUE))
null_dist_price %>% head()
gf_histogram(data = null_dist_price, ~mean) %>% 
  gf_vline(xintercept = obs_diff_price, colour = "red")

2*(sum(null_dist_price >= obs_diff_price + 1)/(100000+1)) #P-value

```

Does not seem to be any significant difference in prices...

Suppose we knock off the Quaker Cereal data item...

```{r}

which(groceries$Product == "Quaker Oats Life Cereal Original")
groceries_less <- groceries[-2,]
groceries_less


obs_diff_price_less = mean( ~ Walmart - Target, data = groceries_less)
obs_diff_price_less
polarity_less <- c(rep(1, 15), rep(-1,14)) # Due to resampling this small bias makes no difference
null_dist_price_less <- do(100000) * mean(data = groceries_less, 
                                    ~(Walmart-Target) * resample(polarity_less,
                                                    replace = TRUE))
null_dist_price_less %>% head()
gf_histogram(data = null_dist_price_less, ~mean) %>% 
  gf_vline(xintercept = obs_diff_price_less, colour = "red")

1- mean(null_dist_price_less >= obs_diff_price_less) #P-value

```

### Case Study 7: Proportions between Categorical Variables

Let us try a dataset with Qualitative / Categorical data. This is a
General Social Survey dataset, and we have people with different levels
of `Education` stating their opinion on the `Death Penalty`. We want to
know if these two Categorical variables have a correlation, i.e. can the
opinions in favour of the Death Penalty be explained by the Education
level?

Since data is Categorical, we need to take `counts` in a table, and then
implement a `chi-square test`. In the test, we will permute the
`Education` variable to see if we can see how significant its *effect
size* is.

```{r}

GSS2002 <- resampledata::GSS2002 # From resampledata package

skim(GSS2002)

```

Note how *all* variables are Categorical !! `Education` has five
`levels`:

```{r}

GSS2002 %>% count(Education)
GSS2002 %>% count(DeathPenalty)

```

Let us drop NA entries in Education and Death Penalty. And set up a
table for the chi-square test.

```{r,warning=FALSE,message=FALSE}

gss2002 <- GSS2002 %>% 
  dplyr::select(Education, DeathPenalty) %>% 
  tidyr::drop_na(., c(Education, DeathPenalty))
dim(gss2002)

gss_summary <- gss2002 %>%
  mutate(
    Education = factor(
      Education,
      levels = c("Bachelors", "Graduate", "Jr Col", "HS", "Left HS"),
      labels = c("Bachelors", "Graduate", "Jr Col", "HS", "Left HS")
    ),
    DeathPenalty = as.factor(DeathPenalty)
  ) %>%
  group_by(Education, DeathPenalty) %>%
  summarise(count = n()) %>% # This is good for a chisq test
  
  # Add two more columns to faciltate mosaic/Marrimekko Plot
  # 
  mutate(edu_count = sum(count), 
         edu_prop = count / sum(count)) %>%
  ungroup() 

gss_summary

# We can plot a heatmap-like `mosaic chart` for this table, using `ggplot`:
# https://stackoverflow.com/questions/19233365/how-to-create-a-marimekko-mosaic-plot-in-ggplot2

ggplot(data = gss_summary, aes( x = Education, y = edu_prop)) +
  geom_bar(aes(width = edu_count, fill = DeathPenalty), stat = "identity", position = "fill", colour = "black") +
  geom_text(aes(label = scales::percent(edu_prop)), position = position_stack(vjust = 0.5)) +


# if labels are desired
 facet_grid(~ Education, scales = "free_x", space = "free_x") + 
  theme(scale_fill_brewer(palette = "RdYlGn")) + 
  # theme(panel.spacing.x = unit(0, "npc")) + # if no spacing preferred between bars
  theme_void() 

```

Let us now perform the base `chisq test`: We need a `table` and then the
`chisq` test:

```{r}

gss_table <- tally(DeathPenalty ~ Education, data = gss2002)
gss_table

# Get the observed chi-square statistic
observedChi2 <- mosaic::chisq(tally(DeathPenalty ~ Education, data = gss2002))
observedChi2

# Actual chi-square test
stats::chisq.test(tally(DeathPenalty ~ Education, data = gss2002))

```

We should now repeat the test with permutations on `Education`:

```{r}

null_chisq <- do(4999) * chisq.test(tally(DeathPenalty ~ shuffle(Education), data = gss2002))

head(null_chisq)

gf_histogram( ~ X.squared, data = null_chisq) %>% 
  gf_vline(xintercept = observedChi2, color = "red")

```

So we would conclude that `Education` has a significant effect on
`DeathPenalty` opinion!


## References

1.  [Chapter 11, Hypothesis Testing with
    Randomization](https://openintro-ims.netlify.app/foundations-randomization.html#chp11-review)
    in [Introduction to Modern Statistics (1st
    Ed)](https://openintro-ims.netlify.app/index.html) by Mine
    Çetinkaya-Rundel and Johanna Hardin.
1. Permutation Analysis of Arnold Schwarzenegger's Veto Letter. <https://www.stat.berkeley.edu/~stark/Preprints/acrosticVeto09.htm>

::: {#refs style="font-size: 60%;"}

{{< iconify lucide package-check >}} R Package Citations

```{r}
#| echo: false
library(grateful)
cite_packages(
  output = "table",
  out.dir = ".",
  out.format = "html",
  pkgs = c("ggformula", "skimr", "mosaic", "resampledata3")
) %>%
  knitr::kable(format = "simple")

```
